---
title: "W1. Introduction & Roadmap"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Say what last semester's bulk RNA-seq answered, and what it structurally could not
-   ⚠️ Distinguish **execution depth** from **judgment depth**, and say which one this course is for
-   Trace one clinical question through all eight omics layers of the semester
-   ⚠️ Name the point in a study after which design mistakes become unrecoverable
-   Recognise batch–group confounding from a sample layout alone
-   Say what a fresh, frozen, or FFPE specimen can and cannot become
-   Set up an R project that still runs in three years
:::

> 🤔 **Motivating question**
>
> A colleague asks you to co-author. They have RNA-seq on 6 patients and 6 controls, and a list of 847 differentially expressed genes. What is the first question you ask — about the genes, or about something else?

------------------------------------------------------------------------

# 1. Where You Are Now 📍 {#w1-s1}

## 1) What last semester actually gave you {#w1-s1-1}

You can take a tissue, get counts, and produce a ranked gene list. That is a real skill, and most clinicians do not have it.

It is also a **single measurement of a mixture**, and it answers exactly one question: *which genes differ on average between these two groups of tissue?*

![Every remaining week of this course attacks that same gene list from a different side.](/assets/img/w1/where_you_are.svg){fig-align="center"}

💡 Notice that none of the six questions on the right is answered by better statistics on the left. They need **different data**.

## 2) The five stages have not changed {#w1-s1-2}

Whatever the technology — genotypes, chromatin, single cells, peptides — the pipeline shape is the same one you already know:

```         
① Quality control     → is this data usable at all?
② Alignment /          → where did each measurement come from,
   quantification         and how much of it is there?
③ Normalization       → make samples comparable to each other
④ Statistical testing → which differences exceed noise?
⑤ Interpretation      → what does that mean biologically?
```

::: {.callout-important collapse="true"}
## The stages are not equally dangerous

-   Stages ① ② fail **loudly** — the pipeline crashes, or the numbers are absurd
-   Stages ③ ④ fail **quietly** — you get a tidy table of 847 genes, sorted by adjusted p, and nothing about its appearance tells you it is wrong

⚠️ The failure mode unique to computation: **the analysis always returns an answer.** A failed PCR gives you an empty gel. A failed analysis gives you a figure.
:::

## 3) ⚠️ Two kinds of depth {#w1-s1-3}

"Advanced bioinformatics" could mean two very different things. This course commits to one of them.

![The left column is a full-time job. The right column is the one you are uniquely qualified for.](/assets/img/w1/two_depths.svg){fig-align="center"}

::: {.callout-tip collapse="true"}
## Why the right column is yours and not the analyst's

Your analyst can tell you the pseudobulk model converged. They cannot tell you that the "depleted" cell type is the one that never survives dissociation from inflamed tissue — because they have never held the tissue.

-   ✅ **You** know what the specimen went through, what the patients were on, and what the diagnosis actually means
-   ✅ **You** know whether a 2% absolute risk difference changes management
-   ❌ Nobody can supply that for you, and no method compensates for its absence

→ The gap this course closes is not a coding gap. It is a **vocabulary gap** — enough to ask the right question and recognise a wrong answer.
:::

------------------------------------------------------------------------

# 2. The Course as One Question 🧭 {#w1-s2}

## 1) One clinical question, eight layers {#w1-s2-1}

The semester is not a tour of techniques. It is one question, pursued downward until it becomes something you could actually test.

![Each layer answers something the layer above it could not.](/assets/img/w1/roadmap.svg){fig-align="center"}

💡 Read the right-hand column downward. Notice that **the answer gets narrower and more actionable** at every step — from 40 loci to one gene in one cell type at one tissue location with one candidate drug.

⚠️ Notice also the W9–W10 row: *protein up, mRNA flat*. Layers disagree more often than papers admit, and disagreement is usually the informative part.

## 2) The syllabus {#w1-s2-2}

::: {style="font-size: 0.92em"}
| Module | Week | Contents | Key tools |
|---------------|---------------|-------------------------|-----------------|
| **Intro** | W1 | Review, roadmap, R environment | R, Bioconductor |
|  | W2 | R scripts, AI-assisted coding, reproducibility | Quarto, git |
| **Clinical Genomics** | W3 | GWAS & post-GWAS | GWAS Catalog, Open Targets |
|  | W4 | PRS & clinical application | PGS Catalog, PRSice-2 |
| **Single-cell** | W5 | scRNA-seq technology, QC, clustering | Seurat, scanpy |
|  | W6 | Trajectory & cell–cell communication | Monocle3, CellChat |
| **Spatial** | W7 | Spatial transcriptomics — the basics | Seurat, Squidpy |
|  | W8 | Spatial + scRNA integration, deconvolution | CARD, SPOTlight |
| **Proteomics** | W9 | MS- and NGS-based proteomics | theory |
|  | W10 | Clinical proteomics & biomarkers | limma, MSstats |
| **Integration** | W11 | Multi-omics integration | MOFA+, mixOmics |
|  | W12 | Network analysis & drug targets | STRING, igraph |
| **Capstone** | W13 | Interpretation & visualisation | ggplot2, Quarto |
|  | W14 | Individual project design | — |
:::

## 3) What you should be able to do in December {#w1-s2-3}

-   ✅ Read a methods section and say **what was actually done**, including what was quietly not done
-   ✅ Look up whether public data already answers your question, before writing a grant to generate it
-   ✅ Sit in a study-design meeting and stop an unrecoverable mistake
-   ✅ Say why a published finding might not hold in Korean patients
-   ✅ Present a data-supported research proposal at the final session

::: {.callout-note collapse="true"}
## What this course deliberately does not promise

❌ That you will be able to run a spatial deconvolution pipeline unsupervised.

That is a reasonable thing to want, and it is not achievable in fourteen 2-hour sessions alongside clinical duties. Anyone promising it is selling something.

→ What is achievable: never again being the person in the room who cannot evaluate the analysis their own name will appear on.
:::

------------------------------------------------------------------------

# 3. Before Any Data Exists ⏱️ {#w1-s3}

This section is the most immediately useful thing in the semester, and it involves no software.

## 1) ⚠️ The timing problem {#w1-s3-1}

![Almost every irreversible decision happens before anyone opens R.](/assets/img/w1/when_to_call.svg){fig-align="center"}

::: {.callout-warning collapse="true"}
## What "we'll figure out the analysis later" actually costs

Three real patterns:

-   ① **"We have n = 3 per group, is that enough?"** — asked after collection. For a well-powered bulk comparison in a homogeneous system, sometimes. For single-cell across conditions, ⚠️ your n is the number of **donors**, not cells (W5), so the answer is usually no
-   ② **"The tissue was flash-frozen."** — then scRNA-seq is off the table and it is snRNA-seq or nothing (W5). Not a disaster, but it changes what you can claim
-   ③ **"Cases were collected at hospital A, controls at hospital B."** — see §3.3. There is no fix

→ Each of these takes 30 seconds to prevent and 18 months to regret.
:::

## 2) Three questions that decide everything {#w1-s3-2}

Ask these before a single specimen is collected.

**① What is my unit of replication?**

The thing you can independently repeat. Usually **the patient** — not the cell, not the well, not the technical replicate.

-   ⚠️ 30,000 cells from 3 patients is **n = 3**
-   ⚠️ Two aliquots of the same tumour is **n = 1**

**② What is the comparison, exactly?**

Write the sentence before collecting: *"I will compare \_\_\_ to \_\_\_, and I expect \_\_\_ to differ in direction \_\_\_."*

-   ❌ "I want to profile my patients" — not a comparison
-   ✅ "I will compare treatment-resistant to treatment-naive, matched on age and stage" — now the design follows

**③ What else differs between my groups?**

Everything that tracks with your grouping is a rival explanation: collection site, storage duration, sequencing run, extraction day, operator, time of day, medication.

## 3) ⚠️ Confounding by batch {#w1-s3-3}

![Both layouts cost the same. Only one of them can be analysed.](/assets/img/w1/batch_design.svg){fig-align="center"}

::: {.callout-important collapse="true"}
## Why "we'll just correct for batch" does not save you

Batch correction estimates the batch effect from samples that share a batch but differ in group. If **no batch contains both groups**, there is nothing to estimate from.

The mathematics is not being difficult — the information genuinely is not in the data. `ComBat`, `limma::removeBatchEffect`, Harmony and every successor have the same limitation.

✅ The fix is free, and it is scheduling: **make sure every run, every extraction day, and every site contains both groups.**

💡 If a collaborator sends you data where this was violated, saying so early is a kindness. Saying so after they have written the discussion section is not.
:::

## 4) What your specimen can become {#w1-s3-4}

| Specimen | scRNA-seq | snRNA-seq | Bulk RNA | Spatial | DNA / genotyping |
|--------------------|:----------:|:----------:|:----------:|:-------------:|:---------------:|
| **Fresh** (\< a few hours) | ✅ | ✅ | ✅ | ✅ | ✅ |
| **Flash-frozen (OCT)** | ❌ | ✅ | ✅ | ✅ | ✅ |
| **FFPE block** | ❌ | ⚠️ limited | ⚠️ degraded | ✅ probe-based | ⚠️ fragmented |
| **Whole blood / buffy coat** | ✅ (PBMC) | — | ✅ | — | ✅ |

💡 **The archive question worth asking your department:** how many FFPE blocks with linked outcome data are sitting in pathology? Probe-based spatial and proteomics both work on FFPE (W7, W9) — a retrospective cohort you already own often beats a prospective one you have to fund.

------------------------------------------------------------------------

# 4. A Setup That Survives Three Years 🛠️ {#w1-s4}

## 1) Reproducibility is a clinical problem {#w1-s4-1}

The gap between analysis and publication is typically 1–3 years. Reviewers ask for a reanalysis. A package updated. Your laptop was replaced.

Three habits cover almost all of it:

1.  **Project-based working** — never `setwd()` again
2.  **A consistent directory structure** — you know where everything lives, and so does a collaborator
3.  **Locked package versions** — the same code gives the same numbers later

## 2) R Project (.Rproj) {#w1-s4-2}

An `.Rproj` file fixes the working directory to wherever it sits. Paths built with `here::here()` then work on any machine.

```         
RStudio → File → New Project → New Directory (or Existing Directory)
```

**A structure that scales:**

```         
my_project/
├── my_project.Rproj          ← working directory anchors here
├── rawdata/                  ← never modified, never overwritten
│   └── clinical_data.xlsx
├── scripts/                  ← numbered, in run order
│   ├── 1_clean.R
│   ├── 2_analysis.R
│   └── 3_figures.R
├── analysis/                 ← intermediate .rds objects
├── results/
│   ├── figures/
│   └── tables/
└── logs/
```

💡 The numbering is the point. `2_analysis.R` cannot be run before `1_clean.R`, and six months later that ordering is the only documentation you will trust.

## 3) renv — version locking {#w1-s4-3}

```{r}
# Once, when starting a project
renv::init()

# Install packages as usual
install.packages("tidyverse")
BiocManager::install("DESeq2")

# Lock the current versions into renv.lock
renv::snapshot()

# On a new machine, or for a collaborator
renv::restore()
```

::: {.callout-note collapse="true"}
## Why this matters more than it sounds

A Seurat major-version update can silently change default clustering behaviour. Without a lock file, "I reran it and got different clusters" has no explanation and no fix.

⚠️ `renv::snapshot()` is only useful if you actually run it. Make it the last line of every analysis session.
:::

------------------------------------------------------------------------

# 5. Hands-On 💻 {#w1-s5}

## 1) No code: does your question already have an answer? {#w1-s5-1}

Before generating anything, spend twenty minutes in the browser. Pick a gene or disease from your own specialty and work through:

| Resource | Ask it |
|-----------------------------|-------------------------------------------|
| [**GWAS Catalog**](https://www.ebi.ac.uk/gwas) | Has anyone found variants for this disease? How many, and in whom? |
| [**Open Targets**](https://platform.opentargets.org) | Is my gene already linked to this disease? Is there a drug? |
| [**GTEx**](https://gtexportal.org) | Which tissues express my gene at all? |
| [**CELLxGENE**](https://cellxgene.cziscience.com) | Which cell types express it? *(often answers the whole question)* |
| [**GEO**](https://www.ncbi.nlm.nih.gov/geo/) | Has someone already run this experiment and deposited it? |

✅ Write down, in one sentence each, what you found — and whether your planned experiment still adds something.

💡 This is not a warm-up exercise. It is the single highest-yield hour in translational research, and almost nobody spends it.

## 2) Set up your project {#w1-s5-2}

```         
File → New Project → New Directory → New Project
Name: advanced_BI
```

```{r}
dirs <- c("rawdata", "analysis",
          "results/figures", "results/tables",
          "scripts", "logs")
sapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)

install.packages("renv")
renv::init()
```

## 3) Install the core packages {#w1-s5-3}

```{r}
install.packages(c("tidyverse", "BiocManager", "here",
                   "patchwork", "data.table"))

BiocManager::install(c("DESeq2", "SingleR", "celldex"))

renv::snapshot()
```

::: {.callout-note collapse="true"}
## If installation fails

-   Bioconductor installs are slow — several minutes is normal, not an error
-   ⚠️ Version mismatch is the usual culprit. `BiocManager::valid()` tells you what is out of step
-   Heavier packages (Seurat, ArchR) are installed in the week we need them, not now
:::

## 4) Before next week {#w1-s5-4}

**W2 — R scripts, AI-assisted coding, and work that survives review.** Bring a script you have actually written, however messy. We will make it reproducible together.

------------------------------------------------------------------------

# 📝 Summary

-   Bulk RNA-seq answers one question about a **mixture**. Six questions it cannot answer define the rest of the semester
-   The five pipeline stages do not change across technologies — ⚠️ and stages ③ ④ fail **silently**
-   ⚠️ **Execution depth** belongs to your analyst. **Judgment depth** belongs to you, and cannot be delegated
-   The semester is one clinical question pursued through eight layers until it becomes testable
-   ⚠️ Almost every irreversible decision — n, sample type, batching — happens **before anyone opens R**
-   Your unit of replication is usually the **patient**. 30,000 cells from 3 people is n = 3
-   ⚠️ **Group perfectly confounded with batch cannot be corrected.** Not by ComBat, not by Harmony, not by anything
-   Fresh vs frozen vs FFPE decides which assays remain possible, permanently
-   💡 The FFPE archive your department already owns may beat the cohort you were going to fund
-   `.Rproj` + a fixed folder structure + `renv::snapshot()` is the whole reproducibility story

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|----------------------------|--------------------------------------------|
| Unit of replication | The thing you can independently repeat — usually the patient |
| Technical replicate | Same sample measured twice; does not increase n |
| Biological replicate | A different individual; this is your n |
| Batch effect | Systematic difference caused by processing, not biology |
| Confounding | Two explanations that the data cannot separate |
| Judgment depth | Being able to tell when an analysis is wrong |
| snRNA-seq | Profiling nuclei — the option for frozen tissue (W5) |
| FFPE | Formalin-fixed paraffin-embedded — the pathology archive format |
| `.Rproj` | RStudio project file; anchors the working directory |
| `here::here()` | Builds paths from the project root, portable across machines |
| `renv` | Records and restores exact package versions |
| Quarto | Literate document format; code and prose in one file (W2) |

------------------------------------------------------------------------

# ➕ References

### Study design and its failures

Lazic SE. [*Experimental Design for Laboratory Biologists.*](https://doi.org/10.1017/9781139696647) Cambridge University Press. 2016. (the clearest treatment of replication units)\
Leek JT, et al. [*Tackling the widespread and critical impact of batch effects in high-throughput data.*](https://doi.org/10.1038/nrg2825) Nat Rev Genet. 2010.\
Blainey P, Krzywinski M, Altman N. [*Points of significance: replication.*](https://doi.org/10.1038/nmeth.2921) Nat Methods. 2014.

### Reproducible practice

Sandve GK, Nekrutenko A, Taylor J, Hovig E. [*Ten simple rules for reproducible computational research.*](https://doi.org/10.1371/journal.pcbi.1003285) PLoS Comput Biol. 2013.\
Wilson G, et al. [*Good enough practices in scientific computing.*](https://doi.org/10.1371/journal.pcbi.1005510) PLoS Comput Biol. 2017.\
[R for Data Science (2nd ed.)](https://r4ds.hadley.nz) · [renv documentation](https://rstudio.github.io/renv/)

### Public data first

Barrett T, et al. [*NCBI GEO: archive for functional genomics data sets.*](https://doi.org/10.1093/nar/gks1193) Nucleic Acids Res. 2013.\
Ochoa D, et al. [*The next-generation Open Targets Platform.*](https://doi.org/10.1093/nar/gkac1046) Nucleic Acids Res. 2023.\
[GWAS Catalog](https://www.ebi.ac.uk/gwas) · [GTEx Portal](https://gtexportal.org) · [CELLxGENE](https://cellxgene.cziscience.com)
