---
title: "W2. R Scripts, AI-Assisted Coding & Reproducibility"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Write an analysis script that you (or a reviewer) can rerun in three years
-   Split an analysis into numbered files with one job each
-   Write a prompt that produces useful code instead of plausible code
-   ⚠️ Apply a verification gate to code you did not write — and know why "it ran" proves nothing
-   ⚠️ Recognise the three ways generated code fails **silently**
-   Produce a Quarto report where every number in the text is computed, not typed
-   Use the five git commands that actually matter for a clinical project
:::

> 🤔 **Motivating question**
>
> An AI writes you 40 lines that run without error and produce a beautiful volcano plot. You cannot fully explain line 23. Do you use the figure?

------------------------------------------------------------------------

# 1. Code That Outlives the Analysis 📜 {#w2-s1}

## 1) The reviewer arrives 14 months later {#w2-s1-1}

Analysis code written under deadline has a predictable shape, and a predictable fate.

![The right-hand file costs about ten extra minutes to write.](/assets/img/w2/script_anatomy.svg){fig-align="center"}

⚠️ Note which detail actually kills the left-hand script. Not the messy names — the **undocumented exclusion**. `df_final[-c(3,17,22), ]` is a scientific decision with no recorded reason, and it is the first thing a careful reviewer will ask about.

## 2) One file, one job {#w2-s1-2}

```
scripts/
├── 1_clean.R       rawdata/  →  analysis/1_clean.rds
├── 2_deseq2.R      analysis/1_clean.rds  →  analysis/2_dds.rds
├── 3_enrich.R      analysis/2_dds.rds    →  results/tables/
└── 4_figures.R     analysis/*            →  results/figures/
```

Every script starts with a header stating what goes in and what comes out:

```{r}
# ============================================================
# 2_deseq2.R — differential expression
# Project : IBD biopsy cohort
# In      : analysis/1_clean.rds
# Out     : analysis/2_dds.rds, results/tables/de_results.tsv
# ============================================================
```

💡 The numbering is documentation. `3_enrich.R` cannot run before `2_deseq2.R`, and in eighteen months that ordering is the only description of the pipeline you will trust.

## 3) Named sections {#w2-s1-3}

Four trailing dashes turn a comment into a navigable section (⌘/Ctrl + Shift + R inserts one):

```{r}
# Libraries ----
library(DESeq2)

# Load ----
dds <- readRDS(here("analysis", "1_clean.rds"))

# Filter — 3 samples excluded, RIN < 6 ----
dds <- dds[, dds$RIN >= 6]

# Test ----
dds <- DESeq(dds)
```

✅ Sections appear in RStudio's outline pane. ✅ And note the third one: the **reason lives in the section title**, where it cannot be separated from the code.

## 4) Functions — when to bother {#w2-s1-4}

The rule is not "always write functions." It is: **the third time you copy-paste, stop.**

```{r}
# The pattern that goes wrong — copy-paste per cell type
cd8 <- FindMarkers(seurat, ident.1 = "CD8 T", min.pct = 0.25)
cd4 <- FindMarkers(seurat, ident.1 = "CD4 T", min.pct = 0.25)
nk  <- FindMarkers(seurat, ident.1 = "NK",    min.pct = 0.5)   # ⚠️ typo, or intentional?

# Write it once
get_markers <- function(obj, cell_type, min_pct = 0.25) {
  FindMarkers(obj, ident.1 = cell_type, min.pct = min_pct) |>
    as_tibble(rownames = "gene") |>
    mutate(cell_type = cell_type)
}

all_markers <- map_dfr(c("CD8 T", "CD4 T", "NK"), \(x) get_markers(seurat, x))
```

⚠️ The point is not elegance. In the copy-pasted version, `min.pct = 0.5` on the third line is invisible — and it silently changes what "NK markers" means.

------------------------------------------------------------------------

# 2. Working With AI ⚙️ {#w2-s2}

## 1) What actually changed {#w2-s2-1}

**Vibe coding** — describing what you want in natural language and iterating on what comes back — became a mainstream working style in 2025. For clinicians it removes a real barrier: syntax was never the interesting part of your question.

| It genuinely removed | It did not remove |
|--------------------------------|----------------------------------------|
| Remembering `dplyr` syntax | Knowing whether the analysis is appropriate |
| Boilerplate plotting code | Knowing what the numbers should look like |
| Reading error messages | ⚠️ Noticing when there is **no** error message |
| Translating between R and Python | Responsibility for what you publish |

## 2) A prompt that works {#w2-s2-2}

Three components: **context, task, constraints.**

```
Context: A Seurat object `seurat` with cell types in `cell_type`
         and disease status in `condition` (levels: "IBD", "control").
         Sample identity is in `sample_id`, 12 donors total.

Task:    Write a pseudobulk differential expression function that
         takes a cell type name, aggregates counts per donor,
         and returns a DESeq2 results tibble with gene symbols.

Constraints: dplyr + DESeq2 + here. Return a tibble sorted by padj.
             Comment any step where a design choice is being made.
```

❌ Compare: `"write code for my analysis"` — you will get code that runs and answers a different question.

💡 That last constraint line is worth adding every time. Forcing the model to mark its own design choices makes them visible to you.

## 3) ⚠️ The gate {#w2-s2-3}

![Everything in this course depends on step ③ actually happening.](/assets/img/w2/ai_verification.svg){fig-align="center"}

::: {.callout-important collapse="true"}
## Why "it ran" is worthless as evidence

Traditional coding errors are **loud** — the script stops. That trained a whole generation to treat "no error" as "correct."

Generated code inverts this. It is fluent, idiomatic, and usually syntactically perfect. When it is wrong, it is wrong **semantically**: it does a coherent thing that is not the thing you asked for.

-   ❌ A crash is a gift. It tells you exactly where to look
-   ⚠️ A plausible data frame of the right dimensions is the dangerous outcome

→ Your verification cannot be "did it run." It has to be **"is the output the shape and magnitude I expected?"** — which requires you to have formed an expectation first.
:::

## 4) ⚠️ Three silent failures {#w2-s2-4}

![All three of these run cleanly, return the right class of object, and are wrong.](/assets/img/w2/silent_failures.svg){fig-align="center"}

::: {.callout-tip collapse="true"}
## The habit that catches most of them

After every join, filter, or relabel, print the shape:

```{r}
nrow(df)                    # did I lose rows?
sum(is.na(df$key_column))   # did the join fail quietly?
table(df$group)             # are both groups still here?
```

Three lines. They will catch more real errors in your career than any statistical method you learn this semester.

💡 Ask the model to insert these itself: *"after each join, add a line printing the row count and NA count."* It will comply, and now the check is in the script permanently.
:::

## 5) Where it is reliable, and where it is not {#w2-s2-5}

-   ✅ **Reliable** — plotting code, data reshaping, explaining an error message, translating between languages, writing the boilerplate around a well-known package
-   ⚠️ **Check carefully** — anything involving a statistical choice, a multiple-testing correction, a join key, or a package that changed API recently
-   ❌ **Do not delegate** — deciding what the comparison is, deciding what n means, deciding whether the result is biologically plausible

------------------------------------------------------------------------

# 3. Reproducible Documents 📄 {#w2-s3}

## 1) The copy-paste problem {#w2-s3-1}

![One filter line versus five manual edits and a wrong abstract.](/assets/img/w2/copy_paste_vs_quarto.svg){fig-align="center"}

## 2) Quarto, minimally {#w2-s3-2}

A `.qmd` file is prose with executable code blocks. Render it and you get HTML, PDF, or Word.

````
---
title: "IBD biopsy cohort — DE analysis"
format: html
---

## Results

```{{r}}
#| echo: false
library(tidyverse)
res <- readRDS(here::here("analysis", "2_dds.rds"))
sig <- res |> filter(padj < 0.05)
```

We identified `r knitr::inline_expr("nrow(sig)")` differentially expressed
genes (FDR < 0.05), of which `r knitr::inline_expr("sum(sig$log2FoldChange > 0)")` were upregulated.
````

```
quarto render report.qmd --to html
quarto render report.qmd --to docx     # for the co-author who wants Word
```

✅ The inline `r knitr::inline_expr("nrow(sig)")` is the whole idea: **the number in the sentence is computed, not typed.** Change the data, re-render, and the sentence is still true.

::: {.callout-note collapse="true"}
## Useful chunk options

Written as a comment at the top of a chunk, prefixed with a hash and a pipe:

| Option | Effect |
|------------------------|------------------------------------------------|
| `echo: false` | run the code, hide it — for a clinical audience |
| `eval: false` | show the code, don't run it — for teaching |
| `warning: false` | suppress package chatter |
| `fig-cap: "..."` | attach a caption, so the figure travels with its meaning |
| `cache: true` | don't re-run a slow chunk every render |

💡 This course website is a Quarto project. Every page you are reading is a `.qmd`.
:::

------------------------------------------------------------------------

# 4. Version Control, Minimally 🔀 {#w2-s4}

## 1) Why a clinician should care {#w2-s4-1}

Not because you are a software developer. Because:

-   ✅ **"When did this number change, and why?"** — answerable in seconds instead of never
-   ✅ **`analysis_final_v3_REAL.R` stops existing** — there is one file, with a history
-   ✅ **A reviewer or collaborator can run your code** without you emailing a zip file
-   ✅ **GitHub Pages hosts your report at a URL** you can put in a paper

## 2) The five commands {#w2-s4-2}

```{bash}
# once per machine
git config --global user.name  "Eunji Ha"
git config --global user.email "you@knu.ac.kr"

# once per project
git init

# after each working session — these three, forever
git add .
git commit -m "Add DESeq2 analysis; exclude 3 low-RIN samples"
git push
```

⚠️ **Never commit patient data.** Put `rawdata/` in `.gitignore` before the first commit:

```
rawdata/
*.rds
.Rhistory
.Rproj.user/
renv/library/
```

::: {.callout-warning collapse="true"}
## Once it is in the history, it is in the history

Deleting a file in a later commit does **not** remove it from the repository — the earlier version is still recoverable, and on a public repo, still public.

✅ Add `.gitignore` **first**, then `git init`, then commit. In that order.

⚠️ If identifiable data has already been pushed to a public repository, treat it as a disclosure event and involve your IRB. Rewriting history is not sufficient.
:::

## 3) Publishing {#w2-s4-3}

```{bash}
quarto publish gh-pages
```

Renders your `.qmd` files, pushes the HTML, and serves it at `https://<username>.github.io/<repo>/`.

------------------------------------------------------------------------

# 5. Hands-On 💻 {#w2-s5}

## 1) Rescue a script of your own {#w2-s5-1}

Bring a real script — the messier the better. In 20 minutes:

-   ① Add the header block: project, in, out
-   ② Insert `# Section ----` markers and open the outline pane
-   ③ Find one unexplained filter or exclusion. **Write its reason into the section title**
-   ④ Replace every absolute path with `here::here()`
-   ⑤ `renv::snapshot()`

## 2) ⚠️ Verify code you did not write {#w2-s5-2}

We will give you a 30-line script that runs cleanly and contains one planted error of the type in §2.4. Your task is to find it **without being told which line**.

The only reliable route:

```{r}
# Insert shape checks between every step, then read the numbers
cat("rows before join:", nrow(clin), "\n")
merged <- left_join(clin, expr, by = "sample_id")
cat("rows after join :", nrow(merged), "\n")
cat("unmatched       :", sum(is.na(merged$gene_1)), "\n")
```

💡 Time yourself. Most people find it in under five minutes **with** the checks, and not at all without them.

## 3) One Quarto report {#w2-s5-3}

Take any result you already have and write a one-page `.qmd` in which **no number is typed by hand**. Render to HTML and to Word.

## 4) Before next week {#w2-s5-4}

**W3 — GWAS & post-GWAS.** No installation needed; we start in the browser.

Come with **one disease you actually treat**. We will look it up together in the GWAS Catalog and Open Targets.

------------------------------------------------------------------------

# 📝 Summary

-   A script's fatal flaw is usually not messiness — it is an **undocumented exclusion**
-   One file, one job, numbered in run order, with an in/out header
-   Put the *reason* for a decision in the section title, where it cannot detach from the code
-   Prompts need **context + task + constraints**; ask the model to flag its own design choices
-   ⚠️ **"It ran without error" is not evidence.** Generated code fails semantically, not syntactically
-   ⚠️ Three silent failures: the mismatched join, adjustment on a pre-filtered list, the relabel that drops a level
-   ✅ Print `nrow()`, `sum(is.na())`, `table()` after every join, filter and relabel. Every time
-   AI is reliable for plotting and reshaping; ❌ never for deciding the comparison, the n, or the plausibility
-   Quarto makes every number in the text **computed**, so a re-run cannot leave a stale figure in the abstract
-   ⚠️ `.gitignore` before `git init`. Patient data in a git history stays in the history

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| Vibe coding | Writing code by describing intent to a language model and iterating |
| Silent failure | Code that runs, returns the right type, and computes the wrong thing |
| Shape check | Printing row counts / NA counts to confirm an operation did what you meant |
| Literate document | Prose and executable code in one file (Quarto, R Markdown) |
| Inline code | `r knitr::inline_expr("expr")` — a value computed into a sentence at render time |
| Chunk option | Per-block setting: `echo`, `eval`, `cache`, `fig-cap` |
| Commit | A saved snapshot of the project with a message explaining why |
| `.gitignore` | List of files git must never track — put raw data here |
| GitHub Pages | Free static hosting; `quarto publish gh-pages` deploys to it |
| `here::here()` | Project-root-relative paths that work on any machine |

------------------------------------------------------------------------

# ➕ References

### Writing code that lasts

Wilson G, et al. [*Good enough practices in scientific computing.*](https://doi.org/10.1371/journal.pcbi.1005510) PLoS Comput Biol. 2017.\
Sandve GK, Nekrutenko A, Taylor J, Hovig E. [*Ten simple rules for reproducible computational research.*](https://doi.org/10.1371/journal.pcbi.1003285) PLoS Comput Biol. 2013.\
[R for Data Science (2nd ed.)](https://r4ds.hadley.nz) — the standard reference for modern R style

### Reproducible reporting

[Quarto documentation](https://quarto.org) · [Quarto → GitHub Pages](https://quarto.org/docs/publishing/github-pages.html)\
Rule A, et al. [*Ten simple rules for writing and sharing computational analyses in Jupyter Notebooks.*](https://doi.org/10.1371/journal.pcbi.1007007) PLoS Comput Biol. 2019. (the principles transfer directly to Quarto)

### Version control without pain

Bryan J. [*Happy Git and GitHub for the useR.*](https://happygitwithr.com) — written for R users, not developers\
Perez-Riverol Y, et al. [*Ten simple rules for taking advantage of Git and GitHub.*](https://doi.org/10.1371/journal.pcbi.1004947) PLoS Comput Biol. 2016.

### On working with generated code

Karpathy A. *"Vibe coding"* (2025) — the coinage, and worth reading in the original for how explicitly it describes **not** reading the code. This chapter is the argument for why a clinical researcher cannot work that way.
