---
title: "Spatial Transcriptomics"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 10"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W10 — Spatial Transcriptomics"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Why Spatial? {background-color="#1D9E75"}

---

## The Problem with Everything We've Done So Far

| Method | What we lose |
|--------|-------------|
| Bulk RNA-seq | Cell type identity, spatial context |
| scRNA-seq (W8–W9) | **Where in the tissue?** |
| GWAS (W2–W3) | Cell type, tissue architecture |

::: {.fragment}
**The fundamental problem:**

Dissociating tissue into single cells destroys the spatial information.  
You know *what* cells are doing — but not *where*, *next to what*, and *in which microenvironment*.
:::

---

## What Spatial Transcriptomics Adds

:::: {.columns}

::: {.column width="50%"}
**scRNA-seq tells you:**

- This is a macrophage
- It's in an inflammatory state
- It's talking to a T cell
:::

::: {.column width="50%"}
**Spatial adds:**

- This macrophage is at the **tumor invasion front**
- It's **5 cell-widths from a blood vessel**
- The neighboring tumor cells are **hypoxic**
:::

::::

::: {.fragment}
> For a pathologist or oncologist, this is exactly how you think about tissue.  
> Spatial transcriptomics is **digital pathology with transcriptomics layered on top.**
:::

---

## Clinical Scenarios Where This Matters

::: {.incremental}
- **Tumor microenvironment** — immune exclusion zone vs. inflamed phenotype → immunotherapy response
- **Tissue borders** — where does normal end and dysplasia begin? Molecularly.
- **Fibrosis** — which cells drive progression at the fibrotic front?
- **Brain** — layer-specific gene expression in neurodegeneration
- **Gut** — crypt-villus axis disruption in IBD
- **Cardiac** — spatial heterogeneity in ischemic zones
:::

---

# Technologies {background-color="#534AB7"}

---

## The Landscape (2024)

| Platform | Resolution | Genes | Throughput | Best for |
|----------|-----------|-------|-----------|---------|
| **10x Visium** | ~55 µm spot | Whole transcriptome | High | Discovery, large tissue |
| **10x Xenium** | Single-cell | 100–500 genes | Very high | Validation, FFPE |
| **NanoString CosMx** | Single-cell | 1,000+ genes | High | Clinical FFPE samples |
| **Slide-seq v2** | ~10 µm | Whole transcriptome | Medium | Mouse brain |
| **MERFISH / seqFISH** | Single-cell | 100–10,000 genes | Low | Mechanistic studies |

::: {.fragment}
::: {.callout-tip}
## For clinical research
**Visium** for discovery (fresh-frozen).  
**Xenium / CosMx** for FFPE validation — the samples you already have in your biobank.
:::
:::

---

## How Visium Works

```
Fresh-frozen tissue section (10 µm)
         ↓
Placed on capture array
(6,500 spots, each 55 µm, containing barcoded oligos)
         ↓
Permeabilization → mRNA released → captured by oligos
         ↓
cDNA synthesis + sequencing
         ↓
Each spot: ~10–20 cells + (x, y) coordinates + expression profile
         ↓
H&E image aligned to array
```

::: {.fragment}
The result: a **gene expression matrix** where every row has a tissue coordinate.
:::

---

## The Data Structure

```
Seurat object with spatial data:
├── counts matrix   (spots × genes)  e.g., 3,000 spots × 30,000 genes
├── spot coordinates (x, y) on tissue
├── H&E image
└── scalefactors    (pixel → µm conversion)
```

```{r}
library(Seurat)
library(here)

# Load 10x Visium output
visium <- Load10X_Spatial(
  data.dir   = here("data", "raw", "visium_output"),
  filename   = "filtered_feature_bc_matrix.h5",
  assay      = "Spatial",
  slice      = "slice1",
  filter.matrix = TRUE
)

visium
```

---

# Analysis Workflow {background-color="#185FA5"}

---

## Overview

```
Raw Visium data
    │
    ├── 1. QC           → filter low-quality spots
    ├── 2. Normalization → SCTransform
    ├── 3. Dim reduction → PCA → UMAP
    ├── 4. Clustering   → spatial domains
    ├── 5. SVGs         → spatially variable genes
    ├── 6. Deconvolution → cell type per spot
    └── 7. Integration  → scRNA-seq reference
```

---

## Step 1: QC

```{r}
# Mitochondrial gene percentage
visium[["percent.mt"]] <- PercentageFeatureSet(visium,
                                               pattern = "^MT-")

# Visualize QC metrics on tissue
SpatialFeaturePlot(visium,
                   features = c("nCount_Spatial",
                                "nFeature_Spatial",
                                "percent.mt"))
```

```{r}
# Filter spots
visium <- subset(visium,
                 nCount_Spatial   > 500    &
                 nFeature_Spatial > 200    &
                 percent.mt       < 25)

cat("Spots remaining:", ncol(visium), "\n")
```

---

## Step 2: Normalization — SCTransform

```{r}
# SCTransform — accounts for technical variation
# (sequencing depth varies across spots)
visium <- SCTransform(visium,
                      assay      = "Spatial",
                      verbose    = FALSE,
                      vars.to.regress = "percent.mt")
```

::: {.callout-note}
## Why not standard normalization?
Spatial data has strong sequencing depth variation across the tissue section  
(edge spots often have fewer counts).  
SCTransform handles this better than simple log-normalization.
:::

---

## Step 3: Dimensionality Reduction & Clustering

```{r}
visium <- RunPCA(visium, assay = "SCT", verbose = FALSE)
visium <- FindNeighbors(visium, reduction = "pca", dims = 1:30)
visium <- FindClusters(visium, verbose = FALSE, resolution = 0.5)
visium <- RunUMAP(visium, reduction = "pca", dims = 1:30)

# Plot clusters in UMAP space AND on tissue
p1 <- DimPlot(visium, reduction = "umap", label = TRUE)
p2 <- SpatialDimPlot(visium, label = TRUE, label.size = 3)

p1 + p2
```

::: {.fragment}
Key question: do the **UMAP clusters** correspond to **anatomical regions**?  
If yes → the spatial domains are transcriptionally distinct.
:::

---

## Step 4: Spatially Variable Genes (SVGs)

Which genes vary across tissue space — not just between clusters, but continuously?

```{r}
library(NNSVG)

# NNSVG: fast, scalable SVG detection
# Fits a spatial model per gene
nnsvg_results <- nnsvg(
  x        = counts(visium),
  coords   = spatialCoords(visium),
  n_threads = 4
)

# Top SVGs
top_svgs <- nnsvg_results |>
  arrange(desc(LR_stat)) |>
  slice_head(n = 20)

SpatialFeaturePlot(visium,
                   features = top_svgs$gene_name[1:6],
                   ncol = 3)
```

---

## Step 5: Deconvolution — Cell Types per Spot

Each Visium spot contains **10–20 cells** → what mix of cell types?

```{r}
library(STdeconvolve)

# Reference: single-cell RNA-seq from same tissue type
# (from W8–W9 analysis — this is why sc comes first!)
counts_spatial <- GetAssayData(visium, slot = "counts",
                               assay = "Spatial")

# Run deconvolution
ldaOut <- fitLDA(t(as.matrix(counts_spatial)),
                 Ks = seq(4, 9, by = 1))

# Optimal K
optLDA <- optimalModel(models = ldaOut, opt = "min")
results <- getBetaTheta(optLDA,
                        perc.filt = 0.05,
                        betaScale = 1000)

# Map cell type proportions back to tissue
vizAllTopics(theta        = results$theta,
             pos          = spatialCoords(visium),
             topicOrder   = seq(ncol(results$theta)),
             topicCols    = rainbow(ncol(results$theta)))
```

---

## Step 6: Integration with scRNA-seq Reference

```{r}
# Transfer cell type labels from scRNA-seq reference
# (Seurat label transfer)

# Load your W8-W9 scRNA-seq reference
sc_ref <- readRDS(here("data", "processed", "scrna_annotated.rds"))

# Find anchors between spatial and single-cell data
anchors <- FindTransferAnchors(
  reference      = sc_ref,
  query          = visium,
  normalization.method = "SCT"
)

# Transfer labels
predictions <- TransferData(
  anchorset    = anchors,
  refdata      = sc_ref$celltype,
  prediction.assay = TRUE,
  weight.reduction = visium[["pca"]],
  dims         = 1:30
)

visium[["predictions"]] <- predictions
```

---

## Visualizing Cell Type Predictions on Tissue

```{r}
# Which cell types land where?
DefaultAssay(visium) <- "predictions"

SpatialFeaturePlot(
  visium,
  features  = c("Tumor", "Macrophage", "T_cell",
                "Fibroblast", "Endothelial"),
  pt.size.factor = 1.5,
  ncol      = 3,
  crop      = TRUE
)
```

::: {.fragment}
Now you can ask:  
**"Where are the immune-excluded tumors, and what genes define that border?"**
:::

---

# Clinical Applications {background-color="#0F6E56"}

---

## Tumor Microenvironment — Immune Exclusion

```{r}
library(tidyverse)

# Score each spot for immune infiltration
visium <- AddModuleScore(
  visium,
  features = list(
    immune_score  = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG"),
    exclusion_score = c("ACTA2", "FAP", "COL1A1", "POSTN")
  ),
  name = c("immune", "exclusion")
)

# Map scores onto tissue
SpatialFeaturePlot(visium,
                   features = c("immune1", "exclusion1"),
                   ncol = 2)
```

::: {.fragment}
**Clinical translation:**  
High immune score + high exclusion score = **immune-excluded phenotype** → likely non-responder to checkpoint inhibitor.
:::

---

## Identifying Spatially Distinct Disease Zones

```{r}
# Example: Colorectal cancer — tumor core vs. invasion front
# Differentially expressed genes between spatial clusters

Idents(visium) <- "seurat_clusters"

# Compare invasion front (cluster 3) vs. tumor core (cluster 1)
invasion_vs_core <- FindMarkers(
  visium,
  ident.1 = "3",     # invasion front cluster
  ident.2 = "1",     # tumor core cluster
  assay   = "SCT"
)

# Top markers of invasion front
invasion_vs_core |>
  filter(p_val_adj < 0.05, avg_log2FC > 1) |>
  arrange(desc(avg_log2FC)) |>
  head(20)
```

---

## Cell–Cell Communication in Spatial Context

```{r}
library(CellChat)

# Unlike standard CellChat (W9), spatial CellChat
# only considers cells within a defined distance

cellchat <- createCellChat(
  object   = visium,
  meta     = visium@meta.data,
  group.by = "predicted.celltype",
  datatype = "spatial",
  coordinates = GetTissueCoordinates(visium),
  spatial.factors = list(ratio = 0.1, tol = 0.08)
)

cellchat <- addMeta(cellchat, meta = visium@meta.data)
cellchat <- setIdent(cellchat, ident.use = "predicted.celltype")
```

::: {.callout-tip}
## Spatial constraint
Standard cell–cell communication assumes all cells can talk to all cells.  
Spatial CellChat only allows interactions between **physically neighboring spots** — far more biologically realistic.
:::

---

## Spatial Trajectory — Along Tissue Axes

```{r}
library(spatialDE)

# Test for expression gradients along a defined tissue axis
# e.g., crypt → villus in intestine, or cortex → medulla in kidney

# Define pseudospatial axis
coords <- GetTissueCoordinates(visium)

# SpatialDE: Gaussian process model for spatial variation
norm_expr <- normalizeExpr(
  counts   = GetAssayData(visium, slot = "counts"),
  method   = "scran"
)

results <- spatialDE::run(norm_expr, sample_info = coords)

# Genes with significant spatial expression patterns
sig_spatial <- results[results$qvalue < 0.05, ]
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: Human Breast Cancer (10x Visium)

Publicly available from 10x Genomics:  
Human Breast Cancer Block A, Section 1

```{r}
library(Seurat)
library(tidyverse)
library(patchwork)
library(here)

# Load data (pre-downloaded)
bc <- Load10X_Spatial(
  data.dir = here("data", "raw", "breast_cancer_visium"),
  filename = "filtered_feature_bc_matrix.h5"
)

# Quick look
cat("Spots:", ncol(bc), "\n")
cat("Genes:", nrow(bc), "\n")
```

---

## Hands-On: QC & Normalization

```{r}
# MT genes
bc[["percent.mt"]] <- PercentageFeatureSet(bc, pattern = "^MT-")

# Visualize QC on tissue
SpatialFeaturePlot(bc, features = "nCount_Spatial") +
  theme(legend.position = "right")

# Filter
bc <- subset(bc,
             nCount_Spatial   > 100 &
             percent.mt       < 30)

# Normalize
bc <- SCTransform(bc, assay = "Spatial", verbose = FALSE)
```

---

## Hands-On: Cluster & Visualize

```{r}
bc <- RunPCA(bc, assay = "SCT", verbose = FALSE)
bc <- FindNeighbors(bc, dims = 1:30)
bc <- FindClusters(bc, resolution = 0.5, verbose = FALSE)
bc <- RunUMAP(bc, dims = 1:30)

# Side-by-side: UMAP + tissue
DimPlot(bc, label = TRUE) +
SpatialDimPlot(bc, label = TRUE, label.size = 3)
```

---

## Hands-On: Find Spatially Variable Genes

```{r}
# Built-in Seurat method (Moran's I)
bc <- FindSpatiallyVariableFeatures(
  bc,
  assay    = "SCT",
  features = VariableFeatures(bc)[1:1000],
  selection.method = "moransi"
)

# Top SVGs
top_svgs <- SpatiallyVariableFeatures(bc, 
                                       selection.method = "moransi")

# Visualize top 6 SVGs on tissue
SpatialFeaturePlot(bc,
                   features = top_svgs[1:6],
                   ncol     = 3,
                   alpha    = c(0.1, 1))
```

---

## Hands-On: Marker Genes per Spatial Cluster

```{r}
# Find marker genes for each spatial cluster
bc_markers <- FindAllMarkers(
  bc,
  assay      = "SCT",
  only.pos   = TRUE,
  min.pct    = 0.25,
  logfc.threshold = 0.5
)

# Top 5 per cluster
bc_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5) |>
  print(n = Inf)

# Visualize a marker on tissue
SpatialFeaturePlot(bc,
                   features = c("ERBB2", "KRT7",
                                "CD8A", "COL1A1"),
                   ncol = 2)
```

---

## Hands-On: Save Results

```{r}
# Save processed object
saveRDS(bc, here("data", "processed", "breast_cancer_spatial.rds"))

# Export cluster annotations for downstream use
cluster_annot <- bc@meta.data |>
  select(barcode = rownames(bc@meta.data),
         cluster = seurat_clusters,
         x = imagerow,
         y = imagecol,
         nCount_Spatial,
         nFeature_Spatial)

write_tsv(cluster_annot,
          here("results", "tables", "spatial_clusters.tsv"))
```

---

# Limitations & Considerations {background-color="#993C1D"}

---

## What to Watch Out For

::: {.incremental}
- **Visium resolution (55 µm)** — each spot is 10–20 cells, not single-cell → need deconvolution
- **Tissue quality** — fresh-frozen preferred; FFPE works for Xenium/CosMx but not standard Visium
- **Capture area** — 6.5 mm × 6.5 mm per section; large tumors need multiple sections
- **Gene dropout** — spatial assays have higher dropout than scRNA-seq
- **Batch effects** — section-to-section variation in permeabilization time matters
- **Alignment** — H&E to array alignment errors propagate into all downstream analysis
:::

---

## GWAS → Spatial: Closing the Loop

Remember the W2 story:

```
GWAS hit: rs12345 near GENE_X (p = 3×10⁻²⁴)
    │
    ├─ eQTL: rs12345 increases GENE_X expression (GTEx)
    ├─ scRNA-seq: GENE_X highly expressed in macrophages (W8)
    └─ Spatial: GENE_X+ macrophages concentrated at tumor front (W10 ← you are here)
    
→ "Macrophages at the invasion front, driven by GENE_X, 
    may facilitate tumor invasion in carriers of rs12345"
→ Testable hypothesis → W13 Capstone
```

---

# Next Week {background-color="#534AB7"}

---

## W11 Preview — Multi-omics Integration & GWAS Gene Prioritization

**What we'll cover:**

- MOFA2: factor analysis across omics layers
- Integrating GWAS, epigenomics, and spatial data
- Gene prioritization: from GWAS locus to candidate gene

**Install before next class:**

```{r}
BiocManager::install("MOFA2")
BiocManager::install("mixOmics")
install.packages("OmicsPLS")
```

---

## Key Resources

- **10x Genomics datasets** — [10xgenomics.com/datasets](https://www.10xgenomics.com/datasets) — free Visium data
- **Seurat Spatial vignette** — [satijalab.org/seurat/articles/spatial_vignette](https://satijalab.org/seurat/articles/spatial_vignette)
- **NNSVG** — [bioconductor.org/packages/NNSVG](https://bioconductor.org/packages/NNSVG)
- **STdeconvolve** — [bioconductor.org/packages/STdeconvolve](https://bioconductor.org/packages/STdeconvolve)
- **Human Cell Atlas Spatial** — [humancellatlas.org](https://www.humancellatlas.org)

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Multi-omics Integration & GWAS Gene Prioritization**

Course materials → [your GitHub Pages URL here]
:::
