---
title: "Epigenomics II: Chromatin Accessibility & 3D Genome"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 5"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W5 — Chromatin & 3D Genome"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Open Chromatin & Gene Regulation {background-color="#1D9E75"}

---

## The Regulatory Genome

Only ~2% of the genome codes for proteins.  
The other 98%? **Largely regulatory.**

::: {.incremental}
- **Promoters** — near TSS, where transcription initiates
- **Enhancers** — distal elements, can be megabases away
- **Insulators** — block enhancer–promoter interactions
- **Silencers** — repress transcription
:::

::: {.fragment}
> **~90% of GWAS hits are non-coding.**  
> To understand what they do, we need to know which regulatory elements are active  
> in which cell types — and how they contact their target genes.
:::

---

## Chromatin Accessibility — The Key Readout

```
Closed chromatin:  DNA wrapped tightly around histones → inaccessible
                   [H][H][H][DNA][H][H][H]
                   Transcription factors cannot bind
                   Gene is OFF

Open chromatin:    nucleosome-depleted region → accessible
                   [H]         [H]
                   Transcription factors can bind
                   Gene is (potentially) ON
```

::: {.fragment}
Open chromatin = **active regulatory elements**  
→ cell-type-specific gene regulation
:::

---

# ATAC-seq {background-color="#534AB7"}

---

## How ATAC-seq Works

```
Nuclei isolation
    ↓
Tn5 transposase treatment
(cuts & inserts sequencing adapters ONLY in open chromatin)
    ↓
PCR amplification
    ↓
Short-read sequencing
    ↓
Reads pile up at open chromatin → "peaks"
```

::: {.fragment}
**Advantages:**
- Works on 500–50,000 cells (vs. DNase-seq needing millions)
- Single-cell compatible (scATAC-seq)
- Fast, cheap, highly reproducible
:::

---

## ATAC-seq Analysis Workflow

```
Raw FASTQ
    │
    ├── 1. QC & trimming         → FastQC, Trimmomatic
    ├── 2. Alignment             → Bowtie2 → hg38
    ├── 3. Post-alignment QC     → TSS enrichment, fragment size
    ├── 4. Peak calling          → MACS2
    ├── 5. Differential peaks    → DESeq2 on peak counts
    ├── 6. TF motif analysis     → chromVAR, HOMER
    └── 7. GWAS integration      → overlap GWAS hits with peaks
```

::: {.callout-note}
Steps 1–4 require a Linux server (pre-computed for this class).  
We start from peak files in R.
:::

---

## Working with Peak Files in R

```{r}
library(ArchR)
library(here)

# ArchR: comprehensive ATAC-seq analysis in R
# Create ArrowFiles from fragment files
ArrowFiles <- createArrowFiles(
  inputFiles  = here("data", "raw", "fragments.tsv.gz"),
  sampleNames = "SampleA",
  minTSS      = 4,       # TSS enrichment score filter
  minFrags    = 1000,    # minimum fragments per cell
  addTileMat  = TRUE,    # 500 bp bins across genome
  addGeneScoreMat = TRUE # gene activity scores
)

proj <- ArchRProject(
  ArrowFiles  = ArrowFiles,
  outputDirectory = here("results", "archr_output"),
  copyArrows  = TRUE
)
```

---

## QC: TSS Enrichment & Fragment Size

```{r}
# TSS enrichment — key QC metric for ATAC-seq
# Good samples: TSS score > 8
plotTSSEnrichment(ArchRProj = proj)

# Fragment size distribution
# Should show nucleosome-free (<200 bp) and mono-nucleosome (~200 bp) peaks
plotFragmentSizes(ArchRProj = proj)

# QC summary
df_qc <- getCellColData(proj, select = c("TSSEnrichment",
                                          "nFrags",
                                          "BlacklistRatio"))
summary(df_qc)
```

---

## Dimensionality Reduction & Clustering

```{r}
# LSI (Latent Semantic Indexing) — standard for ATAC-seq
proj <- addIterativeLSI(
  ArchRProj   = proj,
  useMatrix   = "TileMatrix",
  name        = "IterativeLSI",
  iterations  = 2,
  clusterParams = list(resolution = 0.2, n.start = 10),
  varFeatures = 25000
)

proj <- addClusters(proj, reducedDims = "IterativeLSI",
                    resolution = 0.8)
proj <- addUMAP(proj, reducedDims = "IterativeLSI")

# Visualize
plotEmbedding(proj, colorBy = "cellColData",
              name = "Clusters", embedding = "UMAP")
```

---

## Peak Calling & Differential Accessibility

```{r}
# Call peaks per cluster (pseudo-bulk)
proj <- addGroupCoverages(proj, groupBy = "Clusters")

proj <- addReproduciblePeakSet(
  proj,
  groupBy    = "Clusters",
  pathToMacs2 = "/usr/bin/macs2"
)

proj <- addPeakMatrix(proj)

# Differential accessibility between clusters
markerPeaks <- getMarkerFeatures(
  proj,
  useMatrix   = "PeakMatrix",
  groupBy     = "Clusters",
  bias        = c("TSSEnrichment", "log10(nFrags)"),
  testMethod  = "wilcoxon"
)

# Top peaks per cluster
markerList <- getMarkers(markerPeaks, cutOff = "FDR <= 0.01 & Log2FC >= 1")
```

---

## TF Motif Analysis with chromVAR

```{r}
library(chromVAR)
library(JASPAR2020)
library(TFBSTools)

# Get JASPAR motifs
motifs <- getMatrixSet(JASPAR2020,
                       opts = list(species   = "Homo sapiens",
                                   matrixtype = "PFM"))

# Add motif annotations to ArchR project
proj <- addMotifAnnotations(proj,
                             motifPWMs = motifs,
                             name      = "Motif")

# chromVAR deviation scores — which TFs are active in each cluster?
proj <- addDeviationsMatrix(
  proj,
  peakAnnotation = "Motif",
  force          = TRUE
)

# Plot TF deviations
plotVarDev <- getVarDeviations(proj, name = "MotifMatrix",
                                plot = TRUE)
```

---

# 3D Genome Organization {background-color="#185FA5"}

---

## Why 3D Structure Matters

DNA in the nucleus is not linear — it folds into a hierarchy:

```
Chromosome territory
    └── Compartment A/B (active/inactive, ~Mb scale)
            └── TAD (Topologically Associating Domain, ~100 kb–1 Mb)
                    └── Loop (enhancer–promoter contact, ~10–100 kb)
```

::: {.fragment}
**Clinical relevance:**

- TAD boundary disruption → new enhancer–gene contacts → oncogene activation  
  (e.g., PDGFRA activation in glioma by TAD reorganization)
- Loop formation = enhancer–promoter contact = gene activation
- GWAS hit in an enhancer → which gene's promoter does it contact?
:::

---

## Hi-C — Measuring 3D Contacts

```
Crosslink cells (formaldehyde)
    ↓
Restriction enzyme digestion
    ↓
Proximity ligation (nearby DNA ligated)
    ↓
Sequencing → pairs of genomic loci that were in contact
    ↓
Contact matrix: genomic bins × bins → interaction frequency
```

---

## Loading & Visualizing Hi-C Data

```{r}
library(strawr)
library(tidyverse)
library(here)

# Read .hic file (Juicer format)
hic_file <- here("data", "raw", "sample.hic")

# Extract contact matrix for a region
# Chromosome 17, BRCA1 locus, 25 kb resolution
contacts <- straw(
  norm       = "KR",        # Knight-Ruiz normalization
  fname      = hic_file,
  chr1loc    = "17:41000000:43000000",
  chr2loc    = "17:41000000:43000000",
  unit       = "BP",
  binsize    = 25000
)

# Convert to matrix for visualization
mat <- sparseMatrix(
  i    = contacts$x / 25000 + 1,
  j    = contacts$y / 25000 + 1,
  x    = contacts$counts
)
```

---

## TAD Calling

```{r}
library(HiCcompare)

# TAD boundaries — insulation score method
# Dips in insulation score = TAD boundaries

insulation_score <- function(mat, window = 10) {
  n <- nrow(mat)
  scores <- numeric(n)
  for (i in (window+1):(n-window)) {
    upstream   <- mat[max(1, i-window):(i-1),
                     max(1, i-window):(i-1)]
    downstream <- mat[i:min(n, i+window),
                     i:min(n, i+window)]
    cross      <- mat[max(1, i-window):(i-1),
                     i:min(n, i+window)]
    scores[i]  <- mean(cross) / sqrt(mean(upstream) * mean(downstream))
  }
  scores
}

ins <- insulation_score(as.matrix(mat))

# TAD boundaries = local minima in insulation score
```

---

## Connecting GWAS Hits to Target Genes via 3D Contacts

```{r}
library(GenomicRanges)

# GWAS hit (e.g., CAD hit on chromosome 9p21)
gwas_hit <- GRanges("chr9", IRanges(22100000, 22110000))

# Find which gene promoters contact this GWAS hit locus
# Using published H3K27ac HiChIP loops (active enhancer-promoter)
loops <- import(here("data", "raw", "hichip_loops.bedpe"))

# Loops overlapping the GWAS hit
hit_loops <- loops[queryHits(findOverlaps(loops, gwas_hit))]

# Extract the connected promoters
target_promoters <- resize(hit_loops, width = 1, fix = "end")

# Which genes?
genes <- import(here("data", "raw", "gencode_v38.gtf"))
overlapping_genes <- subsetByOverlaps(genes, target_promoters)
unique(overlapping_genes$gene_name)
```

---

## GWAS + Epigenomics Integration

Full workflow connecting W2 to W4–W5:

```{r}
library(plyranges)

# 1. GWAS hits (from W2)
gwas_gr <- read_tsv(here("results", "gwas_sig_hits.tsv")) |>
  makeGRangesFromDataFrame(seqnames.field = "CHR",
                           start.field    = "BP",
                           end.field      = "BP")

# 2. Open chromatin peaks (from ATAC-seq, this week)
peaks_gr <- import(here("data", "processed", "atac_peaks.bed"))

# 3. Overlap: GWAS hits in open chromatin?
hits_in_peaks <- gwas_gr |>
  join_overlap_inner(peaks_gr)

cat("GWAS hits overlapping open chromatin:",
    length(hits_in_peaks), "/", length(gwas_gr), "\n")

# 4. Which cell type's peaks?
table(hits_in_peaks$cell_type)
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: ENCODE ATAC-seq — Human Heart

```{r}
library(ArchR)
library(here)

# Pre-processed peak matrix from ENCODE
# Heart left ventricle vs. right ventricle ATAC-seq
peaks <- read_tsv(here("data", "raw", "heart_atac_peaks.tsv"))
counts <- readRDS(here("data", "raw", "heart_peak_counts.rds"))

# Differential accessibility: LV vs. RV
library(DESeq2)

dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = data.frame(condition = c(rep("LV", 3), rep("RV", 3))),
  design    = ~ condition
)

dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "LV", "RV"))
summary(res)
```

---

## Hands-On: Motif Enrichment at Differential Peaks

```{r}
library(chromVAR)
library(BSgenome.Hsapiens.UCSC.hg38)
library(JASPAR2020)
library(TFBSTools)

# Get sequences at LV-specific peaks
lv_peaks <- peaks[which(res$padj < 0.05 & res$log2FoldChange > 1), ]
lv_gr    <- makeGRangesFromDataFrame(lv_peaks)

peak_seqs <- getSeq(BSgenome.Hsapiens.UCSC.hg38, lv_gr)

# Get JASPAR motifs
motifs <- getMatrixSet(JASPAR2020,
                       opts = list(species = "Homo sapiens",
                                   matrixtype = "PFM"))

# Match motifs
motif_matches <- matchMotifs(motifs, peak_seqs,
                              genome = BSgenome.Hsapiens.UCSC.hg38)

# Which TFs are enriched at LV-specific peaks?
enrichment <- motifEnrichment(peak_seqs, motif_matches)
enrichment_df <- motifEnrichmentTable(enrichment) |>
  arrange(p.value) |>
  head(20)
```

---

## Hands-On: Visualize a Genomic Region

```{r}
library(Gviz)

# Plot ATAC-seq signal, peaks, and gene annotation at a locus
options(ucscChromosomeNames = FALSE)

# Tracks
genome_track  <- GenomeAxisTrack()
gene_track    <- GeneRegionTrack(txdb, genome = "hg38",
                                  chromosome = "chr17",
                                  name       = "Genes")
atac_track_lv <- DataTrack(range  = lv_bigwig,
                             type   = "histogram",
                             name   = "ATAC LV",
                             col    = "#185FA5")
peak_track    <- AnnotationTrack(lv_gr,
                                  name  = "LV Peaks",
                                  col   = "#D85A30")

plotTracks(list(genome_track, gene_track,
                atac_track_lv, peak_track),
           chromosome = "chr17",
           from       = 41150000,
           to         = 41350000)
```

---

# Next Week {background-color="#534AB7"}

---

## W6 Preview — Metagenomics I: 16S rRNA-seq

**What we'll cover:**

- The human microbiome — composition and function
- 16S rRNA amplicon sequencing — principle and limitations
- DADA2 pipeline — ASV calling, taxonomy assignment
- Alpha and beta diversity metrics

**Install before next class:**

```{r}
BiocManager::install(c("dada2", "phyloseq",
                       "microbiome", "DESeq2"))
install.packages(c("vegan", "ggplot2", "patchwork"))
```

---

## Key Resources

- **ENCODE** — [encodeproject.org](https://www.encodeproject.org) — reference ATAC-seq datasets
- **Roadmap Epigenomics** — [egg2.wustl.edu/roadmap](http://egg2.wustl.edu/roadmap) — 127 reference epigenomes
- **ArchR** — [archrproject.com](https://www.archrproject.com) — comprehensive ATAC-seq in R
- **JASPAR 2022** — [jaspar.genereg.net](https://jaspar.genereg.net) — TF binding motifs
- **4D Nucleome** — [4dnucleome.org](https://www.4dnucleome.org) — Hi-C reference data

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Metagenomics I — 16S rRNA-seq**

Course materials → [your GitHub Pages URL here]
:::
