---
title: "W5. Single-cell RNA-seq — Technology, QC, Clustering"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Say what single cell answers that bulk structurally cannot — and when bulk is still the better buy
-   Choose between droplet and plate-based protocols, and ⚠️ between cells and nuclei
-   ⚠️ Explain why dissociation changes the composition you measure
-   Explain what a UMI is, and why the count matrix is mostly zeros
-   ⚠️ Read a knee plot and a QC scatter, and name the population each threshold deletes
-   ⚠️ Read a UMAP responsibly — and say what its distances do and do not mean
-   ⚠️ Explain why per-cell differential expression across conditions is invalid, and what pseudobulk does instead
:::

> 🤔 **Motivating question**
>
> Your bulk RNA-seq shows a clean two-fold increase, *p* = 0.001. Is that every cell changing a little, 5% of cells changing enormously, or no cell changing at all?

------------------------------------------------------------------------

# 1. What Bulk Cannot See 🔬 {#w5-s1}

## 1) One number, three biologies {#w5-s1-1}

![The mean is compatible with all three. Only one of them is a regulatory story.](/assets/img/w5/bulk_vs_singlecell.svg){fig-align="center"}

-   **① A general modest response** — the case everyone implicitly assumes, and the least interesting
-   **② A rare responsive subpopulation** — usually the interesting one. In bulk it is diluted by the cells doing nothing, and often falls below significance entirely
-   **③ A change in composition** — no cell changed its transcriptome. There are simply more of one cell type

⚠️ ③ deserves attention because it is common in inflamed, fibrotic or infiltrated tissue — exactly the tissue clinicians study — and it is routinely written up as if it were ①.

## 2) When single cell is worth it {#w5-s1-2}

Be specific. Single cell is expensive and frequently used where bulk would have answered better.

**✅ Single cell is right for:**

-   **Discovering or characterising populations**, including rare ones
-   **Composition** — which cell types shifted in disease
-   **Cell-type-specific effects** — a gene up in one cell type and down in another, invisible in bulk
-   **Continuous states** — activation, exhaustion, differentiation, where discrete labels fail

**✅ Bulk is still right for:**

-   A well-defined comparison in a homogeneous system (cell line, sorted population)
-   Deep transcript-level work — isoforms, splicing, low-expression genes
-   **Large cohorts** — n = 200 patients, where 200 single-cell runs is unaffordable

::: {.callout-tip collapse="true"}
## Sorting first is badly underrated

Already know which cell type you care about?

→ FACS-sort it and run deep bulk RNA-seq. Often **cheaper, deeper, and statistically cleaner** than a whole-tissue single-cell experiment — and you can afford far more patients, which (§5) is where your power actually comes from.

💡 "We did single cell" is not by itself a stronger design. Sometimes it is a more expensive way to be underpowered.
:::

------------------------------------------------------------------------

# 2. ⚠️ Decisions Made Before the Machine 🧫 {#w5-s2}

## 1) The platform trade-off {#w5-s2-1}

Every platform buys cells with depth, or depth with cells.

| Platform | Cells | Depth | When |
|--------------------|--------------|--------------|--------------------------|
| **Droplet (10x)** | 10³–10⁵ | a few thousand genes/cell, 3′ end only | the current default |
| **Plate (Smart-seq3)** | 10² | full-length, deep | isoforms, very sensitive detection |
| **Combinatorial (sci-, SPLiT-seq)** | 10⁶ | very shallow | whole-organism atlases |

## 2) ⚠️ Cells or nuclei — and why it is decided at collection {#w5-s2-2}

![The composition you measure is the composition that survived.](/assets/img/w5/dissociation_bias.svg){fig-align="center"}

| | scRNA-seq | snRNA-seq |
|-------------------------|:--------------------:|:--------------------:|
| Frozen / archived tissue | ❌ needs fresh | ✅ |
| Large or fragile cells (neurons, adipocytes, myofibres, cardiomyocytes) | ❌ lost in dissociation | ✅ |
| Dissociation stress response | ⚠️ yes — *FOS*, *JUN*, heat-shock genes induced by the protocol | ✅ minimal |
| Cytoplasmic transcripts | ✅ captured | ❌ missed |
| Intronic reads | low | high — nuclei are rich in unspliced pre-mRNA |

::: {.callout-warning collapse="true"}
## Two consequences worth carrying into a design meeting

-   ⚠️ **Immediate-early genes are the classic false positive.** A cluster defined by *FOS*, *JUN* and *EGR1* is very often a dissociation artifact, not an "activated" state. If one condition sat on the bench longer, it will look more activated
-   ⚠️ **Depletion claims need scrutiny.** "Cell type X is depleted in disease" and "cell type X is harder to release from diseased tissue" produce identical data

→ ✅ Mitigations: dissociate at low temperature with cold-active protease, keep the protocol identical across conditions, and — best — use nuclei when the tissue allows it.
:::

## 3) The design consequence, stated early {#w5-s2-3}

⚠️ **Your unit of replication is the donor.** Everything in §5 follows from this, and it should shape the budget before it shapes the analysis.

💡 Given a fixed budget: **more donors, fewer cells each** beats **fewer donors, more cells each** for almost any question that compares conditions.

------------------------------------------------------------------------

# 3. From Droplets to a Matrix 🧮 {#w5-s3}

## 1) Barcodes and UMIs {#w5-s3-1}

Two tags are attached to every captured transcript:

-   **Cell barcode** — identifies the droplet, and therefore the cell
-   **UMI** (unique molecular identifier) — a random sequence unique to each original mRNA molecule

**Why the UMI matters:** after PCR, one original molecule may be represented by hundreds of reads. Collapsing reads that share a UMI recovers the count of **molecules** rather than reads, removing amplification bias.

## 2) Why the matrix is over 90% zeros {#w5-s3-2}

A droplet experiment detects roughly 1,000–5,000 genes per cell, out of \~20,000. Zeros have two very different meanings:

-   **Biological zero** — the gene really is not expressed in that cell
-   **Technical zero (dropout)** — the transcript was there and was not captured. Droplet capture efficiency is \~10–30%

💡 The field has largely moved past calling this "zero inflation" — for UMI data the zeros are mostly what a simple sampling model predicts from low counts, not an extra process.

⚠️ The practical consequence stands: **you cannot conclude a gene is off in a cell from a single zero.**

## 3) ⚠️ QC is an analysis decision, not cleaning {#w5-s3-3}

![Left: separating cells from empty droplets. Right: the plot every paper should show.](/assets/img/w5/qc_thresholds.svg){fig-align="center"}

| Problem | What it looks like | What to do |
|--------------------|--------------------------|--------------------------|
| **Empty droplets** | Low-UMI barcodes below the knee | `EmptyDrops`, not a hard cutoff |
| **Ambient RNA** | Markers of abundant cell types appearing everywhere | `SoupX`, `CellBender`, `DecontX` |
| **Doublets** | High UMI **and** high gene count; two incompatible marker sets | `scDblFinder`, `DoubletFinder`; \~1% per 1,000 cells loaded |
| **Dying cells** | High mitochondrial fraction, low gene count | Threshold on mito %, tissue-specifically |

::: {.callout-important collapse="true"}
## Every threshold deletes a real population

-   A mito cutoff of 5% is conventional — and **wrong for cardiomyocytes**, which are genuinely mitochondria-rich. Applied to heart tissue it removes the cell type you care about
-   `min.features = 500` removes debris — **and also** platelets, some granulocytes, and any small low-RNA cell
-   An upper `nFeature` cutoff removes doublets — **and also** genuinely large, transcriptionally active cells

→ ✅ Plot the distribution, look at what sits outside your proposed cut, **then** decide. And report the thresholds — they belong in the methods, not the analyst's console history.
:::

------------------------------------------------------------------------

# 4. Finding Cell Types 🗺️ {#w5-s4}

## 1) The standard pipeline {#w5-s4-1}

```
normalise → select highly variable genes → PCA
          → build a k-nearest-neighbour graph
          → cluster (Leiden)
          → UMAP, for display only
          → annotate
```

⚠️ Note the order. **UMAP is computed for display.** Clustering happens on the PCA/graph, not on the two-dimensional picture.

## 2) ⚠️ How to read a UMAP responsibly {#w5-s4-2}

UMAP is a non-linear projection of tens of dimensions into two. It preserves local neighbourhoods reasonably and global structure badly.

::: {.callout-warning collapse="true"}
## What a UMAP does *not* tell you

-   ❌ **Distances between clusters are not meaningful.** Two clusters far apart are not necessarily more different than two that are close
-   ❌ **Cluster areas are not meaningful.** Size reflects the embedding, not cell number
-   ❌ **Apparent bridges are not trajectories.** A thin connection may be doublets, ambient RNA, or a parameter artifact
-   ⚠️ **The layout changes with parameters and random seed.** `n_neighbors` and `min_dist` visibly reshape it

→ A UMAP is a **picture** of your data, not a **measurement** of it. Any claim that matters should be supported by a number computed in the high-dimensional space.

💡 A useful journal-club question: *"which sentence in this results section is supported only by the UMAP?"*
:::

## 3) Clustering resolution is arbitrary {#w5-s4-3}

![Same cells, same algorithm, one parameter changed.](/assets/img/w5/umap_resolution.svg){fig-align="center"}

Leiden clustering takes a **resolution** parameter. Raise it, get more clusters. Lower it, get fewer.

⚠️ There is **no correct value and no statistical criterion that settles it.** The algorithm will split any cluster you ask it to.

✅ What to do instead of agonising:

-   Check clusters against **marker genes** and known biology, not against a score
-   Check **stability** — does the cluster survive a different resolution, a different seed, a subsampled dataset?
-   Ask whether a "new" cluster at higher resolution has genuinely distinct markers, or just gradations of the same ones
-   **Report the resolution you used**

::: {.callout-warning collapse="true"}
## A cluster is not a cell type

Clusters can equally reflect:

-   cell **state** (activated vs resting)
-   cell-cycle phase
-   stress response (§2.2)
-   ambient contamination
-   **donor identity**

→ Naming a cluster after its top marker gene and treating it as a cell type is how the literature accumulates cell types that do not exist.
:::

## 4) Annotation {#w5-s4-4}

-   **Marker-based** — check known markers per cluster. ✅ Transparent. ⚠️ Circular: you find the cell types you looked for
-   **Reference-based** — map onto an annotated atlas (Azimuth, SingleR, scANVI). ✅ Fast and consistent. ⚠️ Inherits the reference's vocabulary and cannot name anything the reference lacks

💡 Use both. **Disagreements between them are the informative part** — that is where a genuinely novel or genuinely artifactual population sits.

------------------------------------------------------------------------

# 5. ⚠️ Statistics 📊 {#w5-s5}

## 1) Why per-cell tests are wrong {#w5-s5-1}

![Both panels analyse identical data. Only one of them answers the question that was asked.](/assets/img/w5/pseudobulk.svg){fig-align="center"}

::: {.callout-important collapse="true"}
## Cells from one donor are not independent samples

A Wilcoxon test across 5,400 cells from 6 donors treats those cells as 5,400 independent observations. They are not — cells from the same person share that person's genetics, environment, medication, batch and processing.

The consequence:

-   p-values inflated by many orders of magnitude
-   add cells → the p-value shrinks indefinitely
-   → you are measuring **how many cells you loaded**, not how many people you studied

⚠️ Seurat's default `FindMarkers` behaves this way.

-   ✅ Appropriate for *identifying cluster markers within one dataset* — a descriptive task
-   ❌ Not appropriate for **comparing conditions**
:::

## 2) Pseudobulk {#w5-s5-2}

**The fix:** for each **donor × cell type**, sum the counts across cells. You now have a small matrix of pseudo-samples — analyse it with DESeq2 or edgeR exactly as in last semester's bulk workflow.

```{r}
library(Seurat); library(DESeq2); library(tidyverse)

pb <- AggregateExpression(seurat,
                          group.by = c("cell_type", "sample_id"),
                          assays   = "RNA")$RNA

# then, per cell type, a completely ordinary DESeq2 run
```

-   Benchmarks consistently find pseudobulk controls false positives far better than per-cell tests
-   **Mixed models** with a donor random effect are a reasonable alternative — more power retained, more computation

## 3) Differential abundance is compositional {#w5-s5-3}

Often the interesting result is not that a cell type changed its expression, but that **there is more or less of it**.

⚠️ Proportions are **compositional** — one cell type expands and every other proportion falls by arithmetic alone, with no biology behind it.

-   ✅ Methods that handle this: `scCODA`, `miloR`, `propeller`, Dirichlet-multinomial models
-   ❌ A χ² test on cell counts is **not valid**
-   ⚠️ And before believing any abundance result, return to §2.2: is this composition, or is this dissociation?

------------------------------------------------------------------------

# 6. Hands-On 💻 {#w5-s6}

## 1) No code: does the atlas already answer it? {#w5-s6-1}

[**CELLxGENE**](https://cellxgene.cziscience.com) → search a gene from your specialty.

-   ① Which cell types express it, in which tissues?
-   ② Is the expression broad, or confined to one population?
-   ③ Find a dataset in your disease area. **How many donors** does it have — not how many cells?
-   ④ 💡 If the atlas already answers "which cell type expresses my gene", you have saved a grant

## 2) The pipeline, end to end {#w5-s6-2}

```{r}
library(Seurat); library(SeuratData); library(tidyverse)

InstallData("pbmc3k"); data("pbmc3k")
pbmc <- pbmc3k

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# ⚠️ LOOK before you cut
VlnPlot(pbmc, c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)

pbmc <- subset(pbmc, nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

pbmc <- NormalizeData(pbmc) |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(verbose = FALSE)

ElbowPlot(pbmc, ndims = 30)

pbmc <- FindNeighbors(pbmc, dims = 1:10) |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:10)

DimPlot(pbmc, label = TRUE)
```

## 3) ✅ Change one thing {#w5-s6-3}

```{r}
# ① How many cells did each threshold remove, and which ones?
table(pbmc3k$seurat_annotations[pbmc3k$percent.mt >= 5])   # who did the mito cut delete?

# ② Re-cluster at three resolutions and count
for (r in c(0.2, 0.5, 1.2)) {
  pbmc <- FindClusters(pbmc, resolution = r, verbose = FALSE)
  cat("resolution", r, "→", nlevels(pbmc$seurat_clusters), "clusters\n")
}

# ③ Re-run UMAP with a different seed and compare the picture
pbmc <- RunUMAP(pbmc, dims = 1:10, seed.use = 99)
DimPlot(pbmc, label = TRUE)
```

💡 ③ is the exercise that changes how you read figures. Nothing about the biology moved.

## 4) Save for next week {#w5-s6-4}

```{r}
saveRDS(pbmc, here::here("analysis", "5_pbmc_annotated.rds"))
```

**W6 — trajectory, cell–cell communication, and multi-sample integration.**

```{r}
BiocManager::install("monocle3")
install.packages(c("harmony", "CellChat"))
```

------------------------------------------------------------------------

# 📝 Summary

-   One bulk fold change is compatible with a uniform shift, a rare responsive subpopulation, or a **pure composition change**
-   ✅ Sorting plus deep bulk on many patients often beats whole-tissue single cell on few
-   ⚠️ **Fresh or frozen is decided at collection** and fixes scRNA-seq vs snRNA-seq permanently
-   ⚠️ **Dissociation induces stress genes and destroys cells selectively.** *FOS*/*JUN* clusters are usually artifacts; depletion claims need scrutiny
-   UMIs count molecules, not reads. The matrix is \>90% zeros and a single zero means little
-   ⚠️ **Every QC threshold deletes a real population** — mito 5% removes cardiomyocytes, min.features 500 removes platelets
-   ❌ UMAP distances, cluster areas and bridges are **not interpretable**
-   ⚠️ Clustering resolution is arbitrary. Check markers and stability, and **report the value you used**
-   ⚠️ **Per-cell differential expression across conditions is invalid.** Use pseudobulk. Your n is the number of **donors**
-   ⚠️ Cell-type proportions are compositional — χ² on cell counts is not valid

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| Cell barcode | Sequence identifying which droplet a read came from |
| UMI | Tag marking one original mRNA molecule |
| Dropout | An expressed transcript that was not captured — a technical zero |
| Knee plot | Ranked UMI counts, used to separate cells from empty droplets |
| Ambient RNA | Free-floating RNA contaminating every droplet |
| Doublet | Two cells captured in one droplet |
| snRNA-seq | Profiling nuclei rather than whole cells |
| Dissociation artifact | Gene expression induced by the protocol, not the biology |
| HVG | Highly variable genes selected before PCA |
| UMAP / t-SNE | Non-linear 2D embeddings — for visualisation only |
| Leiden clustering | Graph-based community detection on the kNN graph |
| Resolution | Parameter controlling how many clusters you get |
| Pseudobulk | Summing counts per donor × cell type before testing |
| Differential abundance | Testing whether cell-type proportions changed |
| Compositional data | Proportions that must sum to 1, so they cannot move independently |

------------------------------------------------------------------------

# ➕ References

### Overviews and best practice

Heumos L, et al. [*Best practices for single-cell analysis across modalities.*](https://doi.org/10.1038/s41576-023-00586-w) Nat Rev Genet. 2023.\
[Single-cell best practices (online book)](https://www.sc-best-practices.org/) — free, current, and the one to keep open\
Luecken MD, Theis FJ. [*Current best practices in single-cell RNA-seq analysis: a tutorial.*](https://doi.org/10.15252/msb.20188746) Mol Syst Biol. 2019.

### Technology, QC, and artifacts

Lun ATL, et al. [*EmptyDrops: distinguishing cells from empty droplets in droplet-based single-cell RNA sequencing data.*](https://doi.org/10.1186/s13059-019-1662-y) Genome Biol. 2019.\
Young MD, Behjati S. [*SoupX removes ambient RNA contamination.*](https://doi.org/10.1093/gigascience/giaa151) GigaScience. 2020.\
van den Brink SC, et al. [*Single-cell sequencing reveals dissociation-induced gene expression in tissue subpopulations.*](https://doi.org/10.1038/nmeth.4437) Nat Methods. 2017. **← read this before designing a dissociation protocol**\
Denisenko E, et al. [*Systematic assessment of tissue dissociation and storage biases in single-cell and single-nucleus RNA-seq workflows.*](https://doi.org/10.1186/s13059-020-02048-6) Genome Biol. 2020.

### Embeddings and clustering

Chari T, Pachter L. [*The specious art of single-cell genomics.*](https://doi.org/10.1371/journal.pcbi.1011288) PLoS Comput Biol. 2023. **← a sharp, useful critique of 2D embeddings**\
Becht E, et al. [*Dimensionality reduction for visualizing single-cell data using UMAP.*](https://doi.org/10.1038/nbt.4314) Nat Biotechnol. 2019.

### Statistics

Squair JW, et al. [*Confronting false discoveries in single-cell differential expression.*](https://doi.org/10.1038/s41467-021-25960-2) Nat Commun. 2021.\
Zimmerman KD, Espeland MA, Langefeld CD. [*A practical solution to pseudoreplication bias in single-cell studies.*](https://doi.org/10.1038/s41467-021-21038-1) Nat Commun. 2021.\
Murphy AE, Skene NG. [*A balanced measure shows superior performance of pseudobulk methods.*](https://doi.org/10.1038/s41467-022-35519-4) Nat Commun. 2022.\
Büttner M, et al. [*scCODA is a Bayesian model for compositional single-cell data analysis.*](https://doi.org/10.1038/s41467-021-27150-6) Nat Commun. 2021.

### Atlases

[Human Cell Atlas](https://www.humancellatlas.org/) · [CELLxGENE](https://cellxgene.cziscience.com/) · [Azimuth](https://azimuth.hubmapconsortium.org/) · [Tabula Sapiens](https://tabula-sapiens.sf.czbiohub.org/)
