---
title: "Metagenomics I: 16S rRNA-seq"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 6"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W6 — 16S rRNA-seq"
    css: custom.css
execute:
  echo: true
  eval: false
---

# The Human Microbiome {background-color="#1D9E75"}

---

## You Are Not Alone

::: {.incremental}
- ~38 trillion microbial cells in/on the human body (~1:1 with human cells)
- ~3.3 million microbial genes (vs. ~20,000 human protein-coding genes)
- 99% of the microbiome is in the **gut**
- Collectively called the **gut microbiome** or **gut flora**
:::

::: {.fragment}
**Clinically relevant in:**
IBD, IBS, colorectal cancer, T2D, obesity, cardiovascular disease,  
mental health (gut–brain axis), autoimmune disease, drug metabolism,  
immunotherapy response
:::

---

## Why This Matters for Drug Response

::: {.incremental}
- **Irinotecan (CPT-11)** — gut bacteria reactivate the drug in the colon → diarrhea
- **Checkpoint inhibitors** — *Faecalibacterium prausnitzii*, *Akkermansia* associated with response
- **Metformin** — microbiome mediates glycemic effects
- **Antibiotics** → dysbiosis → *C. difficile* → CDI
:::

::: {.fragment}
::: {.callout-tip}
## The gut–immunotherapy connection
Patients with high *Akkermansia muciniphila* abundance respond significantly better to anti-PD1 therapy.  
This has been replicated in multiple cohorts across multiple tumor types.
:::
:::

---

# 16S rRNA Sequencing {background-color="#534AB7"}

---

## Why 16S rRNA?

```
16S rRNA gene:
├── Conserved regions  → universal PCR primers
└── Variable regions   → species identification
    (V1-V2, V3-V4, V4 most common)

Workflow:
Stool/swab DNA → PCR amplify V3-V4 → Sequence → Assign taxonomy
```

::: {.fragment}
**Pros:** Cheap (~$50/sample), easy, large reference databases  
**Cons:** Amplification bias, ~genus level resolution, no functional info

→ For functional metagenomics: **Whole Metagenome Shotgun (WMS)** — next week (W7)
:::

---

## ASV vs. OTU — Modern Standard

**OTU (Operational Taxonomic Unit)** — old approach:
- Cluster reads at 97% similarity → OTU table
- Information loss, not reproducible

**ASV (Amplicon Sequence Variant)** — current standard (DADA2):
- Exact sequence variants, corrected for sequencing errors
- Reproducible across studies
- Higher resolution (can distinguish strains)

::: {.fragment}
> **Always use DADA2 / ASVs for new studies.**  
> QIIME2 wraps DADA2 — both acceptable.
:::

---

## Analysis Workflow

```
Raw paired-end FASTQ (R1 + R2)
    │
    ├── 1. Quality inspection   → plotQualityProfile()
    ├── 2. Filtering & trimming → filterAndTrim()
    ├── 3. Error learning       → learnErrors()
    ├── 4. Denoising (DADA2)   → dada()
    ├── 5. Merging pairs        → mergePairs()
    ├── 6. Chimera removal      → removeBimeraDenovo()
    ├── 7. Taxonomy assignment  → assignTaxonomy() → SILVA
    └── 8. Phyloseq object      → diversity, composition analysis
```

---

# DADA2 Pipeline {background-color="#185FA5"}

---

## Step 1: Quality Profiles

```{r}
library(dada2)
library(here)

# List raw FASTQ files
fnFs <- sort(list.files(here("data", "raw", "fastq"),
                         pattern = "_R1_001.fastq.gz",
                         full.names = TRUE))
fnRs <- sort(list.files(here("data", "raw", "fastq"),
                         pattern = "_R2_001.fastq.gz",
                         full.names = TRUE))

sample_names <- sub("_R1_001.fastq.gz", "",
                    basename(fnFs))

# Visualize quality profiles
plotQualityProfile(fnFs[1:2])   # forward reads
plotQualityProfile(fnRs[1:2])   # reverse reads
# Quality usually drops at end of reads — decide truncation point
```

---

## Step 2: Filter & Trim

```{r}
# Output paths
filtFs <- file.path(here("data", "processed", "filtered"),
                    paste0(sample_names, "_F_filt.fastq.gz"))
filtRs <- file.path(here("data", "processed", "filtered"),
                    paste0(sample_names, "_R_filt.fastq.gz"))

# Filter — adjust truncLen based on quality profile
out <- filterAndTrim(
  fnFs, filtFs, fnRs, filtRs,
  truncLen    = c(240, 200),   # forward, reverse
  maxN        = 0,             # no Ns allowed
  maxEE       = c(2, 2),       # max expected errors
  truncQ      = 2,
  rm.phix     = TRUE,
  compress    = TRUE,
  multithread = TRUE
)

head(out)
# Check: >70% reads should pass filter
```

---

## Step 3–4: Learn Errors & Denoise

```{r}
# Learn error rates (takes ~10 min on 20 samples)
errF <- learnErrors(filtFs, multithread = TRUE)
errR <- learnErrors(filtRs, multithread = TRUE)

# Visualize error model — points should follow the line
plotErrors(errF, nominalQ = TRUE)

# DADA2 denoising — the core step
dadaFs <- dada(filtFs, err = errF, multithread = TRUE,
               pool = "pseudo")   # pool for rare ASVs
dadaRs <- dada(filtRs, err = errR, multithread = TRUE,
               pool = "pseudo")

dadaFs[[1]]
# dada-class: 128 sample sequences
```

---

## Step 5–7: Merge, Remove Chimeras, Assign Taxonomy

```{r}
# Merge paired reads
mergers <- mergePairs(dadaFs, filtFs,
                      dadaRs, filtRs,
                      verbose = TRUE)

# Make sequence table
seqtab <- makeSequenceTable(mergers)

# Inspect ASV length distribution — should center on ~253 bp (V3-V4)
table(nchar(getSequences(seqtab)))

# Remove chimeras
seqtab_nochim <- removeBimeraDenovo(seqtab,
                                    method     = "consensus",
                                    multithread = TRUE,
                                    verbose    = TRUE)

# Fraction of reads retained
sum(seqtab_nochim) / sum(seqtab)
```

---

## Taxonomy Assignment

```{r}
# SILVA database (download v138.1 from https://zenodo.org/record/4587955)
taxa <- assignTaxonomy(
  seqtab_nochim,
  here("data", "raw", "silva_nr99_v138.1_train_set.fa.gz"),
  multithread = TRUE
)

# Add species-level assignment
taxa <- addSpecies(
  taxa,
  here("data", "raw", "silva_species_assignment_v138.1.fa.gz")
)

# Preview
taxa_df <- as.data.frame(taxa)
head(taxa_df)
# Kingdom  Phylum          Class     Order      Family     Genus   Species
# Bacteria Bacteroidota    ...
```

---

## Track Reads Through Pipeline

```{r}
getN <- function(x) sum(getUniques(x))

track <- cbind(
  out,
  sapply(dadaFs, getN),
  sapply(dadaRs, getN),
  sapply(mergers, getN),
  rowSums(seqtab_nochim)
)

colnames(track) <- c("input", "filtered", "denoisedF",
                      "denoisedR", "merged", "nonchim")
rownames(track) <- sample_names

head(track)
# Should retain >70% from input to nonchim
# Large drops at merge → primer trimming needed
# Large drops at chimera → primer length issues
```

---

# Diversity Analysis with phyloseq {background-color="#0F6E56"}

---

## Building the phyloseq Object

```{r}
library(phyloseq)
library(tidyverse)
library(here)

# Load sample metadata
metadata <- read_tsv(here("data", "raw", "sample_metadata.tsv")) |>
  column_to_rownames("sample_id")

# Build phyloseq object
ps <- phyloseq(
  otu_table(seqtab_nochim, taxa_are_rows = FALSE),
  sample_data(metadata),
  tax_table(taxa)
)

# Rename ASVs to short names
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- paste0("ASV", seq(ntaxa(ps)))
taxa_names(ps) <- names(dna)

ps
```

---

## Alpha Diversity — Within-Sample Diversity

```{r}
# Rarefy to equal depth first
ps_rare <- rarefy_even_depth(ps,
                              sample.size = min(sample_sums(ps)),
                              rngseed     = 42)

# Alpha diversity metrics
alpha_div <- estimate_richness(ps_rare,
                                measures = c("Observed",
                                             "Shannon",
                                             "Simpson",
                                             "Chao1"))

alpha_div <- alpha_div |>
  rownames_to_column("sample_id") |>
  left_join(as(sample_data(ps), "data.frame") |>
            rownames_to_column("sample_id"))

# Compare between groups
ggplot(alpha_div, aes(x = Group, y = Shannon, fill = Group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1) +
  stat_compare_means() +
  labs(title = "Shannon diversity by group") +
  theme_minimal()
```

---

## Beta Diversity — Between-Sample Diversity

```{r}
library(vegan)

# Bray-Curtis dissimilarity — most common for microbiome
bray_dist <- distance(ps_rare, method = "bray")

# PCoA ordination
ord_pcoa <- ordinate(ps_rare, method = "PCoA",
                      distance = bray_dist)

# Plot
plot_ordination(ps_rare, ord_pcoa,
                color = "Group",
                shape = "Group") +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse() +
  labs(title = "PCoA — Bray-Curtis dissimilarity") +
  theme_minimal()

# PERMANOVA — is the group difference significant?
adonis2(bray_dist ~ Group + age + sex,
        data = as(sample_data(ps_rare), "data.frame"),
        permutations = 999)
```

---

## Differential Abundance — DESeq2 for Microbiome

```{r}
library(DESeq2)

# Note: do NOT rarefy before DESeq2 — use raw counts
ps_deseq <- phyloseq_to_deseq2(ps, ~ Group)
ps_deseq <- DESeq(ps_deseq, test = "Wald",
                   fitType = "parametric")

res <- results(ps_deseq, cooksCutoff = FALSE)
res <- res[order(res$padj), ]

# Significant ASVs
sig_asvs <- res |>
  as_tibble(rownames = "ASV") |>
  filter(!is.na(padj), padj < 0.05) |>
  left_join(as.data.frame(tax_table(ps)) |>
            rownames_to_column("ASV"))

sig_asvs |> select(ASV, Genus, Species, log2FoldChange, padj)
```

---

## Composition Visualization

```{r}
# Relative abundance bar chart — collapse to genus level
ps_genus <- tax_glom(ps_rare, taxrank = "Genus")
ps_rel   <- transform_sample_counts(ps_genus,
                                     function(x) x / sum(x))

# Top 15 genera
top_genera <- names(sort(taxa_sums(ps_rel),
                          decreasing = TRUE))[1:15]
ps_top <- prune_taxa(top_genera, ps_rel)

plot_bar(ps_top, fill = "Genus") +
  facet_wrap(~ Group, scales = "free_x") +
  labs(title = "Relative abundance — top 15 genera") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: IBD Microbiome (HMP2)

Human Microbiome Project 2 — Inflammatory Bowel Disease cohort  
(subset: 30 CD, 30 UC, 30 healthy controls; pre-processed ASV table)

```{r}
library(phyloseq)
library(tidyverse)
library(here)

# Load pre-processed phyloseq object
ps_ibd <- readRDS(here("data", "raw", "hmp2_phyloseq.rds"))
ps_ibd

# Sample breakdown
table(sample_data(ps_ibd)$diagnosis)
# CD: 30, UC: 30, nonIBD: 30
```

---

## Hands-On: Alpha & Beta Diversity

```{r}
# Alpha diversity
ps_ibd_rare <- rarefy_even_depth(ps_ibd, rngseed = 42)

alpha <- estimate_richness(ps_ibd_rare,
                            measures = c("Shannon", "Observed")) |>
  rownames_to_column("sample") |>
  left_join(as(sample_data(ps_ibd), "data.frame") |>
            rownames_to_column("sample"))

ggplot(alpha, aes(diagnosis, Shannon, fill = diagnosis)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.5) +
  scale_fill_manual(values = c("#1D9E75", "#D85A30", "#378ADD")) +
  labs(title = "Alpha diversity — IBD vs. healthy") +
  theme_minimal()

# Beta diversity — PERMANOVA
bray <- distance(ps_ibd_rare, method = "bray")
adonis2(bray ~ diagnosis, data = as(sample_data(ps_ibd_rare),
                                    "data.frame"))
```

---

## Hands-On: Find Dysbiotic Taxa

```{r}
# Differential abundance: CD vs. nonIBD
ps_cd <- subset_samples(ps_ibd, diagnosis %in% c("CD", "nonIBD"))
dds   <- phyloseq_to_deseq2(ps_cd, ~ diagnosis)
dds   <- DESeq(dds)
res   <- results(dds, contrast = c("diagnosis", "CD", "nonIBD"),
                  cooksCutoff = FALSE)

# Annotate with taxonomy
res_df <- res |>
  as_tibble(rownames = "ASV") |>
  filter(!is.na(padj)) |>
  left_join(as.data.frame(tax_table(ps_ibd)) |>
            rownames_to_column("ASV")) |>
  arrange(padj)

# Top depleted in CD (dysbiosis signature)
res_df |>
  filter(padj < 0.05, log2FoldChange < -1) |>
  select(Genus, Species, log2FoldChange, padj)
```

---

# Next Week {background-color="#534AB7"}

---

## W7 Preview — Metagenomics II: WMS

**What we'll cover:**

- Shotgun metagenomics vs. 16S — when to use which
- HUMAnN3 functional profiling
- Taxonomic profiling with MetaPhlAn4
- Resistome analysis — AMR gene detection

**Install before next class:**

```{r}
install.packages(c("MicrobiomeStat", "ggpicrust2"))
BiocManager::install("curatedMetagenomicData")
# HUMAnN3 / MetaPhlAn4: command-line tools
# Pre-computed output will be provided
```

---

## Key Resources

- **SILVA** — [arb-silva.de](https://www.arb-silva.de) — 16S reference database
- **DADA2 tutorial** — [benjjneb.github.io/dada2](https://benjjneb.github.io/dada2)
- **phyloseq** — [joey711.github.io/phyloseq](https://joey711.github.io/phyloseq)
- **MicrobiomeAnalyst** — [microbiomeanalyst.ca](https://www.microbiomeanalyst.ca) — web-based analysis
- **curatedMetagenomicData** — curated public microbiome datasets in R

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Metagenomics II — Whole Metagenome Shotgun**

Course materials → [your GitHub Pages URL here]
:::
