---
title: "W6. Integration, Trajectory, Regulation & Communication"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Say what integration assumes, and ⚠️ recognise when it has erased the biology you were looking for
-   ✅ Demand the two colourings that make over-correction visible
-   Explain what pseudotime orders — and ⚠️ why it is not time
-   ⚠️ Say why a trajectory algorithm always returns a trajectory
-   Explain where RNA velocity's direction comes from, and which of its assumptions are contested
-   ⚠️ Separate what a regulon built from TF–target co-expression does and does not tell you
-   ⚠️ Explain why ligand–receptor co-expression is not evidence of an interaction
-   Decide which of these results belongs in a grant as a hypothesis, and which belongs in a paper as a finding
:::

> 🤔 **Motivating question**
>
> A paper shows a beautiful UMAP where patient and control cells overlap perfectly, then reports that a disease-specific fibroblast state exists. Are those two claims compatible?

------------------------------------------------------------------------

# 1. Integration and What It Costs 🔗 {#w6-s1}

## 1) The problem integration solves {#w6-s1-1}

In single-cell data the **donor** is usually the dominant source of variation. Cells cluster by patient before they cluster by cell type — which makes the data uninterpretable as biology.

**Integration methods** (Harmony, scVI, Seurat RPCA, scANVI) find a shared representation in which cells of the same type from different donors sit together.

## 2) ⚠️ Integration is an assumption in disguise {#w6-s1-2}

![Both right-hand panels look perfectly mixed. Only one colouring reveals what happened.](/assets/img/w6/integration_overcorrection.svg){fig-align="center"}

The assumption: **the same cell types exist in every batch, and any batch-associated variation is technical.**

⚠️ That assumption fails exactly when your biology is interesting:

-   a disease-specific cell state
-   a population present only in one condition
-   a treatment-induced program

→ Aggressive integration merges it away, and the resulting UMAP looks beautifully mixed.

::: {.callout-important collapse="true"}
## The diagnostic, and why it is rarely shown

-   ✅ Colour by batch **and** by cell type. Publish both
-   ✅ Confirm that known cell-type markers still separate after integration
-   ✅ Confirm that a population unique to one condition **survived**
-   ✅ Benchmarks such as **scIB** score exactly this batch-removal / biology-preservation trade-off

⚠️ Most papers show only the batch colouring — because that is the one that demonstrates the method worked. It is also the one that cannot reveal the failure.

💡 Back to the motivating question: a perfectly mixed UMAP and a disease-specific state are **in tension**. Not impossible — but the paper owes you the second colouring.
:::

## 3) ⚠️ When integration cannot help at all {#w6-s1-3}

If disease status is perfectly confounded with batch — all patients processed on one day, all controls on another — **no integration method can separate them.** Exactly the argument from 🔗 [W1 §3.3](../part1/w1_intro.qmd#w1-s3-3), and just as unrecoverable here.

------------------------------------------------------------------------

# 2. Trajectory and Pseudotime 🧭 {#w6-s2}

## 1) The idea {#w6-s2-1}

Biology is dynamic; a single-cell experiment is a snapshot. If cells were caught mid-process, they can be **ordered** along an axis of transcriptional similarity — a **pseudotime**.

-   Stem cell → progenitor → differentiated
-   Naive T cell → activated → effector → exhausted
-   Fibroblast → myofibroblast → senescent

Tools: **Monocle3**, **Slingshot**, **PAGA**.

## 2) ⚠️ The algorithm always succeeds {#w6-s2-2}

![Same tool, same settings, same confident output.](/assets/img/w6/trajectory_caveats.svg){fig-align="center"}

::: {.callout-warning collapse="true"}
## Three caveats worth stating plainly

-   ⚠️ **A trajectory algorithm returns a trajectory for any dataset**, including one with no underlying process. There is no "no trajectory found" output
-   ⚠️ **Pseudotime is not time.** It measures neither rate nor duration. Two cells far apart in pseudotime may be minutes or months apart in reality — or the ordering may be reversed
-   ⚠️ **The root has to be chosen by you.** Reverse the root and the biology reverses with it. That choice is prior knowledge entering as if it were a result

→ ✅ Sanity check: shuffle your cell labels, or run the algorithm on a homogeneous population, and see what comes back. It is a sobering five minutes.
:::

## 3) RNA velocity {#w6-s2-3}

![Where the arrows come from.](/assets/img/w6/rna_velocity.svg){fig-align="center"}

RNA velocity adds **direction** by comparing unspliced (new) to spliced (mature) transcripts per gene. More unspliced than steady state predicts → the gene is turning on.

⚠️ The assumptions about splicing and degradation kinetics are **contested**, and known to fail for some genes and systems. The arrows are also drawn on a UMAP, so they inherit every distortion from 🔗 [W5 §4.3](w5_scrna_qc.qmd#w5-s4-3).

✅ What settles a trajectory claim: a **time course**, or **lineage tracing**. Not a better algorithm.

------------------------------------------------------------------------

# 3. Inferring Mechanism from Co-expression 📡 {#w6-s3}

Same raw material — **co-expression** within one snapshot. Only where you look differs:

-   **Within a cell** — TF ↔ target genes → a **regulon** (gene regulatory network)
-   **Between cells** — ligand ↔ receptor → **cell–cell communication**

→ Same limit, too: co-expression is not regulation, and not interaction. Both are hypothesis generators.

## 1) Within a cell: gene regulatory networks (regulons) {#w6-s3-1}

> Which TF is switching this cell state on?

![](/assets/img/w6/scenic_regulon.svg){fig-align="center"}

**SCENIC — three steps**

1.  **Co-expression (GRNBoost2)** — predict each gene's expression from TF expression (tree-based regression) → candidate TF–target pairs. Mostly indirect correlation
2.  **Motif pruning (cisTarget)** — keep only targets whose promoter region is enriched for that TF's motif → a **regulon** (TF + motif-backed targets). This step is what lifts SCENIC above plain co-expression
3.  **Activity per cell (AUCell)** — how strongly the regulon's genes sit at the top of each cell's expression ranking (AUC) → a cell × regulon matrix → cluster, colour the UMAP, compare groups

Tools: **pySCENIC** (Python, the one used in practice), SCENIC (R, the original, slow)

::: {.callout-warning collapse="true"}
## Reading regulons

-   ⚠️ **TF mRNA ≠ TF activity.** Phosphorylation, nuclear translocation (NF-κB) and ligand binding (nuclear receptors) are invisible in RNA — and TFs are lowly expressed, so dropout is common
-   ⚠️ **A motif ≠ binding.** Closed chromatin blocks it, and scRNA alone cannot tell (with multiome / scATAC: SCENIC+, beyond this course)
-   ⚠️ **Different every run.** GRNBoost2 is stochastic → run it several times and keep the regulons that recur
-   ⚠️ **Circularity.** Comparing regulon activity between clusters that were defined by expression → differences are guaranteed
-   ⚠️ **Comparing conditions = the W5 per-cell trap again.** Aggregate per donor before testing (🔗 [W5 §5.2](w5_scrna_qc.qmd#w5-s5-2))

💡 Lighter alternative: **decoupleR + CollecTRI** — no network inference; score TF activity from known TF–target sets. Fast, reproducible, runs directly on a Seurat object (§5.4). The price: answers only within known biology

✅ Validation: binding by ChIP-seq / CUT&RUN; function by TF knockout / CRISPR / Perturb-seq
:::

## 2) Between cells: cell–cell communication {#w6-s3-2}

CellPhoneDB, CellChat and NicheNet score ligand–receptor pairs between cluster pairs based on **co-expression** — the ligand's mRNA is high in cluster A, its receptor's mRNA is high in cluster B.

![The arrow in the figure is doing far more work than the data supports.](/assets/img/w6/ccc_fallacy.svg){fig-align="center"}

💡 NicheNet is different in kind — it links the two axes: it asks whether a ligand's (between cells) *downstream target genes* are induced inside the receiver cell (within a cell). Still correlative, but a stronger correlation than co-expression alone.

## 3) Using them honestly (both) {#w6-s3-3}

-   ✅ **As a hypothesis generator** — "which TF / which axis is worth testing?" is exactly what these tools are for
-   ✅ **Validated spatially (CCC)** — if the two cell types are physically adjacent (🔗 [W7](../part4/w7_spatial_basics.qmd)), the prediction becomes plausible
-   ✅ **Validated functionally** — CCC: blocking antibody, receptor knockout, ligand stimulation / regulons: TF knockout, Perturb-seq
-   ❌ **As a standalone result** — neither "macrophages signal to T cells via TNF" nor "TF X drives the fibrosis program" is shown by co-expression

------------------------------------------------------------------------

# 4. Putting the Week Together 🧩 {#w6-s4}

Everything in this chapter produces a confident-looking output from data that cannot fully support it. That is not a reason to avoid these methods — it is a reason to be precise about what they are.

| Result | Strength of evidence | Where it belongs |
|----|----|----|
| Cluster + validated markers | ✅ Descriptive, solid | Results |
| Pseudobulk DE across donors (W5) | ✅ Inferential, solid | Results |
| Differential abundance with a compositional method | ⚠️ Solid if n is donors | Results, with caveats |
| Pseudotime ordering | ⚠️ Hypothesis | Results, phrased as "consistent with" |
| RNA velocity direction | ⚠️ Hypothesis, contested method | Supplementary, or a hypothesis |
| Regulon / TF activity (SCENIC, decoupleR) | ⚠️ Hypothesis | Results, phrased as "consistent with" |
| Ligand–receptor prediction | ❌ Hypothesis only | Discussion, or a grant aim |

::: {.callout-tip collapse="true"}
## The sentence that gets a paper into trouble

*"Trajectory analysis revealed that fibroblasts differentiate into myofibroblasts, and cell–cell communication analysis showed that macrophage-derived TGF-β drives this transition."*

Nothing here was revealed or shown. Two algorithms produced two orderings from one snapshot of dissociated cells.

✅ The honest version: *"Pseudotime ordering is consistent with a fibroblast-to-myofibroblast transition, and ligand–receptor analysis nominates macrophage TGF-β as a candidate driver. We are testing this by \[spatial validation / receptor blockade\]."*

→ The second version is also the one that reads like a fundable next step.
:::

------------------------------------------------------------------------

# 5. Hands-On 💻 {#w6-s5}

## 1) Integration, with the honest diagnostic {#w6-s5-1}

```{r}
library(Seurat); library(harmony); library(patchwork)

obj <- readRDS(here::here("analysis", "5_pbmc_annotated.rds"))

obj <- obj |>
  RunHarmony(group.by.vars = "donor") |>
  FindNeighbors(reduction = "harmony", dims = 1:20) |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(reduction = "harmony", dims = 1:20)

# ✅ the two panels that must always appear together
DimPlot(obj, group.by = "donor")     |
DimPlot(obj, group.by = "cell_type", label = TRUE)
```

✅ **Change one thing:** raise Harmony's `theta` (more aggressive correction) and watch whether any small population disappears from the right-hand panel.

## 2) A trajectory, and its sanity check {#w6-s5-2}

```{r}
library(monocle3)

cds <- as.cell_data_set(obj)
cds <- cluster_cells(cds) |> learn_graph()
cds <- order_cells(cds)              # ⚠️ you choose the root — that is an assumption
plot_cells(cds, color_cells_by = "pseudotime")
```

⚠️ **Then do this**, and look at the output honestly:

```{r}
# Run the same pipeline on one homogeneous cluster with no expected process
mono <- subset(obj, cell_type == "CD14 Mono")
cds_null <- as.cell_data_set(mono) |> cluster_cells() |> learn_graph()
plot_cells(cds_null, color_cells_by = "pseudotime")
```

💡 You will get a trajectory. Discuss what that means for the previous figure.

## 3) Cell–cell communication {#w6-s5-3}

```{r}
library(CellChat)

cc <- createCellChat(obj, group.by = "cell_type") |>
  subsetData() |> identifyOverExpressedGenes() |>
  identifyOverExpressedInteractions() |> computeCommunProb() |>
  filterCommunication(min.cells = 10) |> aggregateNet()

netVisual_circle(cc@net$weight, weight.scale = TRUE)
```

✅ Pick the strongest predicted interaction and write **one sentence** describing the experiment that would test it. That sentence is the deliverable, not the plot.

## 4) TF activity (decoupleR) {#w6-s5-4}

```{r}
library(decoupleR)

net  <- get_collectri(organism = "human", split_complexes = FALSE)   # known TF–target pairs
mat  <- as.matrix(LayerData(obj, assay = "RNA", layer = "data"))      # log-normalised
acts <- run_ulm(mat = mat, net = net, .source = "source", .target = "target",
                .mor = "mor", minsize = 5)

obj[["tf"]] <- acts |>
  tidyr::pivot_wider(id_cols = "source", names_from = "condition", values_from = "score") |>
  tibble::column_to_rownames("source") |>
  CreateAssayObject()

# TF mRNA vs TF activity, side by side
DefaultAssay(obj) <- "RNA"; p1 <- FeaturePlot(obj, "SPI1")
DefaultAssay(obj) <- "tf";  p2 <- FeaturePlot(obj, "SPI1")
p1 | p2
```

✅ **Change one thing:** swap SPI1 (PU.1) for a TF you care about and compare its mRNA with its activity — how different are they? To compare conditions, average per donor, not per cell.

## 5) Before next week {#w6-s5-5}

**W7 — spatial transcriptomics.** Come with an answer to one question about your own department:

> *Do we have FFPE blocks with linked outcome data, and roughly how many?*

Everything in W7–W8 depends on that answer.

------------------------------------------------------------------------

# 📝 Summary

-   Integration assumes the same cell types exist in every batch and that batch variation is technical
-   ⚠️ **Over-correction erases exactly the disease-specific population you were looking for** — and the resulting UMAP looks perfect
-   ✅ Always colour by batch **and** cell type. Demand both panels when reviewing
-   ⚠️ Confounded design (all cases one day, all controls another) cannot be rescued by any integration method
-   ⚠️ **A trajectory algorithm returns a trajectory for random data.** There is no null output
-   ⚠️ **Pseudotime is not time** — no rate, no duration — and the root is your assumption, not a result
-   RNA velocity's kinetic assumptions are contested, and its arrows sit on a distorted 2D embedding
-   Regulons (within a cell) and cell–cell communication (between cells) use the same raw material: snapshot co-expression
-   ⚠️ **TF–target co-expression is not regulation.** SCENIC filters by motif once, but TF mRNA ≠ activity and motif ≠ binding
-   ⚠️ **Ligand–receptor co-expression is not interaction**: cells may never be adjacent, the protein may never be made, and the database is biased toward what is already known
-   ✅ These are hypothesis generators. Phrase them as "consistent with" and name the experiment that would test them

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|----|----|
| Integration | Aligning datasets so shared cell types co-cluster |
| Over-correction | Integration that removes real biological differences |
| scIB | Benchmark scoring batch removal against biology preservation |
| Pseudotime | Ordering of cells along an inferred process |
| Root cell | The chosen starting point — reverses the whole ordering |
| Trajectory inference | Fitting a path through cells in expression space |
| RNA velocity | Direction inferred from unspliced vs spliced reads |
| Spliced / unspliced | Mature mRNA vs intron-containing nascent transcript |
| Regulon | A TF plus the target genes backed by its motif |
| SCENIC | Three-step GRN inference: co-expression → motif pruning → per-cell regulon activity |
| TF activity | A TF's activity inferred from its targets' expression — not the TF's own mRNA level |
| Ligand–receptor pair | A curated interaction used to score communication |
| CellChat / CellPhoneDB / NicheNet | Communication inference tools, all correlative |
| Hypothesis generator | A method whose output is a question, not an answer |

------------------------------------------------------------------------

# ➕ References

### Integration

Korsunsky I, et al. [*Fast, sensitive and accurate integration of single-cell data with Harmony.*](https://doi.org/10.1038/s41592-019-0619-0) Nat Methods. 2019.\
Luecken MD, et al. [*Benchmarking atlas-level data integration in single-cell genomics.*](https://doi.org/10.1038/s41592-021-01336-8) Nat Methods. 2022. (scIB)\
Lopez R, et al. [*Deep generative modeling for single-cell transcriptomics.*](https://doi.org/10.1038/s41592-018-0229-2) Nat Methods. 2018. (scVI)

### Trajectory and velocity

Saelens W, et al. [*A comparison of single-cell trajectory inference methods.*](https://doi.org/10.1038/s41587-019-0071-9) Nat Biotechnol. 2019. **← the benchmark that shows how much method choice matters**\
Cao J, et al. [*The single-cell transcriptional landscape of mammalian organogenesis.*](https://doi.org/10.1038/s41586-019-0969-x) Nature. 2019. (Monocle3)\
Bergen V, et al. [*Generalizing RNA velocity to transient cell states through dynamical modeling.*](https://doi.org/10.1038/s41587-020-0591-3) Nat Biotechnol. 2020. (scVelo)\
Bergen V, Soldatov RA, Kharchenko PV, Theis FJ. [*RNA velocity — current challenges and future perspectives.*](https://doi.org/10.15252/msb.202110282) Mol Syst Biol. 2021. **← read this before using velocity in a figure**

### Gene regulatory networks

Aibar S, et al. [*SCENIC: single-cell regulatory network inference and clustering.*](https://doi.org/10.1038/nmeth.4463) Nat Methods. 2017.\
Van de Sande B, et al. [*A scalable SCENIC workflow for single-cell gene regulatory network analysis.*](https://doi.org/10.1038/s41596-020-0336-2) Nat Protoc. 2020. **← the practical pySCENIC protocol, including repeated runs**\
Badia-i-Mompel P, et al. [*decoupleR: ensemble of computational methods to infer biological activities from omics data.*](https://doi.org/10.1093/bioadv/vbac016) Bioinform Adv. 2022.\
Müller-Dott S, et al. [*Expanding the coverage of regulons from high-confidence prior knowledge for accurate estimation of transcription factor activities.*](https://doi.org/10.1093/nar/gkad841) Nucleic Acids Res. 2023. (CollecTRI)

### Cell–cell communication

Jin S, et al. [*Inference and analysis of cell-cell communication using CellChat.*](https://doi.org/10.1038/s41467-021-21246-w) Nat Commun. 2021.\
Browaeys R, Saelens W, Saeys Y. [*NicheNet: modeling intercellular communication by linking ligands to target genes.*](https://doi.org/10.1038/s41592-019-0667-5) Nat Methods. 2020.\
Dimitrov D, et al. [*Comparison of methods and resources for cell-cell communication inference from single-cell RNA-Seq data.*](https://doi.org/10.1038/s41467-022-30755-0) Nat Commun. 2022. **← how much the answer depends on which tool and database you picked**
