---
title: "Metagenomics II: Whole Metagenome Shotgun (WMS)"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 7"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W7 — Whole Metagenome Shotgun"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Beyond 16S {background-color="#1D9E75"}

---

## 16S vs. WMS — When to Use Which

| | 16S rRNA-seq | WMS |
|--|-------------|-----|
| **What you sequence** | Single marker gene | All DNA in sample |
| **Resolution** | Genus level | Species / strain |
| **Functional info** | ❌ | ✅ |
| **Cost** | $ | $$$ |
| **Throughput** | Very high | Medium |
| **Host contamination** | Not an issue | Must remove |
| **Best for** | Discovery, large cohorts | Mechanism, function |

::: {.fragment}
::: {.callout-tip}
## Clinical rule of thumb
**16S** for screening / large-scale studies.  
**WMS** when you need to know **what the microbiome is doing**, not just who's there.
:::
:::

---

## What WMS Gives You

::: {.incremental}
- **Taxonomic profiling** — species and strain level (MetaPhlAn4)
- **Functional profiling** — which metabolic pathways are active (HUMAnN3)
- **Resistome** — antimicrobial resistance genes (CARD, AMRFinderPlus)
- **Virulence factors** — VFDB database
- **Phageome** — bacteriophage composition
- **De novo assembly** — MAGs (Metagenome-Assembled Genomes)
:::

---

# Taxonomic Profiling with MetaPhlAn4 {background-color="#534AB7"}

---

## How MetaPhlAn4 Works

```
WMS reads
    ↓
Map against clade-specific marker genes
(~5.1 million markers from ~99,000 genomes)
    ↓
Estimate relative abundance at each taxonomic level:
Kingdom → Phylum → Class → Order → Family → Genus → Species → Strain
```

::: {.fragment}
**Advantages over 16S:**
- No PCR amplification bias
- Species and strain level resolution
- Includes viruses, fungi, archaea
:::

---

## Working with MetaPhlAn Output in R

```{r}
library(tidyverse)
library(here)

# MetaPhlAn4 output (pre-computed)
# Tab-separated: clade_name | clade_taxid | relative_abundance
metaphlan <- read_tsv(here("data", "processed", "metaphlan_merged.tsv"),
                       skip = 1)

# Filter to species level
species <- metaphlan |>
  filter(str_detect(`#clade_name`, "s__"),
         !str_detect(`#clade_name`, "t__")) |>
  mutate(species = str_extract(`#clade_name`,
                                "s__[^|]+$") |>
           str_remove("s__") |>
           str_replace("_", " "))

# Wide to long
species_long <- species |>
  pivot_longer(-c(`#clade_name`, NCBI_tax_id, species),
               names_to  = "sample",
               values_to = "relative_abundance")
```

---

## Heatmap of Top Species

```{r}
library(ComplexHeatmap)

# Top 30 most variable species
top_species <- species |>
  rowwise() |>
  mutate(variance = var(c_across(where(is.numeric)))) |>
  ungroup() |>
  slice_max(variance, n = 30)

mat <- top_species |>
  select(-c(`#clade_name`, NCBI_tax_id, variance)) |>
  column_to_rownames("species") |>
  as.matrix()

# Heatmap with annotation
ha <- HeatmapAnnotation(
  Group = sample_data$Group,
  col   = list(Group = c(IBD = "#D85A30", Control = "#1D9E75"))
)

Heatmap(
  log10(mat + 1e-4),
  name            = "log10(rel. abund.)",
  top_annotation  = ha,
  show_column_names = FALSE,
  clustering_distance_rows = "pearson"
)
```

---

# Functional Profiling with HUMAnN3 {background-color="#185FA5"}

---

## What HUMAnN3 Does

```
WMS reads
    ↓
MetaPhlAn4 taxonomic profile (which bugs are present)
    ↓
Map reads against pangenomes of identified species
    ↓
Unmapped reads → translated search against UniRef90
    ↓
Output:
├── Gene family abundances (UniRef90 RPK)
├── Pathway abundances (MetaCyc)
└── Pathway coverages
```

---

## Working with HUMAnN3 Output

```{r}
library(tidyverse)
library(here)

# Load pathway abundance table (pre-computed)
pathways <- read_tsv(here("data", "processed",
                           "humann3_pathabundance_relab.tsv"))

# Filter to community-level (remove per-species stratification)
pathways_comm <- pathways |>
  filter(!str_detect(`# Pathway`, "\\|")) |>
  filter(`# Pathway` != "UNMAPPED",
         `# Pathway` != "UNINTEGRATED")

cat("Pathways detected:", nrow(pathways_comm), "\n")

# Long format
pathways_long <- pathways_comm |>
  pivot_longer(-`# Pathway`,
               names_to  = "sample",
               values_to = "relative_abundance") |>
  rename(pathway = `# Pathway`)
```

---

## Differential Pathway Analysis — MaAsLin2

```{r}
library(Maaslin2)

# MaAsLin2: linear model for microbial multi-omics
# Handles compositional data correctly

fit_data <- Maaslin2(
  input_data     = pathways_comm |>
    column_to_rownames("# Pathway") |>
    t() |> as.data.frame(),
  input_metadata = sample_metadata,
  output         = here("results", "maaslin2_pathways"),
  fixed_effects  = c("diagnosis", "age", "bmi"),
  random_effects = c("subject"),     # if repeated measures
  min_abundance  = 0.0001,
  min_prevalence = 0.1,
  normalization  = "NONE",           # already normalized
  transform      = "LOG"
)

# Significant pathways
fit_data$results |>
  filter(qval < 0.05) |>
  arrange(qval)
```

---

## Interpreting Pathway Results

```{r}
library(ggplot2)

# Top differentially abundant pathways
top_paths <- fit_data$results |>
  filter(qval < 0.05, metadata == "diagnosis") |>
  slice_max(abs(coef), n = 20)

ggplot(top_paths, aes(x = coef,
                       y = reorder(feature, coef),
                       fill = coef > 0)) +
  geom_col() +
  scale_fill_manual(values = c("#1D9E75", "#D85A30"),
                    labels  = c("Depleted in IBD",
                                "Enriched in IBD")) +
  labs(x    = "Coefficient (IBD vs. Control)",
       y    = "MetaCyc Pathway",
       fill = "",
       title = "Differentially abundant pathways — IBD") +
  theme_minimal()
```

---

# Resistome Analysis {background-color="#0F6E56"}

---

## Antimicrobial Resistance Genes in the Gut

::: {.incremental}
- The gut microbiome is a **reservoir of AMR genes** (the resistome)
- Antibiotic exposure → selection → enrichment of resistant strains
- Horizontal gene transfer → spread between commensal and pathogenic bacteria
- **Clinical relevance:** pre-treatment resistome may predict ICU infection outcomes,
  C. diff. colonization risk, and post-transplant infection
:::

---

## Resistome Profiling with CARD

```{r}
library(tidyverse)
library(here)

# AMR gene abundance table from AMRFinderPlus / RGI
# (aligned against CARD database — pre-computed)
amr <- read_tsv(here("data", "processed", "amr_gene_counts.tsv"))

# Summarize by drug class
amr_class <- amr |>
  group_by(sample, drug_class) |>
  summarise(total_rpk = sum(reads_per_kbp), .groups = "drop")

# Diversity of resistance genes
amr_div <- amr |>
  group_by(sample) |>
  summarise(
    n_genes  = n_distinct(gene_family),
    total    = sum(reads_per_kbp)
  ) |>
  left_join(sample_metadata)

# More antibiotic use → more AMR genes?
cor.test(amr_div$antibiotic_days, amr_div$n_genes,
          method = "spearman")
```

---

## Curatedmetagenomicdata — Pre-Analyzed Public Datasets

```{r}
library(curatedMetagenomicData)
library(SummarizedExperiment)

# Access 10,000+ published WMS samples
# Pre-computed MetaPhlAn4 profiles
available_studies <- sampleMetadata |>
  count(study_name, disease) |>
  arrange(desc(n))

# Download CRC microbiome meta-analysis dataset
crc <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE
)

crc_se <- crc[[1]]   # SummarizedExperiment
colData(crc_se) |> head()
```

---

## Multi-Study Meta-Analysis

```{r}
# Combine multiple CRC cohorts
studies <- c("ZellerG_2014", "YuJ_2015",
             "FengQ_2015", "VogtmannE_2016")

all_crc <- curatedMetagenomicData(
  paste0(studies, ".relative_abundance"),
  dryrun = FALSE
)

# Merge into one SummarizedExperiment
merged <- mergeData(all_crc)

# Batch correction by study
library(sva)
combat_corrected <- ComBat(
  dat     = assay(merged),
  batch   = colData(merged)$study_name,
  mod     = model.matrix(~ study_condition, data = colData(merged))
)
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: CRC Microbiome (curatedMetagenomicData)

```{r}
library(curatedMetagenomicData)
library(tidyverse)
library(phyloseq)
library(here)

# Zeller 2014 — CRC vs. healthy controls
crc_se <- curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE
)[[1]]

# Convert to phyloseq
ps_crc <- makePhyloseqFromSE(crc_se)
ps_crc

table(sample_data(ps_crc)$study_condition)
# CRC: 53, control: 61, adenoma: 42
```

---

## Hands-On: Species-Level Composition

```{r}
# Filter to species level
ps_sp <- subset_taxa(ps_crc,
                      !is.na(Species) & Species != "")

# Relative abundance
ps_rel <- transform_sample_counts(ps_sp,
                                   function(x) x / sum(x))

# Top taxa enriched in CRC
ps_crc_only <- subset_samples(ps_crc,
                               study_condition %in% c("CRC", "control"))

dds_crc <- phyloseq_to_deseq2(ps_crc_only, ~ study_condition)
dds_crc <- DESeq(dds_crc)
res_crc  <- results(dds_crc, cooksCutoff = FALSE)

res_df <- res_crc |>
  as_tibble(rownames = "taxon") |>
  filter(!is.na(padj)) |>
  arrange(padj)

# Fusobacterium nucleatum — known CRC marker?
res_df |> filter(str_detect(taxon, "Fusobacterium"))
```

---

## Hands-On: Pathway Analysis

```{r}
# Download pathway data for same cohort
crc_path <- curatedMetagenomicData(
  "ZellerG_2014.pathway_abundance",
  dryrun = FALSE
)[[1]]

# MaAsLin2 on pathways
library(Maaslin2)

path_mat <- assay(crc_path) |> t() |> as.data.frame()
meta      <- colData(crc_path) |> as.data.frame()

fit <- Maaslin2(
  input_data     = path_mat,
  input_metadata = meta,
  output         = here("results", "maaslin2_crc"),
  fixed_effects  = c("study_condition", "age", "gender"),
  min_prevalence = 0.1,
  normalization  = "NONE",
  transform      = "LOG"
)

fit$results |>
  filter(qval < 0.05, metadata == "study_condition") |>
  arrange(qval) |>
  head(20)
```

---

# Next Week {background-color="#534AB7"}

---

## W8 Preview — scRNA-seq Fundamentals

**What we'll cover:**

- From bulk to single-cell — why single-cell?
- 10x Genomics Chromium workflow
- Seurat pipeline: QC → normalization → clustering → annotation
- Cell type annotation with SingleR

**Install before next class:**

```{r}
install.packages("Seurat")
BiocManager::install(c("SingleR", "celldex",
                        "scran", "scater"))
install.packages("SeuratData")   # practice datasets
```

---

## Key Resources

- **curatedMetagenomicData** — [bioconductor.org/packages/curatedMetagenomicData](https://bioconductor.org/packages/curatedMetagenomicData)
- **HUMAnN3** — [huttenhower.sph.harvard.edu/humann](http://huttenhower.sph.harvard.edu/humann)
- **MetaPhlAn4** — [github.com/biobakery/MetaPhlAn](https://github.com/biobakery/MetaPhlAn)
- **CARD** — [card.mcmaster.ca](https://card.mcmaster.ca) — AMR gene database
- **MaAsLin2** — [github.com/biobakery/Maaslin2](https://github.com/biobakery/Maaslin2)

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Single-cell RNA-seq Fundamentals**

Course materials → [your GitHub Pages URL here]
:::
