---
title: "Advanced Single-cell Analysis: Trajectory & Cell–Cell Communication"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 9"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W9 — Trajectory & CCC"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Beyond Clustering {background-color="#1D9E75"}

---

## What Clustering Doesn't Tell You

From last week — we can cluster cells and annotate them.

But biology is **dynamic**, not static:

::: {.incremental}
- Stem cells → progenitors → differentiated cells
- Naive T cells → activated → effector → exhausted
- Fibroblasts → myofibroblasts → senescent cells
:::

::: {.fragment}
> **Pseudotime trajectory analysis** asks:  
> Given a snapshot of cells, can we infer the **order** and **direction** of differentiation?
:::

---

## Three Questions This Week

1. **Where are cells going?** → Trajectory / pseudotime (Monocle3)
2. **How fast and in which direction?** → RNA velocity (scVelo)
3. **Who is talking to whom?** → Cell–cell communication (CellChat)

Plus: **Integration** of multiple samples with Harmony

---

# Pseudotime Trajectory {background-color="#534AB7"}

---

## The Core Idea

```
scRNA-seq is a snapshot:
All cells captured at the same moment

But cells are at different stages of a process:
Cell A: early progenitor
Cell B: mid-differentiation
Cell C: terminal state

Pseudotime: order cells along a trajectory
             based on transcriptional similarity
```

---

## Monocle3 Workflow

```{r}
library(monocle3)
library(SeuratWrappers)
library(tidyverse)
library(here)

# Load annotated Seurat object from W8
seurat <- readRDS(here("data", "processed", "pbmc_annotated.rds"))

# Convert to CellDataSet (Monocle3 format)
cds <- as.cell_data_set(seurat)

# Preprocessing
cds <- cluster_cells(cds, resolution = 0.5)
cds <- learn_graph(cds, use_partition = TRUE,
                    verbose = FALSE)

# Learn trajectory graph
plot_cells(cds,
           color_cells_by   = "cell_type",
           label_groups_by_cluster = FALSE,
           label_leaves     = FALSE,
           label_branch_points = FALSE)
```

---

## Ordering Cells in Pseudotime

```{r}
# Define the root — where does the trajectory start?
# Biologically: the most undifferentiated cell state
# Select the "root" cells interactively or by marker gene

# Option 1: Automatic — cells with lowest UMI (proxy for progenitors)
root_cells <- colnames(cds)[which(
  Matrix::colSums(counts(cds)) == min(Matrix::colSums(counts(cds)))
)]

# Option 2: Manual — set root by a known marker gene
# e.g., HSCs in a hematopoiesis dataset: CD34+ cells

cds <- order_cells(cds, root_cells = root_cells)

# Visualize pseudotime
plot_cells(cds,
           color_cells_by   = "pseudotime",
           label_cell_groups = FALSE,
           label_leaves      = FALSE,
           trajectory_graph_color = "grey40",
           trajectory_graph_segment_size = 0.8)
```

---

## Pseudotime-Dependent Gene Expression

```{r}
# Which genes change significantly along pseudotime?
pr_test <- graph_test(cds,
                       neighbor_graph = "principal_graph",
                       cores          = 4)

pr_deg <- pr_test |>
  filter(q_value < 0.01, morans_I > 0.1) |>
  arrange(desc(morans_I))

# Top pseudotime-variable genes
cat("Pseudotime-variable genes:", nrow(pr_deg), "\n")
head(pr_deg)

# Visualize gene expression along pseudotime
plot_genes_in_pseudotime(
  cds[c("CD34", "GATA1", "HBA1", "HBB"), ],
  color_cells_by  = "pseudotime",
  min_expr        = 0.5
)
```

---

## Gene Modules Along Trajectory

```{r}
# Find co-regulated gene modules along pseudotime
gene_modules <- find_gene_modules(
  cds[pr_deg$gene_id, ],
  resolution     = 0.01,
  random_seed    = 42
)

cell_group_df <- tibble(
  cell    = row.names(colData(cds)),
  group   = colData(cds)$cell_type
)

agg_mat <- aggregate_gene_expression(
  cds, gene_modules, cell_group_df
)

pheatmap::pheatmap(
  t(scale(t(agg_mat))),
  clustering_method   = "ward.D2",
  fontsize            = 10
)
```

---

# RNA Velocity {background-color="#185FA5"}

---

## The Idea Behind RNA Velocity

```
For any gene in a cell:

Unspliced pre-mRNA (u):  newly transcribed → proxy for FUTURE state
Spliced mRNA (s):         mature message   → CURRENT state

If u > expected given s:  gene is being upregulated → cell moving in that direction
If u < expected given s:  gene is being downregulated → cell moving away
```

::: {.fragment}
RNA velocity predicts the **direction** and **speed** of cell state transitions —  
like an arrow on the UMAP showing where each cell is heading.
:::

---

## scVelo in R via reticulate

```{r}
library(reticulate)
library(Seurat)
library(here)

# scVelo is Python-based — use via reticulate
scv <- import("scvelo")

# Load loom file (contains spliced/unspliced counts)
# Generated by velocyto during Cell Ranger processing
ldata <- scv$read(here("data", "raw", "sample.loom"),
                   cache = TRUE)

# Basic preprocessing
scv$pp$filter_and_normalize(ldata,
                              min_shared_counts = 20,
                              n_top_genes       = 2000)
scv$pp$moments(ldata, n_pcs = 30, n_neighbors = 30)

# Estimate RNA velocity
scv$tl$velocity(ldata, mode = "stochastic")
scv$tl$velocity_graph(ldata)

# Plot
scv$pl$velocity_embedding_stream(
  ldata, basis = "umap",
  color         = "cell_type",
  title         = "RNA Velocity"
)
```

---

## Interpreting Velocity Plots

What to look for:

::: {.incremental}
- **Arrows pointing toward a cluster** → that cluster is a terminal state (attractor)
- **Arrows pointing away from a cluster** → that cluster is a progenitor
- **Circular arrows** → cycling cells (proliferating)
- **Scattered / no arrows** → cells in a stable state or not enough signal
:::

::: {.fragment}
::: {.callout-tip}
## Caveat
RNA velocity requires **unspliced** count information (velocyto loom files).  
Not all public datasets provide this.  
When not available: use Monocle3 pseudotime as the alternative.
:::
:::

---

# Cell–Cell Communication {background-color="#0F6E56"}

---

## Cells Don't Work Alone

::: {.incremental}
- **Cytokines** — IL-6, TNF-α, IFN-γ → paracrine signaling
- **Growth factors** — EGF, VEGF, FGF → receptor–ligand pairs
- **Extracellular matrix** — integrins, collagen → structural signals
- **Direct contact** — Notch–Jagged, E-cadherin
:::

::: {.fragment}
scRNA-seq lets us infer these interactions from expression data:

> If cell type A expresses **ligand X** and cell type B expresses **receptor Y**,  
> and X–Y is a known signaling pair,  
> → infer that A is signaling to B via X–Y.
:::

---

## CellChat Workflow

```{r}
library(CellChat)
library(Seurat)
library(patchwork)
library(here)

seurat <- readRDS(here("data", "processed", "pbmc_annotated.rds"))

# Create CellChat object
cellchat <- createCellChat(
  object   = seurat,
  group.by = "cell_type",
  assay    = "RNA"
)

# Use CellChatDB (human)
CellChatDB <- CellChatDB.human
cellchat@DB <- CellChatDB

# Subset to signaling genes only
cellchat <- subsetData(cellchat)
```

---

## Infer Communication Network

```{r}
# Identify overexpressed ligands and receptors
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)

# Compute communication probability
cellchat <- computeCommunProb(cellchat,
                               type    = "triMean",
                               nboot   = 100)

# Filter out weak interactions (few cells)
cellchat <- filterCommunication(cellchat, min.cells = 10)

# Infer at pathway level
cellchat <- computeCommunProbPathway(cellchat)

# Aggregate network
cellchat <- aggregateNet(cellchat)
```

---

## Visualizing Communication

```{r}
# Chord diagram — all interactions between cell types
netVisual_circle(
  cellchat@net$count,
  vertex.weight = as.numeric(table(cellchat@idents)),
  weight.scale  = TRUE,
  label.edge    = FALSE,
  title.name    = "Number of interactions"
)

# Top pathways
cellchat@netP$pathways

# Visualize a specific pathway — e.g., MHC-II
netVisual_aggregate(
  cellchat,
  signaling       = "MHC-II",
  layout          = "chord",
  vertex.receiver = c(1, 2, 3)   # T cell clusters
)
```

---

## Comparing CellChat — Disease vs. Control

```{r}
# Load two CellChat objects (disease vs. control)
cc_disease <- readRDS(here("data", "processed", "cellchat_disease.rds"))
cc_control <- readRDS(here("data", "processed", "cellchat_control.rds"))

# Merge for comparison
cc_list <- list(Control = cc_control, Disease = cc_disease)
cc_list <- lapply(cc_list, netAnalysis_computeCentrality)

# Compare number of interactions
compareInteractions(cc_list,
                    show.legend   = FALSE,
                    group         = c(1, 2),
                    measure       = "count")

# Differential interactions — gained/lost in disease
netVisual_diffInteraction(cc_list,
                           weight.scale = TRUE)
```

---

# Multi-Sample Integration with Harmony {background-color="#3B6D11"}

---

## The Batch Effect Problem in scRNA-seq

```
Patient A (sample 1) ─┐
Patient B (sample 2) ─┤ → Cluster by PATIENT, not cell type!
Patient C (sample 3) ─┘
(batch effect)
```

::: {.fragment}
**Solution — Harmony:**
- Integrates multiple samples / batches
- Preserves biological variation
- Removes technical variation
- Fast (minutes, not hours like Seurat integration)
:::

---

## Harmony Integration

```{r}
library(harmony)
library(Seurat)
library(here)

# Load multi-sample Seurat object
seurat_multi <- readRDS(here("data", "processed",
                              "multi_sample_seurat.rds"))

# Before Harmony — PCA colored by sample
DimPlot(seurat_multi, group.by = "sample_id",
        reduction = "pca") +
  ggtitle("Before integration — colored by sample")

# Run Harmony
seurat_multi <- RunHarmony(
  seurat_multi,
  group.by.vars   = "sample_id",
  reduction       = "pca",
  reduction.save  = "harmony",
  verbose         = FALSE
)

# Use Harmony embeddings for downstream analysis
seurat_multi <- seurat_multi |>
  RunUMAP(reduction = "harmony", dims = 1:30) |>
  FindNeighbors(reduction = "harmony", dims = 1:30) |>
  FindClusters(resolution = 0.5)

# After Harmony — cells should mix by biology, not sample
DimPlot(seurat_multi, group.by = "sample_id") +
  ggtitle("After Harmony — samples integrated")
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: Hematopoiesis (BioC)

```{r}
library(Seurat)
library(monocle3)
library(SeuratWrappers)
library(tidyverse)
library(here)

# Kleinberg et al. — human hematopoiesis
# HSC → progenitors → differentiated blood cells
# Perfect for trajectory analysis

library(scRNAseq)
sce_hsc <- KleinIESCData(which = "2i")

# Convert to Seurat
seurat_hsc <- as.Seurat(sce_hsc)

# Standard pipeline (abbreviated)
seurat_hsc <- NormalizeData(seurat_hsc) |>
  FindVariableFeatures() |>
  ScaleData() |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:20) |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:20)
```

---

## Hands-On: Trajectory

```{r}
# Convert to CellDataSet
cds <- as.cell_data_set(seurat_hsc)
cds <- cluster_cells(cds)
cds <- learn_graph(cds, verbose = FALSE)

# Set root — HSC cluster (CD34+, highest expression)
root_idx <- which.max(
  rowMeans(t(assay(cds, "logcounts")[c("Cd34", "Slamf1"), ]))
)

cds <- order_cells(cds,
                   root_cells = colnames(cds)[root_idx])

plot_cells(cds,
           color_cells_by = "pseudotime",
           label_branch_points = FALSE,
           label_leaves = FALSE,
           trajectory_graph_color = "grey40") +
  ggtitle("Hematopoiesis trajectory — pseudotime")
```

---

## Hands-On: CellChat on PBMC

```{r}
library(CellChat)

pbmc <- readRDS(here("data", "processed", "pbmc_annotated.rds"))

# Quick CellChat
cc <- createCellChat(pbmc, group.by = "cell_type", assay = "RNA")
cc@DB <- CellChatDB.human

cc <- subsetData(cc) |>
  identifyOverExpressedGenes() |>
  identifyOverExpressedInteractions() |>
  computeCommunProb(nboot = 20) |>   # low nboot for speed
  filterCommunication(min.cells = 5) |>
  computeCommunProbPathway() |>
  aggregateNet()

# How many interactions between each cell type pair?
groupSize <- as.numeric(table(cc@idents))
netVisual_circle(cc@net$count,
                  vertex.weight = groupSize,
                  weight.scale  = TRUE,
                  title.name    = "PBMC — interaction count")
```

---

# Next Week {background-color="#534AB7"}

---

## W10 Preview — Spatial Transcriptomics

**What we'll cover:**

- Visium, Xenium, CosMx — platform overview
- Seurat spatial pipeline
- Spatially variable genes (NNSVG)
- Cell type deconvolution (STdeconvolve)
- Integration with scRNA-seq reference

**Install before next class:**

```{r}
install.packages(c("Seurat", "STdeconvolve"))
BiocManager::install("NNSVG")
install.packages("SpatialDE")
```

---

## Key Resources

- **Monocle3** — [cole-trapnell-lab.github.io/monocle3](https://cole-trapnell-lab.github.io/monocle3)
- **CellChat** — [github.com/sqjin/CellChat](https://github.com/sqjin/CellChat)
- **Harmony** — [github.com/immunogenomics/harmony](https://github.com/immunogenomics/harmony)
- **scVelo** — [scvelo.readthedocs.io](https://scvelo.readthedocs.io)
- **CELLxGENE** — [cellxgene.cziscience.com](https://cellxgene.cziscience.com) — public scRNA-seq datasets

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Spatial Transcriptomics**

Course materials → [your GitHub Pages URL here]
:::
