---
title: "W13. Interpretation & Visualisation"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   ⚠️ Spot the four ways a figure routinely overstates its data
-   Choose a chart form from the job the data has to do
-   ⚠️ Say why "enriched for immune response" is almost never a result
-   Set the right background for an enrichment test, and say why it changes the answer
-   Write a figure caption as one sentence — before drawing the figure
-   Decide what belongs in a main figure and what belongs in the supplement
:::

> 🤔 **Motivating question**
>
> A colleague shows you a heatmap of 50 genes across 12 patients. Two clean blocks. What do you need to know before believing it?

------------------------------------------------------------------------

# 1. ⚠️ Four Ways a Figure Overstates 📉 {#w13-s1}

![None of these is dishonest. All four are defaults.](/assets/img/w13/figure_traps.svg){fig-align="center"}

::: {.callout-important collapse="true"}
## Back to the motivating question

Before believing the heatmap, you need three things:

-   ① **How were those 50 genes chosen?** If they were selected *because* they differ between the groups, then blocks are guaranteed — the selection created the pattern. This is circularity, and it is extremely common
-   ② **Was it row-scaled?** Z-scoring by row forces every gene to span the full colour range, so noise looks structured
-   ③ **What is the colour scale, and is it centred?** A diverging scale with an off-centre midpoint exaggerates one direction

→ ✅ The honest version: select genes on one split of the data, display them on the other. Or show an unselected set alongside.
:::

## 1) The others, briefly {#w13-s1-1}

-   ⚠️ **Truncated axes.** For bars, always start at zero — a bar's length *is* the value. If zero is uninformative, use points or a dot plot instead
-   ⚠️ **Rainbow colour maps.** Not perceptually ordered, invent boundaries where data is smooth, and fail in greyscale and for colour-vision deficiency. ✅ Use a single-hue sequential ramp for magnitude, and a two-hue diverging ramp with a neutral midpoint for signed values
-   ⚠️ **Bars that hide n.** With n ≤ 10, ✅ plot the individual points. A bar plus SEM at n = 3 implies a precision you do not have

------------------------------------------------------------------------

# 2. Choosing the Form 📐 {#w13-s2}

Pick the chart from the **job**, not from habit.

| The job | The form |
|--------------------------|----------------------------------------------|
| Compare a value across a few groups | Points per patient, box or mean overlaid |
| Show a distribution | Histogram, density, or violin **with points** |
| Relationship between two measures | Scatter, with the correlation stated |
| Magnitude across a grid | Heatmap — ✅ single-hue sequential |
| Signed change | Diverging ramp, neutral at zero |
| Change over time | Line, one line per patient where n allows |
| One headline number | ✅ Just write the number. Not every result needs a chart |

⚠️ **Never a dual y-axis.** Two scales on one plot lets the author choose the apparent relationship. Two panels, or index both to a common baseline.

::: {.callout-tip collapse="true"}
## Three habits that fix most omics figures

-   ✅ **Show the patients.** Points per patient, coloured by nothing else, is more convincing than any summary statistic
-   ✅ **Label directly.** Two to four series labelled on the plot beats a legend the reader has to look back and forth to
-   ✅ **Recede the furniture.** Grid lines and axes in light grey; the data is the only thing that should be dark
:::

------------------------------------------------------------------------

# 3. ⚠️ Enrichment Analysis 🧾 {#w13-s3}

![Seven terms. One fact. And the fact may be an artifact of the background.](/assets/img/w13/enrichment_traps.svg){fig-align="center"}

## 1) The background is the whole test {#w13-s3-1}

Enrichment asks whether your gene list contains more of a category than expected **by chance from some universe**. Change the universe, change the answer.

-   ❌ Default "all human genes" — but genes not expressed in your tissue could never have been in your list
-   ✅ The universe is **the genes you actually tested**: expressed above threshold, passing filters
-   ⚠️ In tissue-specific data this typically halves the universe, and it can remove an entire "significant" category

```{r}
library(clusterProfiler); library(org.Hs.eg.db)

universe <- rownames(dds)[rowSums(counts(dds)) > 10]   # ✅ what was testable

ego <- enrichGO(gene = sig_genes, universe = universe,     # ⚠️ not the default
                OrgDb = org.Hs.eg.db, keyType = "SYMBOL",
                ont = "BP", pAdjustMethod = "BH")

ego <- simplify(ego, cutoff = 0.7)     # collapse redundant nested terms
```

## 2) GSEA on the ranked list {#w13-s3-2}

✅ Often better than an over-representation test: it uses the whole ranking rather than a thresholded list, so there is no arbitrary cut-off, and modest coordinated shifts are detectable.

```{r}
ranks <- res |> filter(!is.na(stat)) |> arrange(desc(stat)) |>
  select(gene, stat) |> deframe()

gseaGO <- gseGO(ranks, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")
```

## 3) ⚠️ Then name genes {#w13-s3-3}

"Enriched for immune response" is compatible with almost any inflammatory tissue. **"Interferon signalling, driven by *IFI44L*, *ISG15* and *MX1*"** is a sentence a reader can act on — and check.

------------------------------------------------------------------------

# 4. One Message Per Figure 🎯 {#w13-s4}

![The left-hand version is what the analysis produced. The right-hand version is what the reader needs.](/assets/img/w13/one_message.svg){fig-align="center"}

✅ **Write the caption first.** If you cannot state the figure's point as one sentence with a blank for the number, the figure is not ready — and possibly the analysis is not either.

-   ✅ A caption should state the **direction**, the **n**, and the **number**
-   ✅ QC panels are not wasted; they belong in the supplement
-   ⚠️ A six-panel "characterisation" figure asks the reader to do your interpreting for you

------------------------------------------------------------------------

# 5. Hands-On 💻 {#w13-s5}

## 1) Take apart the figure you brought {#w13-s5-1}

Work in pairs. For your neighbour's figure:

-   ① Write the caption as one sentence. Can you?
-   ② Does the axis start at zero, and should it?
-   ③ Can you see the n? Can you see individual patients?
-   ④ If there is a heatmap: how were the genes chosen, and was it row-scaled?
-   ⑤ If there is a colour scale: sequential, diverging, or rainbow — and does it match the job?
-   ⑥ Which panel could move to the supplement without weakening the point?

## 2) Rebuild one plot {#w13-s5-2}

```{r}
library(tidyverse)

# ❌ before
ggplot(d, aes(group, value, fill = group)) +
  stat_summary(fun = mean, geom = "col") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2)

# ✅ after — the patients are visible, and n is on the axis
d |>
  mutate(group = fct_relevel(group, "control")) |>
  ggplot(aes(group, value)) +
  geom_boxplot(width = .45, outlier.shape = NA, colour = "#8A7FA3") +
  geom_jitter(width = .12, size = 2.6, alpha = .85, colour = "#185FA5") +
  scale_x_discrete(labels = \(x) paste0(x, "\n(n = ", table(d$group)[x], ")")) +
  labs(y = "CD8 T-cell fraction", x = NULL,
       title = "Responders have a higher CD8 T-cell fraction") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank())
```

## 3) Redo an enrichment with the right universe {#w13-s5-3}

```{r}
# Run enrichGO twice: default universe, then the expressed-gene universe.
# How many terms survive? Does the top term change?
```

💡 This takes two minutes and changes results often enough to be worth doing every time.

## 4) Before next week {#w13-s5-4}

**W14 — your own project.** Bring a one-sentence research question and a note of what data you could realistically get hold of.

------------------------------------------------------------------------

# 📝 Summary

-   ⚠️ Four routine overstatements: **truncated axes**, **rainbow colour maps**, **bars hiding n**, **row-scaled heatmaps of selected genes**
-   ⚠️ Selecting genes because they differ, then showing them as a heatmap, **creates** the blocks. Circularity
-   ✅ Choose the form from the job. Never a dual y-axis
-   ✅ Show the patients; label directly; recede the grid
-   ⚠️ **The enrichment background is the whole test.** Use the genes you actually tested, not the genome
-   ✅ Collapse redundant GO terms; prefer GSEA on the ranked list; prefer specific gene sets
-   ✅ **Then name genes** — "interferon signalling, driven by *ISG15*" beats "immune response"
-   ✅ **Write the caption first.** Direction, n, number — in one sentence
-   ✅ One message per figure; QC goes to the supplement

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| Truncated axis | A bar axis not starting at zero — exaggerates differences |
| Sequential ramp | One hue, light to dark; for magnitude |
| Diverging ramp | Two hues with a neutral midpoint; for signed values |
| Row scaling (z-score) | Rescaling each row — makes noise look structured |
| Circular selection | Choosing features by the effect you then display |
| Over-representation test | Is this category more frequent than expected? |
| Universe / background | The set your list could have been drawn from |
| GSEA | Enrichment on a ranked list, without a cut-off |
| Term redundancy | Nested GO terms reporting the same finding repeatedly |
| Direct labelling | Naming series on the plot instead of in a legend |

------------------------------------------------------------------------

# ➕ References

### Figures

Rougier NP, Droettboom M, Bourne PE. [*Ten simple rules for better figures.*](https://doi.org/10.1371/journal.pcbi.1003833) PLoS Comput Biol. 2014.\
Weissgerber TL, et al. [*Beyond bar and line graphs: time for a new data presentation paradigm.*](https://doi.org/10.1371/journal.pbio.1002128) PLoS Biol. 2015. **← the argument for showing the points**\
Crameri F, Shephard GE, Heron PJ. [*The misuse of colour in science communication.*](https://doi.org/10.1038/s41467-020-19160-7) Nat Commun. 2020. **← why rainbow maps fail**\
Wilke CO. [*Fundamentals of Data Visualization.*](https://clauswilke.com/dataviz/) (free online)

### Enrichment

Timmons JA, Szkop KJ, Gallagher IJ. [*Multiple sources of bias confound functional enrichment analysis of global -omics data.*](https://doi.org/10.1186/s13059-015-0761-7) Genome Biol. 2015. **← including the background problem**\
Wijesooriya K, et al. [*Urgent need for consistent standards in functional enrichment analysis.*](https://doi.org/10.1371/journal.pcbi.1009935) PLoS Comput Biol. 2022.\
Subramanian A, et al. [*Gene set enrichment analysis.*](https://doi.org/10.1073/pnas.0506580102) PNAS. 2005.\
Wu T, et al. [*clusterProfiler 4.0: a universal enrichment tool.*](https://doi.org/10.1016/j.xinn.2021.100141) Innovation. 2021.

### Circularity

Kriegeskorte N, et al. [*Circular analysis in systems neuroscience: the dangers of double dipping.*](https://doi.org/10.1038/nn.2303) Nat Neurosci. 2009. **← written for neuroscience, applies verbatim to omics heatmaps**
