---
title: "Network Analysis & Drug Repositioning"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 12"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W12 — Network & Drug Repositioning"
    css: custom.css
execute:
  echo: true
  eval: false
---

# From Gene Lists to Networks {background-color="#1D9E75"}

---

## The Limitation of Gene Lists

Every analysis we've done produces a gene list:

- GWAS → prioritized genes
- scRNA-seq → marker genes
- Multi-omics → convergent genes

::: {.fragment}
But genes don't act alone.

> A list of 200 DE genes tells you **what** is changing.  
> A network tells you **how** those genes are connected,  
> which ones are **hubs**, and what **drugs** might target the network.
:::

---

## What Network Analysis Adds

::: {.incremental}
- **Module detection** — groups of co-expressed / co-regulated genes
- **Hub genes** — highly connected nodes → often best drug targets
- **Shortest paths** — how does signal propagate from receptor to effector?
- **Drug target mapping** — which approved drugs hit nodes in this network?
- **Disease–drug similarity** — repositioning based on network proximity
:::

---

# Protein–Protein Interaction Networks {background-color="#534AB7"}

---

## STRINGdb — The Standard PPI Database

```{r}
library(STRINGdb)
library(igraph)
library(tidyverse)
library(here)

# Initialize STRINGdb (human, high confidence)
string_db <- STRINGdb$new(
  version      = "11.5",
  species      = 9606,         # Homo sapiens
  score_threshold = 700,       # high confidence (0–1000)
  network_type = "full",
  input_directory = here("data", "raw", "string")
)

# Map your gene list to STRING IDs
candidate_genes <- read_tsv(here("results",
                                  "prioritized_genes.tsv"))

genes_mapped <- string_db$map(candidate_genes,
                               "gene",
                               removeUnmappedRows = TRUE)

# Get interaction network
ppi_network <- string_db$get_interactions(genes_mapped$STRING_id)
head(ppi_network)
```

---

## Build igraph Network

```{r}
library(igraph)

# Create graph from PPI edges
g <- graph_from_data_frame(
  d        = ppi_network |>
    select(from = from, to = to, weight = combined_score),
  directed = FALSE,
  vertices = genes_mapped |>
    select(name = STRING_id, gene)
)

# Basic network statistics
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Density:", round(edge_density(g), 4), "\n")

# Hub genes — highest degree
hub_genes <- sort(degree(g), decreasing = TRUE) |> head(20)
cat("Top hubs:\n")
print(hub_genes)
```

---

## Network Visualization

```{r}
library(ggraph)
library(tidygraph)

# Convert to tidygraph
tg <- as_tbl_graph(g) |>
  mutate(
    degree      = centrality_degree(),
    betweenness = centrality_betweenness(),
    community   = as.factor(group_louvain())
  )

# Plot
ggraph(tg, layout = "fr") +
  geom_edge_link(aes(alpha = weight),
                 color    = "gray70",
                 show.legend = FALSE) +
  geom_node_point(aes(size  = degree,
                      color = community)) +
  geom_node_text(aes(label = gene,
                     filter = degree > quantile(degree, 0.9)),
                 repel = TRUE, size = 3) +
  scale_size_continuous(range = c(2, 10)) +
  labs(title = "PPI network — prioritized GWAS genes",
       color = "Module") +
  theme_graph()
```

---

## Community Detection — Gene Modules

```{r}
# Louvain community detection
community <- cluster_louvain(g, weights = E(g)$weight)

cat("Number of modules:", length(community), "\n")
cat("Modularity:", round(modularity(community), 3), "\n")

# Module membership
module_df <- tibble(
  STRING_id = V(g)$name,
  module    = membership(community)
) |>
  left_join(genes_mapped |> select(STRING_id, gene)) |>
  arrange(module)

# Genes per module
module_df |> count(module) |> arrange(desc(n))

# Functional enrichment per module
module_df |>
  group_by(module) |>
  summarise(genes = list(gene)) |>
  mutate(enrichment = map(genes, ~{
    string_db$get_enrichment(.x)
  }))
```

---

# Drug Target Identification {background-color="#185FA5"}

---

## Druggability of Network Hubs

```{r}
library(tidyverse)
library(here)

# Open Targets — drug-target associations database
# Download from https://platform.opentargets.org/downloads
ot_drugs <- read_parquet(
  here("data", "raw", "open_targets_drugs.parquet")
)

# Map network hubs to drug targets
hub_gene_names <- names(sort(degree(g), decreasing = TRUE))[1:50]

druggable_hubs <- ot_drugs |>
  filter(targetSymbol %in% hub_gene_names) |>
  select(gene       = targetSymbol,
         drug       = drugId,
         drug_name  = drugName,
         phase      = maximumClinicalTrialPhase,
         action     = actionType,
         indication = disease) |>
  arrange(desc(phase))

# Phase 3/4 drugs targeting our hub genes
druggable_hubs |>
  filter(phase >= 3) |>
  distinct(gene, drug_name, phase, action)
```

---

## Network Proximity — Disease–Drug Distance

```{r}
# Network proximity: how close are drug targets to disease genes?
# Short distance → drug is mechanistically relevant

# Disease module: genes from our analysis
disease_genes <- V(g)$name

# Drug targets: from DrugBank / Open Targets
drug_targets <- ot_drugs |>
  filter(phase >= 2) |>
  group_by(drug_name) |>
  summarise(targets = list(targetSymbol))

# Compute shortest path distance
# between each drug's targets and our disease genes

compute_proximity <- function(drug_tgts, disease_gns, graph) {
  # Get STRING IDs for drug targets
  dt_ids <- genes_mapped |>
    filter(gene %in% drug_tgts) |>
    pull(STRING_id)

  dg_ids <- genes_mapped |>
    filter(gene %in% disease_gns) |>
    pull(STRING_id)

  if (length(dt_ids) == 0 | length(dg_ids) == 0) return(NA_real_)

  d <- distances(graph, v = dt_ids, to = dg_ids)
  mean(apply(d, 2, min))   # average shortest path
}

# Run for all drugs
drug_proximity <- drug_targets |>
  mutate(proximity = map_dbl(targets, compute_proximity,
                              disease_gns = disease_genes,
                              graph = g)) |>
  arrange(proximity)

head(drug_proximity, 20)
```

---

# Drug Repositioning with CMAP/L1000 {background-color="#0F6E56"}

---

## The Connectivity Map Idea

::: {.incremental}
- **Disease signature** — genes UP in disease vs. control
- **Drug signature** — genes UP/DOWN after drug treatment (L1000)
- **Key insight:** if a drug's signature is the **reverse** of the disease signature
  → the drug might **reverse** the disease transcriptome
:::

::: {.fragment}
First demonstrated: ibuprofen reverses aging gene expression signature.  
Used to identify: thalidomide analogs for multiple myeloma, metformin for cancer.
:::

---

## Running CMAP Query

```{r}
library(cmapR)
library(here)

# Load L1000 GCTX file (CMap Phase II — 978 landmark genes)
ds <- parse_gctx(here("data", "raw",
                       "level5_beta_trt_cp_n720216x978.gctx"),
                 rid  = NULL,
                 cid  = NULL)

# Your disease signature
# Genes up in disease (from DESeq2 / scRNA-seq)
up_genes   <- c("IL6", "TNF", "CXCL8", "CCL2", "MMP9")
down_genes <- c("ADIPOQ", "PPARG", "IRS1", "GLUT4")

# Score each drug against your signature
# Kolmogorov-Smirnov enrichment test
ks_score <- function(drug_profile, up_set, down_set, n_total) {
  up_idx   <- which(names(drug_profile) %in% up_set)
  down_idx <- which(names(drug_profile) %in% down_set)

  ks_up   <- ks.test(up_idx,   1:n_total)$statistic
  ks_down <- ks.test(down_idx, 1:n_total)$statistic

  if (sign(ks_up) == sign(ks_down)) return(0)
  (ks_up - ks_down) / 2
}
```

---

## Visualizing CMAP Results

```{r}
# Top repositioning candidates (negative score = disease reversal)
cmap_results |>
  arrange(connectivity_score) |>
  slice_head(n = 20) |>
  ggplot(aes(x = connectivity_score,
             y = reorder(drug_name, -connectivity_score),
             fill = connectivity_score < -0.5)) +
  geom_col() +
  scale_fill_manual(values = c("gray70", "#1D9E75"),
                    labels  = c("Weak", "Strong reversal")) +
  geom_vline(xintercept = -0.5, linetype = "dashed",
             color = "#D85A30") +
  labs(x     = "Connectivity score",
       y     = "Drug",
       fill  = "Reversal strength",
       title = "CMAP — top disease-reversing drugs") +
  theme_minimal()
```

---

## Prioritizing Repositioning Candidates

Combine multiple lines of evidence:

```{r}
# Final drug prioritization table
drug_priority <- drug_proximity |>
  left_join(cmap_results |>
            select(drug_name, cmap_score = connectivity_score)) |>
  left_join(druggable_hubs |>
            select(drug_name, phase) |>
            group_by(drug_name) |>
            summarise(max_phase = max(phase))) |>
  mutate(
    # Normalize scores 0–1
    proximity_score = 1 - (proximity / max(proximity, na.rm = TRUE)),
    cmap_score_norm = abs(pmin(cmap_score, 0)),

    # Composite repositioning score
    repo_score = (proximity_score * 0.4) +
                 (cmap_score_norm * 0.4) +
                 (max_phase / 4 * 0.2)
  ) |>
  arrange(desc(repo_score))

head(drug_priority, 15)
```

---

# Mendelian Randomization Revisited {background-color="#3B6D11"}

---

## MR as Causal Drug Target Validation

> If the gene product is causally involved in disease (via MR),  
> **modulating it pharmacologically** should have a therapeutic effect.

```{r}
library(TwoSampleMR)

# Example: Is PCSK9 causally linked to CAD?
# Genetic variants that raise PCSK9 → raise LDL → CAD?

# Instruments: GWAS SNPs for PCSK9 expression (cis-eQTL)
pcsk9_instruments <- extract_instruments("eqtl-a-ENSG00000169174")

# Outcome: CAD GWAS
cad_outcome <- extract_outcome_data(
  snps     = pcsk9_instruments$SNP,
  outcomes = "ieu-a-7"
)

harmonised <- harmonise_data(pcsk9_instruments, cad_outcome)
mr_result  <- mr(harmonised)
mr_result

# If causal → PCSK9 inhibitors (evolocumab, alirocumab) should work
# → This is exactly how they were validated pre-trial
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: AD GWAS Network Analysis

Alzheimer's Disease GWAS genes → PPI network → drug targets

```{r}
library(STRINGdb)
library(igraph)
library(ggraph)
library(tidyverse)
library(here)

# AD GWAS-prioritized genes (from MAGMA, published)
ad_genes <- c("APOE", "BIN1", "CLU", "ABCA7", "CR1",
              "PICALM", "MS4A6A", "CD33", "EPHA1",
              "CD2AP", "SORL1", "FERMT2", "SLC24A4",
              "ZCWPW1", "CELF1", "NME8", "PTK2B", "MEF2C")

# Map to STRING
string_db <- STRINGdb$new(version = "11.5", species = 9606,
                           score_threshold = 400)
ad_df     <- data.frame(gene = ad_genes)
ad_mapped <- string_db$map(ad_df, "gene", removeUnmappedRows = TRUE)
```

---

## Hands-On: Build & Analyze Network

```{r}
# Get interactions
ppi <- string_db$get_interactions(ad_mapped$STRING_id)

# Build graph
g_ad <- graph_from_data_frame(
  d        = ppi |> select(from, to, weight = combined_score),
  directed = FALSE,
  vertices = ad_mapped |> select(name = STRING_id, gene)
)

# Network stats
cat("Nodes:", vcount(g_ad), "\n")
cat("Edges:", ecount(g_ad), "\n")

# Hub genes
V(g_ad)$degree <- degree(g_ad)
top_hubs <- sort(degree(g_ad), decreasing = TRUE) |> head(5)
cat("Top hub genes:", names(top_hubs), "\n")
```

---

## Hands-On: Visualize & Find Drug Targets

```{r}
# Visualize
tg_ad <- as_tbl_graph(g_ad) |>
  mutate(degree    = centrality_degree(),
         community = as.factor(group_louvain()))

ggraph(tg_ad, layout = "stress") +
  geom_edge_link(alpha = 0.3, color = "gray70") +
  geom_node_point(aes(size = degree, color = community)) +
  geom_node_label(aes(label = gene, filter = degree > 3),
                  repel = TRUE, size = 3) +
  theme_graph() +
  labs(title = "AD GWAS genes — PPI network",
       size  = "Degree", color = "Module")

# Which hub genes have approved drugs?
string_db$get_enrichment(ad_mapped$STRING_id) |>
  filter(category == "KEGG") |>
  arrange(fdr)
```

---

# Capstone Preparation {background-color="#993C1D"}

---

## The Full Analytical Journey

```
W2  GWAS         → which variants are associated?
W3  PRS          → quantify individual genetic risk
W4  Methylation  → which genes are epigenetically silenced?
W5  Chromatin    → which regulatory elements are active?
W6  16S          → what is the microbial community composition?
W7  WMS          → what are the microbes doing?
W8  scRNA-seq    → which cell types are involved?
W9  Trajectory   → how do cells transition in disease?
W10 Spatial      → where in the tissue?
W11 Multi-omics  → what converges across layers?
W12 Network      → what are the key hubs? Drug targets?
```

::: {.fragment}
**Your capstone:** pick a disease, a public dataset, and answer one question from this toolkit.
:::

---

## What Makes a Good Capstone?

::: {.incremental}
- **One clear biological question** — not "analyze everything"
- **Publicly available data** — GEO, TCGA, UK Biobank summary stats, curatedMetagenomicData
- **At least 2 analysis methods** from this course
- **A hypothesis** — not just a description
- **Clinical relevance** — why does this matter for patients?
:::

---

# Next Week {background-color="#534AB7"}

---

## W13 — Data Interpretation & Visualization

**What we'll cover:**

- Publication-quality figures in R (ggplot2 advanced)
- Combining plots with patchwork / cowplot
- Color theory for scientific figures
- Figure legends and reporting standards

**Also: capstone topic check-in**  
Come with a **one-sentence research question** and a **dataset in mind**.

---

## Key Resources

- **STRINGdb** — [string-db.org](https://string-db.org) — PPI database
- **Open Targets** — [platform.opentargets.org](https://platform.opentargets.org)
- **iLINCS** — [ilincs.org](http://www.ilincs.org) — CMAP/L1000 web interface
- **Clue.io** — [clue.io](https://clue.io) — Broad CMap portal
- **TwoSampleMR** — [mrcieu.github.io/TwoSampleMR](https://mrcieu.github.io/TwoSampleMR)

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Data Interpretation & Visualization**  
Start thinking about your capstone topic!

Course materials → [your GitHub Pages URL here]
:::
