---
title: "W7. Spatial Transcriptomics — The Basics"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Say what spatial data restores that dissociation destroyed
-   Place the major platforms on the resolution-versus-breadth trade-off, and pick one for a given question
-   ⚠️ Say why **FFPE compatibility** is the single most important practical fact in this chapter
-   ⚠️ Explain why a Visium spot is not a cell, and what follows from that
-   ⚠️ Recognise the physical artifacts — folds, edges, necrosis — that masquerade as biology
-   Say why neighbouring spots are not independent observations
-   State what your n is in a spatial experiment
:::

> 🤔 **Motivating question**
>
> Your single-cell data says TAM-like macrophages are enriched in non-responders. Your collaborator asks: *"where were they?"* Can you answer from the data you have?

------------------------------------------------------------------------

# 1. What Dissociation Threw Away 🗺️ {#w7-s1}

## 1) The same cell type, two different cells {#w7-s1-1}

![Both are annotated identically in the single-cell object.](/assets/img/w7/what_spatial_adds.svg){fig-align="center"}

Everything through W6 was done on cells in suspension. The protocol's first step destroys the variable that most of your clinical intuition runs on: **where the cell was**.

## 2) Questions only spatial can answer {#w7-s1-2}

-   **Architecture** — is the immune infiltrate inside the tumour, or stuck at the margin? (immune-excluded vs inflamed vs desert)
-   **Neighbourhoods** — which cell types are actually adjacent, as opposed to merely co-present
-   **Gradients** — distance from a vessel, from the invasion front, from a germinal centre
-   **Regional heterogeneity** — one part of the tumour resistant, another sensitive
-   **Linking back to histology** — the images pathologists have been reading for 150 years

💡 Notice how many of these are things a pathologist already assesses by eye. Spatial transcriptomics is, in large part, **putting molecular numbers on judgements your department already makes** — which is also why your pathologist is the most useful collaborator in this chapter.

------------------------------------------------------------------------

# 2. The Platforms 🔬 {#w7-s2}

## 1) One trade-off, everything else is detail {#w7-s2-1}

![Resolution and breadth are bought with the same budget.](/assets/img/w7/platform_tradeoff.svg){fig-align="center"}

-   **Sequencing-based** (Visium, Visium HD, GeoMx, Slide-seq) — capture transcripts on a grid or in chosen regions. ✅ Whole transcriptome. ⚠️ Each capture area may contain several cells
-   **Imaging-based** (Xenium, MERFISH, CosMx) — image individual molecules in place. ✅ Single-cell, sometimes subcellular. ⚠️ A targeted panel of a few hundred to a few thousand genes, chosen in advance

## 2) ⚠️ FFPE is the fact that matters most to you {#w7-s2-2}

Almost every current platform works on **formalin-fixed paraffin-embedded** tissue.

::: {.callout-important collapse="true"}
## Why this changes what is possible in a hospital

Your pathology department is storing blocks going back years or decades, and every one of them is attached to an outcome: who responded, who relapsed, who survived.

-   ✅ A **retrospective spatial study with real outcomes** is available to you now, without waiting five years to accrue a prospective cohort
-   ✅ You can select cases *by outcome* — responders vs non-responders — which is a far stronger design than profiling whoever consented
-   ⚠️ Old blocks vary in RNA quality. **DV200** is the standard metric; ask for it before committing
-   ⚠️ Ethics and consent for archived tissue is its own process. Start it early — it is usually the long pole

💡 This is the concrete answer to "what could I actually do with this?" for most residents in this course.
:::

## 3) Choosing {#w7-s2-3}

| Your question | Platform class |
|-----------------------------------|-------------------------------------|
| "Which cell types sit where?" — markers known | ✅ Imaging panel (Xenium, CosMx) |
| "What is transcriptionally different about this region?" | ✅ Whole-transcriptome (Visium, Visium HD) |
| "Compare these two histologically defined regions" | ✅ GeoMx — you draw the regions |
| "Are these two cell types touching?" | ✅ Imaging; ❌ standard Visium cannot resolve it |
| "Screen 60 patient blocks cheaply" | ⚠️ Cost dominates — talk to the core before designing |

------------------------------------------------------------------------

# 3. ⚠️ A Spot Is Not a Cell 🔍 {#w7-s3}

![The single most misread feature of spot-based spatial data.](/assets/img/w7/spot_is_not_a_cell.svg){fig-align="center"}

A standard Visium spot is 55 µm across and typically contains **1–10 cells**, often of several types. What you measure is their **sum**.

::: {.callout-warning collapse="true"}
## Three claims that do not follow from a spot

-   ❌ **"These two genes are co-expressed"** — they may be in different cells in the same spot. The single-cell co-expression fallacy (🔗 [W6 §3](../part3/w6_scrna_dynamics.qmd#w6-s3)) returns in a new costume
-   ❌ **"This spot is a macrophage"** — a spot has no cell type. It has a composition
-   ⚠️ **"These cell types are interacting"** — being in one 55 µm spot is weak evidence of contact. It is better than dissociated data, and it is not adjacency

→ ✅ What you can say: *this region is enriched for macrophage signal.* That is a real, useful, defensible claim — and getting from there to proportions is what W8 is about.
:::

------------------------------------------------------------------------

# 4. Analysing Spatial Data 📐 {#w7-s4}

## 1) The pipeline {#w7-s4-1}

```
QC (per spot, and against the H&E)
   → normalise
   → cluster in expression space  →  "spatial domains"
   → find spatially variable genes
   → relate domains to histology
   → [W8] deconvolve, then analyse neighbourhoods
```

⚠️ Note that standard clustering **ignores coordinates entirely** — it clusters spots the way Seurat clusters cells. That domains come out spatially coherent anyway is a good sign; that they do not is informative.

## 2) Spatially variable genes {#w7-s4-2}

Different question from differential expression: not *"does this gene differ between groups?"* but *"is this gene's expression organised in space at all?"*

-   Tools: `SpatialDE`, `nnSVG`, Moran's I, `SPARK`
-   ✅ Useful for discovery when you do not yet know which regions matter

## 3) ⚠️ Neighbouring spots are not independent {#w7-s4-3}

::: {.callout-important collapse="true"}
## Spatial autocorrelation, and why it inflates everything

Adjacent spots share diffused transcripts, share the same local tissue, and sometimes physically share cells. They are **positively correlated by construction**.

Any test that treats 4,000 spots as 4,000 independent observations will produce p-values as inflated as the per-cell tests in 🔗 [W5 §5.1](../part3/w5_scrna_qc.qmd#w5-s5-1) — and for the same reason.

-   ⚠️ "This domain has 3,412 differentially expressed genes, all *p* \< 10⁻¹⁵" is a symptom, not a result
-   ✅ Methods that account for spatial dependence exist (`nnSVG`, SPARK-X, mixed models with a spatial term)
-   ✅ And across patients, the fix is the same one as always: **aggregate to the patient, then compare patients**
:::

## 4) Domains and niches {#w7-s4-4}

Beyond clustering spots by expression, you can cluster by **what is around them** — the composition of the local neighbourhood. That is where the genuinely spatial biology lives, and it needs deconvolution first. 🔗 [W8](w8_spatial_integration.qmd).

------------------------------------------------------------------------

# 5. ⚠️ Tissue Reality 🧪 {#w7-s5}

![Every panel here has been published as biology at least once.](/assets/img/w7/tissue_reality.svg){fig-align="center"}

::: {.callout-tip collapse="true"}
## The reviewing question that catches most of these

*"Please show the H&E image beside the expression map, at the same scale."*

If a "novel spatial domain" coincides exactly with the edge of the section, a fold, or a necrotic area, that is visible in two seconds and invisible in any statistic.

💡 And the corresponding advice for your own work: sit down with your pathologist **before** picking the block, not after getting the data.
:::

⚠️ **Your n is still the patient.** Three sections from one tumour is n = 1 — the same rule as 🔗 [W1 §3.2](../part1/w1_intro.qmd#w1-s3-2) and 🔗 [W5 §5](../part3/w5_scrna_qc.qmd#w5-s5). Spatial experiments are expensive per sample, which pushes designs toward few patients and many spots. That is the wrong direction for anything you want to generalise.

------------------------------------------------------------------------

# 6. Hands-On 💻 {#w7-s6}

## 1) No code: look at a real section {#w7-s6-1}

[**10x Genomics datasets**](https://www.10xgenomics.com/datasets) → pick any Visium or Xenium dataset with an H&E image.

-   ① Find the H&E and the total-counts map. Do any high-count areas correspond to folds or thick regions?
-   ② Find the clustering figure. Does any cluster trace the **edge** of the section?
-   ③ How many patients contributed? How many sections each?
-   ④ 💡 Then look at the paper's claims and ask which of them survive ①–③

## 2) The pipeline {#w7-s6-2}

```{r}
library(Seurat); library(tidyverse); library(patchwork)

br <- LoadData("stxBrain", type = "anterior1")

# ⚠️ QC against the image, not just the numbers
SpatialFeaturePlot(br, features = "nCount_Spatial") +
  ggtitle("Look for folds and edges before anything else")

br <- SCTransform(br, assay = "Spatial", verbose = FALSE) |>
  RunPCA() |> FindNeighbors(dims = 1:30) |>
  FindClusters(resolution = 0.6) |> RunUMAP(dims = 1:30)

# the two views that must always appear together
DimPlot(br, reduction = "umap", label = TRUE) |
  SpatialDimPlot(br, label = TRUE, label.size = 3)

# ✅ save it — W8 picks this object up
saveRDS(br, here::here("analysis", "7_visium.rds"))
```

## 3) Spatially variable genes {#w7-s6-3}

```{r}
svg_genes <- FindSpatiallyVariableFeatures(
  br, assay = "SCT", features = VariableFeatures(br)[1:1000],
  selection.method = "moransi")

top <- SpatiallyVariableFeatures(br, selection.method = "moransi")[1:6]
SpatialFeaturePlot(br, features = top, ncol = 3, alpha = c(0.1, 1))
```

## 4) ✅ Change one thing {#w7-s6-4}

```{r}
# Are the domains driven by biology, or by how much tissue was in each spot?
br$log_counts <- log10(br$nCount_Spatial)
VlnPlot(br, "log_counts", group.by = "seurat_clusters")
```

💡 If one cluster has systematically higher total counts, ask whether it is a tissue region or a **thickness** region before naming it.

## 5) Before next week {#w7-s6-5}

**W8 — deconvolution and integration.** Bring the answer to the question from W6: how many FFPE blocks with linked outcomes does your department hold?

------------------------------------------------------------------------

# 📝 Summary

-   Spatial data restores **position** — the variable dissociation destroys, and the one most of your clinical reasoning uses
-   Resolution and transcriptome breadth trade against each other. Imaging = single cell, targeted panel. Sequencing = whole transcriptome, multi-cell spots
-   ✅ **Most platforms work on FFPE** — which puts a retrospective, outcome-linked cohort within reach of a resident today
-   ⚠️ **A 55 µm spot is 1–10 cells.** It has a composition, not a cell type. Co-expression within a spot is not co-expression within a cell
-   ⚠️ Standard clustering ignores coordinates; spatially coherent domains are a good sign, not a guarantee
-   ⚠️ **Neighbouring spots are not independent.** Thousands of tiny p-values is a symptom of ignoring that
-   ⚠️ Folds, section edges and necrosis all produce convincing "domains". ✅ Always show the H&E beside the map
-   ⚠️ **n is still the patient.** Three sections from one tumour is n = 1

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| Spot / capture area | One measurement location; on Visium, 55 µm and several cells |
| Visium HD | Continuous 2 µm binning — near single-cell, whole transcriptome |
| Imaging-based ST | Direct molecular imaging (Xenium, MERFISH, CosMx); targeted panel |
| GeoMx | Region-of-interest profiling — you choose the areas |
| FFPE | Formalin-fixed paraffin-embedded; the pathology archive format |
| DV200 | RNA quality metric for FFPE — the number to ask for |
| Spatial domain | A cluster of spots with similar expression |
| Spatially variable gene | A gene whose expression is organised in space |
| Moran's I | Classical statistic for spatial autocorrelation |
| Spatial autocorrelation | Nearby measurements resemble each other by construction |
| Niche / neighbourhood | The local cellular composition around a location (W8) |

------------------------------------------------------------------------

# ➕ References

### Overviews

Moses L, Pachter L. [*Museum of spatial transcriptomics.*](https://doi.org/10.1038/s41592-022-01409-2) Nat Methods. 2022. **← the landscape, honestly assessed**\
Rao A, Barkley D, França GS, Yanai I. [*Exploring tissue architecture using spatial transcriptomics.*](https://doi.org/10.1038/s41586-021-03634-9) Nature. 2021.\
Bressan D, Battistoni G, Hannon GJ. [*The dawn of spatial omics.*](https://doi.org/10.1126/science.abq4964) Science. 2023.

### Methods and statistics

Weber LM, et al. [*nnSVG for the scalable identification of spatially variable genes using nearest-neighbor Gaussian processes.*](https://doi.org/10.1038/s41467-023-39748-z) Nat Commun. 2023.\
Svensson V, Teichmann SA, Stegle O. [*SpatialDE: identification of spatially variable genes.*](https://doi.org/10.1038/nmeth.4636) Nat Methods. 2018.\
Palla G, et al. [*Squidpy: a scalable framework for spatial omics analysis.*](https://doi.org/10.1038/s41592-021-01358-2) Nat Methods. 2022.

### Clinical and FFPE

Janesick A, et al. [*High resolution mapping of the tumor microenvironment using integrated single-cell, spatial and in situ analysis.*](https://doi.org/10.1038/s41467-023-43458-x) Nat Commun. 2023.\
[10x Genomics public datasets](https://www.10xgenomics.com/datasets) · [Squidpy documentation](https://squidpy.readthedocs.io)
