---
title: "W2 — GWAS & Post-GWAS"
subtitle: "Clinical Genomics for Bioinformatics"
author: "Eunji Ha"
date: last-modified
format:
  html
execute:
  eval: false      # code is for teaching; not executed at render time
  echo: true
  warning: false
  message: false
---

# Why GWAS belongs in this course {background-color="#1D9E75"}

------------------------------------------------------------------------

## GWAS is directly clinical

::: incremental
-   **Drug target validation** — targets with human genetic support are \~2× more likely to succeed to approval *\[Reay & Cairns 2021\]*
-   **Disease subtyping** — same diagnosis, different genetic architecture → different biology & treatment
-   **Risk stratification** — polygenic risk scores in preventive medicine (W3)
-   **Shared aetiology** — genetic correlation quantifies overlap between diseases
-   **Causal inference** — Mendelian randomization tests causality without an RCT
:::

::: fragment
▶️ Open a recent *Nature Medicine* paper — there's a good chance GWAS data sits underneath the clinical claim.
:::

------------------------------------------------------------------------

## What a GWAS actually asks

::: fragment
**"Which genetic variants are more common in people with disease X than in matched people without it?"**
:::

::: incremental
-   Compares allele frequencies at **millions of variants** at once
-   **Hypothesis-free** — agnostic to which gene matters
-   Output is a list of associated **loci**, not causal **genes** (yet)
:::

SNP (single-nucleotide polymorphism)

:   a single base-pair position where two or more alleles occur commonly (MAF ≥1%).

    Humans carry \~10 million common SNPs; arrays measure 0.5–5M directly and **impute** the rest.

# Genetic architecture: the shape of the signal {#sec-architecture background-color="#534AB7"}

------------------------------------------------------------------------

## Know your trait before you run anything

**Genetic architecture** = the landscape of genetic contributions: *how many* variants, *how large* their effects, *how frequent* they are, and how they interact *\[Timpson 2017\]*.

| Architecture | Variants | Example | Tools |
|------------------|------------------|------------------|------------------|
| **Monogenic** | one | phenylketonuria, CF (*CFTR*) | family, WES/WGS |
| **Oligogenic** | few | some lipid traits | targeted seq |
| **Polygenic** | many, small | height, T2D | large GWAS |
| **Omnigenic** | \~all expressed genes | most complex traits | huge GWAS + networks |

::: fragment
The **omnigenic model**: regulatory networks are so interconnected that nearly every expressed gene exerts a tiny "peripheral" effect, funnelled through fewer "core" genes — this is *why* PRS (W3) works.
:::

------------------------------------------------------------------------

## Allele frequency × effect size

Common variants tend to have **small** effects; large-effect variants tend to be **rare** (purifying selection removes common, large-effect, deleterious variants) *\[Timpson 2017\]*.

```         
 effect │ rare, large effect
  size  │  •           (Mendelian-like)
        │     • •
        │        • • •
        │             • • • • •   common, small effect
        │                       • • • • •  (typical GWAS hits)
        └──────────────────────────────────── MAF
```

::: incremental
-   **Sample size is destiny** — schizophrenia: \~0 hits at *n*≈hundreds, **113 loci** at *n* in tens of thousands
-   **Effect size ≠ druggability** — common *HMGCR* SNP explains \~0.26% of LDL variance, yet statins are transformative
:::

::: callout-important
Architecture can't be predicted from heritability alone. Choose your study design *after* you understand the likely architecture.
:::

# Study design & QC {#sec-design background-color="#185FA5"}

------------------------------------------------------------------------

## GWAS study design

```         
Participants
├── Cases    (n ≥ 5,000)    patients with the disease
└── Controls (n ≥ 10,000)   ancestry-matched, disease-free
          ↓
Genotyping array  →  imputation to a reference panel
          ↓
Quality control (samples + variants)
          ↓
Per-variant association test (+ covariates)
          ↓
Summary statistics → Manhattan / QQ → post-GWAS
```

::: callout-note
**Sample size is everything.** Early GWAS (*n*≈2,000) found almost nothing; biobank GWAS (*n*\>500K) find hundreds of loci. Most "failed" early studies were just underpowered for a polygenic architecture.
:::

------------------------------------------------------------------------

## Quality control

| QC step | Threshold | Why |
|----|----|----|
| Sample call rate | \> 95–98% | poor DNA / failed samples |
| SNP call rate | \> 99% | unreliable genotyping |
| Hardy–Weinberg eq. | *p* \> 1×10⁻⁶ (controls) | genotyping error |
| Minor allele freq. | \> 1% | underpowered for rarer variants |
| Heterozygosity | ±3 SD | contamination / inbreeding |
| Relatedness (kinship) | π̂ \< 0.125 | inflated statistics |
| Imputation INFO/R² | \> 0.8 | unreliable dosages |
| **Stratification** | PCA | **biggest confounder** |

------------------------------------------------------------------------

## Population stratification

::::: columns
::: {.column width="50%"}
If cases & controls differ in **ancestry**, *any* ancestry-associated variant looks like a disease hit.

▶️ **Fix:** PCA on genotypes → include top PCs as covariates so ancestry is absorbed by the model.
:::

::: {.column width="50%"}
``` r
glm(disease ~ SNP +
      PC1 + PC2 + PC3 + PC4 + PC5 +  # ancestry
      age + sex + batch,             # nuisance
    family = binomial)
```
:::
:::::

::: callout-important
## Always report ancestry — and its trade-off

Mixing populations uncorrected → false positives. Restricting to one ancestry → cleaner stats but **limited generalizability** (see [Diversity](#sec-diversity)).
:::

# Reading GWAS results {#sec-reading background-color="#185FA5"}

------------------------------------------------------------------------

## The Manhattan plot

```{r}
library(qqman)

manhattan(gwas_results,
          chr = "CHR", bp = "BP", p = "P", snp = "SNP",
          col = c("#185FA5", "#1D9E75"),
          suggestiveline = -log10(1e-5),   # dashed
          genomewideline = -log10(5e-8),   # solid
          highlight = lead_snps,
          main = "GWAS — Type 2 Diabetes (n = 150,000)")
```

Each point = a variant; *x* = position, *y* = −log₁₀(*p*). "Skyscrapers" are associated loci.

------------------------------------------------------------------------

## Why the threshold is 5×10⁻⁸

::: incremental
-   \~1 million **independent** common-variant tests after LD pruning
-   Bonferroni: 0.05 / 1,000,000 = **5×10⁻⁸** → genome-wide significant
-   `p < 1×10⁻⁵` → conventional **suggestive** line
:::

::: callout-note
This is calibrated for **common SNPs**. Rare-variant and structural-variant tests (see [Beyond SNPs](#sec-beyond)) need their own thresholds.
:::

------------------------------------------------------------------------

## QQ plot & genomic inflation (λ)

```{r}
qq(gwas_results$P, main = "QQ — observed vs expected −log10(p)")

chisq  <- qchisq(gwas_results$P, df = 1, lower.tail = FALSE)
lambda <- median(chisq) / qchisq(0.5, df = 1)
cat("Genomic inflation λ =", round(lambda, 3), "\n")
```

::: incremental
-   **Well-calibrated**: hug the diagonal, lift off at the tail (true signal)
-   **Inflated (λ \> 1.05)**: all above diagonal → uncorrected stratification / relatedness
-   **Deflated (λ \< 0.95)**: over-conservative QC → lost power
-   At very large *n*, modest λ can be genuine **polygenicity** — LD-score regression separates the two
:::

# Beyond SNPs {#sec-beyond background-color="#BA7517"}

------------------------------------------------------------------------

## Variation that arrays miss

For some traits, common-SNP GWAS nears **signal saturation** — the next frontier is variation arrays capture poorly *\[Harris 2025\]*.

| Variant class | What it is | Clinical relevance |
|------------------------|------------------------|------------------------|
| **CNV** | del/dup, kb–Mb | dosage effects; under-tested |
| **Structural** | inversions, translocations | disrupt genes / distal regulation |
| **Tandem repeats** | variable repeat counts | expansions → neuro disorders |
| **HLA / MHC** | hyper-polymorphic | dominates autoimmune signals |

------------------------------------------------------------------------

## Two examples worth remembering

::: incremental
-   **UGT2B17 deletion → prostate cancer** — alters testosterone metabolism; frequency varies by population → **population-specific risk**, invisible to SNP-only analysis *\[Harris 2025\]*
-   **CYP2D6 copy number → drug dosing** — deletions vs duplications shift codeine→morphine conversion (poor vs ultra-rapid metabolizers). **Pharmacogenetics**: the variant changes the right *dose*
:::

::: callout-note
## Dosage sensitivity & "human knockouts"

Sequencing enables base-pair CNV calling at biobank scale. Aggregating loss-of-function variants (gnomAD **LOEUF**) builds **dosage-sensitivity maps** flagging genes intolerant to disruption — directly useful for clinical variant interpretation.
:::

# Post-GWAS: from hit to biology {#sec-postgwas background-color="#534AB7"}

------------------------------------------------------------------------

## A hit is rarely the cause

A GWAS hit is usually a **tag** in LD with the real driver.

```         
rs1234567  p = 3×10⁻²⁴            ← lead SNP (reported)
   ├── rs1234568  r² = 0.98
   ├── rs1234591  r² = 0.95       ← one (or more) is causal
   └── rs1234612  r² = 0.91
   (inherited together in a haplotype block)
```

**Post-GWAS pipeline:** 1) Fine-mapping → 2) Functional annotation → 3) Variant→gene → 4) Causal inference

------------------------------------------------------------------------

## Step 0 — Linkage disequilibrium

LD both **enables** GWAS (one tag stands in for many) and **complicates** it (can't tell tag from cause by *p* alone).

```{r}
library(LDlinkR)

ld_matrix <- LDmatrix(
  snps  = c("rs1234567", "rs1234568", "rs1234591"),
  pop   = "CEU",            # 1000 Genomes population
  r2d   = "r2",
  token = "your_token"      # free at ldlink.nci.nih.gov
)

LDproxy(snp = "rs1234567", pop = "CEU", r2d = "r2", token = "your_token")
```

::: callout-warning
The LD reference panel **must match your sample's ancestry** — a CEU panel on an East Asian cohort corrupts fine-mapping *\[Schaid 2018\]*.
:::

------------------------------------------------------------------------

## Step 1 — Fine-mapping {#sec-finemap}

Modern fine-mapping is **Bayesian**: compute a **posterior inclusion probability (PIP)** per variant, then a **credible set** — the smallest set containing the causal variant(s) with (e.g.) 95% probability *\[Schaid 2018\]*.

```{r}
library(susieR)

# SuSiE = Sum of Single Effects — allows multiple causal signals
susie_fit <- susie(X = genotype_matrix, y = phenotype_vector, L = 10)

cs  <- susie_get_cs(susie_fit, coverage = 0.95)  # 95% credible set(s)
pip <- susie_get_pip(susie_fit)                  # per-variant PIP
```

::: callout-note
Fine-mapping shrinks **\~200 correlated SNPs** to a credible set of a handful — *if* the data allow it.
:::

------------------------------------------------------------------------

## What controls fine-mapping resolution

Expected PIP rises with **λ = N·R²/(1−R²)** and falls as nearby SNPs get more correlated *\[Schaid 2018\]*:

::: incremental
-   **Larger *N*** → tighter credible sets (cheapest win)
-   **High LD (ρ→1)** → causal variants statistically indistinguishable in one sample
-   **Functional annotation as a prior** → modest help (90% set \~17.5 → \~13.5 SNPs for lipids); matters most in weak, high-LD regions
:::

:::: fragment
::: callout-tip
## Robustness checklist

-   PIPs spread across many SNPs? → you likely **can't** resolve one variant here
-   LD reference matches the sample? (compare allele frequencies)
-   Do **SuSiE / FINEMAP / PAINTOR** agree?
-   Multi-ancestry: assume **shared causal variants**, allow **different effect sizes**
:::
::::

------------------------------------------------------------------------

## Step 2 — Functional annotation

| Location                   | Likely mechanism                     |
|----------------------------|--------------------------------------|
| Coding (missense/nonsense) | direct protein change                |
| Splice site                | altered mRNA / isoform               |
| Promoter / 5′UTR           | transcription regulation             |
| Enhancer                   | distal gene regulation               |
| **Intergenic / intronic**  | **unknown — needs eQTL / chromatin** |

::: callout-important
**\~90% of GWAS hits are non-coding**, enriched in regulatory elements and eQTLs — which is exactly why epigenomics (W4–W5) is essential for interpretation.
:::

------------------------------------------------------------------------

## Step 3 — Variant → gene: colocalization {#sec-var2gene}

An **eQTL** affects a gene's expression. **Colocalization** tests whether the GWAS and eQTL signals share the *same* causal variant.

```{r}
library(coloc)

coloc_result <- coloc.abf(
  dataset1 = list(pvalues = gwas_pvals, N = 150000, type = "cc", s = 0.3),
  dataset2 = list(pvalues = eqtl_pvals, N = 500,    type = "quant")
)
coloc_result$summary   # PP.H4 = share one causal variant
```

::: callout-tip
**PP.H4 \> 0.8** → strong evidence the variant acts *through* that gene → strong effector candidate.
:::

::: callout-warning
## Watch direction of effect (coherence)

A risk allele can *down*-regulate a gene that looks *up*-regulated in patients ("incoherence") — often an impaired protective response. Direction decides agonist vs antagonist *\[Reay & Cairns 2021\]*.
:::

------------------------------------------------------------------------

## Step 3 (cont.) — which cell type?

Colocalization names a gene; it doesn't say *which cell*. Frameworks like **sc-linker** integrate **scRNA-seq** with GWAS via **stratified LD-score regression (S-LDSC)** *\[Minton 2022\]*:

::: incremental
-   Recovered known biology (T cells in eczema; microglia in Alzheimer's & MS)
-   Surfaced candidates (GABAergic neurons in depression — current drugs target serotonin)
-   Separated **disease-state** programmes (M cells in ulcerative colitis) from healthy-state
:::

::: fragment
A locus needs **both** a gene (coloc) **and** a cell type/state (single-cell) before you claim a mechanism.
:::

------------------------------------------------------------------------

## Step 4 — Mendelian randomization

> *Does elevated LDL **cause** CAD, or just correlate?*

Variants = **instruments**; randomized at conception → mimic a lifelong RCT, immune to lifestyle/SES confounding.

```         
Genetic variants → LDL level → CAD
   (instrument)    (exposure)  (outcome)
```

```{r}
library(TwoSampleMR)

exposure_dat <- extract_instruments("ieu-a-300")            # LDL
outcome_dat  <- extract_outcome_data(snps = exposure_dat$SNP,
                                     outcomes = "ieu-a-7")  # CAD
dat    <- harmonise_data(exposure_dat, outcome_dat)
result <- mr(dat)                                            # IVW + sensitivity
```

::: callout-warning
**IV assumptions**: (1) relevance, (2) independence of confounders, (3) exclusion restriction. #3 is broken by **pleiotropy** — always run MR-Egger / weighted median *\[Reay & Cairns 2021\]*.
:::

# GWAS for drug repurposing {#sec-drugs background-color="#0F6E56"}

------------------------------------------------------------------------

## Existing drugs sit on GWAS signals

Human genetic evidence \~doubles clinical success, so loci are mined for targets *\[Reay & Cairns 2021\]*.

| Trait                | Gene    | Drug / class             |
|----------------------|---------|--------------------------|
| Schizophrenia        | *DRD2*  | antipsychotics           |
| Type 2 diabetes      | *PPARG* | thiazolidinediones       |
| Breast cancer        | *ESR1*  | tamoxifen                |
| Bone mineral density | *RANKL* | denosumab                |
| Rheumatoid arthritis | *IL6R*  | tocilizumab              |
| Psoriasis            | *IL23R* | ustekinumab              |
| LDL cholesterol      | *HMGCR* | statins                  |
| Atrial fibrillation  | *SCN5A* | class-1a antiarrhythmics |

::: callout-note
A SNP *near* a gene doesn't prove the gene is causal — follow-up required.
:::

------------------------------------------------------------------------

## Three genetic strategies for repurposing

::: incremental
-   **Locus → target** — Open Targets Genetics, Priority index, MAGMA, TWAS (FUSION, S-PrediXcan)
-   **MR for target validation** — *cis*-variants in a druggable gene as instrument (PCSK9, IL6R)
-   **Direction of effect** — decides agonist vs antagonist (ties back to coherence)
:::

# PheWAS: flip the question {#sec-phewas background-color="#BA7517"}

------------------------------------------------------------------------

## One variant, many phenotypes

```         
GWAS   : one disease  → scan all variants    → "which variants?"
PheWAS : one variant  → scan all phenotypes   → "what else does it do?"
```

Typically run over **electronic health records** (ICD/CPT codes as phenotypes) *\[Denny 2016\]*.

::: incremental
-   **Pleiotropy mapping** — repurposing leads *and* predictable side-effects
-   **EHR-scale** — one biobank → thousands of tests at once
-   **Pharmacogenomic PheWAS ("MedWAS")** — variant → drug response / adverse events (free-text extraction is the bottleneck)
:::

# Diversity: one ancestry is not enough {#sec-diversity background-color="#5B4D80"}

------------------------------------------------------------------------

## Why diverse cohorts matter

Most GWAS (→ fine-mapping → PRS) are **European-ancestry** — a scientific *and* equity problem *\[Wojcik 2019\]*.

::: incremental
-   **Missed discovery** — variants rare/absent in Europeans are invisible; multi-ethnic PAGE found new associations
-   **Better fine-mapping** — different LD patterns break haplotype blocks → **smaller credible sets**
-   **Poor portability** — effect sizes & risk scores don't transfer → biased clinical use **widens disparities**
:::

::: callout-important
## For Korean cohorts

Use Korean / East-Asian resources (KoGES, Korea Biobank Array "KCHIP", TOPMed). Check your **base GWAS ancestry matches your target** before trusting an association, credible set, or PRS.
:::

# Hands-on {#sec-hands background-color="#BA7517"}

------------------------------------------------------------------------

## Dataset: T2D summary statistics

Public T2D GWAS summary stats (e.g. Mahajan et al. 2022, *Nat Genet*) — one row per variant, no individual genotypes, hence shareable.

```{r}
library(tidyverse)
library(here)
library(qqman)

gwas <- read_tsv(here("data", "raw", "t2d_gwas_summary.tsv"))
glimpse(gwas)
# CHR, BP, SNP, A1, A2, FREQ, BETA, SE, P, N, INFO
```

------------------------------------------------------------------------

## Step 1 — QC on summary stats

```{r}
gwas_clean <- gwas |>
  filter(!is.na(P), P > 0, P <= 1,
         FREQ > 0.01, FREQ < 0.99,   # common variants
         INFO > 0.8) |>              # imputation quality
  mutate(CHR = as.integer(CHR),
         neglogP = -log10(P))

cat("SNPs before QC:", nrow(gwas), "\n")
cat("SNPs after  QC:", nrow(gwas_clean), "\n")
```

------------------------------------------------------------------------

## Step 2 — QQ plot & λ

```{r}
chisq  <- qchisq(gwas_clean$P, df = 1, lower.tail = FALSE)
lambda <- median(chisq) / qchisq(0.5, df = 1)
cat("Lambda (GC) =", round(lambda, 4), "\n")

qq(gwas_clean$P,
   main = paste0("QQ — T2D GWAS  (λ = ", round(lambda, 3), ")"),
   col  = "#185FA5")
```

------------------------------------------------------------------------

## Step 3 — Manhattan plot

```{r}
sig_snps <- gwas_clean |> filter(P < 5e-8) |> arrange(P)
cat("Genome-wide significant variants:", nrow(sig_snps), "\n")

manhattan(gwas_clean,
          chr = "CHR", bp = "BP", p = "P", snp = "SNP",
          col = c("#185FA5", "#1D9E75"),
          suggestiveline = -log10(1e-5),
          genomewideline = -log10(5e-8),
          highlight      = head(sig_snps$SNP, 10),
          main           = "T2D GWAS — Manhattan plot")
```

------------------------------------------------------------------------

## Step 4 — Annotate top hits

```{r}
library(biomaRt)

mart     <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
top_hits <- sig_snps |> slice_min(P, n = 20)

annotations <- getBM(
  attributes = c("chromosome_name", "start_position",
                 "end_position", "external_gene_name"),
  filters    = "chromosomal_region",
  values     = paste0(top_hits$CHR, ":",
                      top_hits$BP - 5e5, ":", top_hits$BP + 5e5),
  mart       = mart
)
head(annotations)
```

------------------------------------------------------------------------

## Step 5 — Regional (locus-zoom) plot

```{r}
library(gassocplot)

# TCF7L2 — most replicated T2D locus
tcf7l2 <- gwas_clean |>
  filter(CHR == 10, BP >= 114700000, BP <= 115000000)

stack_assoc_plot(
  markers = tcf7l2[, c("SNP", "CHR", "BP")],
  z       = tcf7l2$neglogP,
  corr    = ld_matrix_tcf7l2,   # from LDlinkR, ancestry-matched
  traits  = "T2D",
  ylab    = "-log10(p)"
)
```

------------------------------------------------------------------------

## Interrogating any locus

::: incremental
1.  **Replicated?** in an independent cohort?
2.  **LD structure?** how big is the credible set?
3.  **Functional context?** coding, regulatory, eQTL?
4.  **Which gene & cell type?** nearest ≠ causal
5.  **Effect size & direction?** small; direction guides drug strategy
6.  **Druggable?** known/tractable target?
:::

# Wrap-up {background-color="#534AB7"}

------------------------------------------------------------------------

## What GWAS cannot tell you

::: incremental
-   **Causality alone** — association ≠ causation (MR helps, with assumptions)
-   **Rare variants** — common-SNP GWAS misses MAF \< 1% (need WES/WGS)
-   **Non-SNP variation** — CNV/SV/repeats need dedicated methods
-   **Mechanism alone** — an enhancer hit says "something regulates something"
-   **Transferability** — Euro-centric results may not generalize
-   **Gene × environment** — static associations miss dynamic biology
:::

::: fragment
> **GWAS is the beginning of the story, not the end** — which is why the rest of the course exists.
:::

------------------------------------------------------------------------

## Where this connects

```         
GWAS hit identified
   ├─ W3       PRS            → genome-wide risk prediction
   ├─ W4–W5    Epigenomics    → active enhancer?
   ├─ W6–W7    Metagenomics   → microbiome mediation?
   ├─ W8–W9    scRNA-seq      → which cell type?
   ├─ W10      Spatial        → where in tissue?
   ├─ W11      Multi-omics    → convergent evidence
   └─ W12      Drug target    → druggable?
```

------------------------------------------------------------------------

## Next week — PRS (W3)

::: incremental
-   Single hits → genome-wide prediction
-   Construction: C+T, LDpred2, PRSice2, PRS-CS(x)
-   Clinical utility — screening, stratification, barriers
-   Hands-on: build a PRS from summary stats
:::

```{r}
install.packages(c("bigsnpr", "bigstatsr"))
# PRSice2 is a command-line binary: https://www.prsice.info/
```

------------------------------------------------------------------------

## Key resources

-   **GWAS Catalog** — <https://www.ebi.ac.uk/gwas>
-   **Open Targets Genetics** — <https://genetics.opentargets.org>
-   **UK Biobank** — <https://www.ukbiobank.ac.uk>
-   **IEU OpenGWAS** — <https://gwas.mrcieu.ac.uk>
-   **GTEx** — <https://gtexportal.org>
-   **LDlink** — <https://ldlink.nci.nih.gov>
-   **gnomAD** — <https://gnomad.broadinstitute.org>

------------------------------------------------------------------------

## References {#references}

1.  Timpson NJ et al. **Genetic architecture: the shape of the genetic contribution to human traits and disease.** *Nat Rev Genet* (2017).
2.  Schaid DJ, Chen W, Larson NB. **From genome-wide associations to candidate causal variants by statistical fine-mapping.** *Nat Rev Genet* (2018).
3.  Minton K. **Linking disease-associated genetic variants to cell types and processes** (on Jagadeesh, Dey et al., *sc-linker*). *Nat Rev Genet* (2022).
4.  Reay WR, Cairns MJ. **Advancing the use of GWAS for drug repurposing.** *Nat Rev Genet* (2021).
5.  Denny JC et al. **Unravelling the human genome–phenome relationship using phenome-wide association studies.** *Nat Rev Genet* (2016).
6.  Harris L et al. **Genome-wide association testing beyond SNPs.** *Nat Rev Genet* (2025).
7.  Wojcik GL et al. **Genetic analyses of diverse populations improves discovery for complex traits.** *Nature* (2019).

------------------------------------------------------------------------

## [Questions?]{style="color:#ffffff"} {background-color="#5B4D80"}

📧 eha\@knu.ac.kr

Course materials → *\[your GitHub Pages URL\]*
