---
title: "W4. Polygenic Risk Scores"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Say what a PRS is, and what it changes for the patient in front of you
-   Recognise the four method names in a methods section without needing to implement them
-   ⚠️ Separate the **base** GWAS from the **target** cohort — and say why that distinction decides everything
-   Read a PRS performance claim: incremental R², ΔAUC, calibration, and the decile trick
-   ⚠️ Say why a European-derived PRS should not be applied to a Korean patient unchanged
-   Judge clinical readiness with ACCE rather than with AUC
-   Communicate a genetic risk result in a framing a patient can act on
:::

> 🤔 **Motivating question**
>
> A 42-year-old man with no family history brings a direct-to-consumer report: *"Coronary artery disease — 92nd percentile."* His LDL is 130 mg/dL. Do you start a statin?

------------------------------------------------------------------------

# 1. What a PRS Is 🧬 {#w4-s1}

## 1) One variant is useless. A million are not {#w4-s1-1}

From 🔗 [W3](w3_gwas.qmd): GWAS reports variants at *p* \< 5×10⁻⁸, each with an odds ratio of roughly **1.02–1.15**. No single one is worth acting on.

A PRS sums them — each GWAS effect size $\hat\beta_j$ multiplied by that person's allele dosage $G_{ij}$:

$$\text{PRS}_i = \sum_{j=1}^{M} \hat\beta_j \, G_{ij}$$

💡 Crucially, a PRS deliberately includes variants **below** genome-wide significance. Individually they are noise; in aggregate they carry real signal. This is the clinical pay-off of the omnigenic model (🔗 [W3 §1.3](w3_gwas.qmd#w3-s1-3)).

## 2) What it actually changes {#w4-s1-2}

A PRS does **not** tell you whether someone has a disease. It shifts a **pre-test probability** — the same thing a family history does, arrived at differently.

| It is | It is not |
|---------------------------------------------|--------------------------|
| A risk **indicator**, like LDL or a Framingham score | A **diagnosis** |
| Fixed from birth, measured once | A monitoring biomarker |
| **Modifiable in consequence** — statins work regardless of PRS | Deterministic |
| One input alongside age, sex, BMI, family history, labs | A replacement for any of them |

::: {.callout-tip collapse="true"}
## The only question that matters at the bedside

Does it **change management**?

If a high PRS moves the patient across a guideline threshold you were going to act on anyway — earlier statin, earlier mammography, earlier colonoscopy — it is useful.

If it does not cross a threshold, you have given the patient anxiety and yourself nothing.
:::

------------------------------------------------------------------------

# 2. How a PRS Gets Built 🏗️ {#w4-s2}

You will almost certainly never build one. But you will read papers that did, and be asked whether one applies to your patients.

## 1) ⚠️ Two datasets that must never have met {#w4-s2-1}

![Almost every inflated PRS result traces back to one of the two questions at the bottom.](/assets/img/w4/prs_construction.svg){fig-align="center"}

::: {.callout-important collapse="true"}
## The single most common way a PRS paper is wrong

If any of the **target** individuals also contributed to the **base** GWAS, the score has partly memorised their disease status. Performance is inflated, sometimes dramatically.

-   ⚠️ Sample overlap between base and target → **overfitting**
-   ⚠️ Tuning the p-value threshold on the same set you report performance on → **overfitting**

✅ A trustworthy paper states base and target explicitly, confirms independence, and evaluates in a **held-out** cohort.

💡 When you read "we optimised the threshold and achieved AUC 0.71", check on *what data* that optimisation happened. It is often one sentence in the supplement.
:::

## 2) The four names you will see {#w4-s2-2}

You need to recognise these, not run them.

| Method | Idea in one line | When you see it |
|-------------------|---------------------------|--------------------------|
| **C+T** (clumping + thresholding) | Keep one variant per LD block, drop those above a *p* cutoff | Older papers; still a sane baseline |
| **PRSice-2** | C+T with the threshold auto-tuned, plus standard plots | Very common in clinical papers |
| **LDpred2** | Bayesian — models LD and reweights *all* variants, no hard cutoff | Current default for accuracy |
| **PRS-CSx** | Continuous shrinkage across **multiple ancestries** at once | The one that matters for non-European cohorts |

::: {.callout-note collapse="true"}
## Does the method choice change your reading?

Mostly no. Differences between modern methods are modest compared with the differences that follow from ancestry and cohort.

Two exceptions worth flagging:

-   ⚠️ A **threshold-based** method (C+T, PRSice-2) has a tunable knob. Ask where it was tuned
-   ✅ A **multi-ancestry** method is the honest choice when the target is not European — its absence in a Korean-cohort paper is a fair criticism
:::

------------------------------------------------------------------------

# 3. ⚠️ Reading a PRS Paper 🔍 {#w4-s3}

This is the section to keep. Five numbers, in the order you should look for them.

## 1) Incremental value, not the PRS alone {#w4-s3-1}

A PRS **alone** reaching AUC 0.64 sounds respectable. Age and sex alone reach about that for many diseases.

![Same model, two questions. Most papers answer only the left one.](/assets/img/w4/discrimination_vs_calibration.svg){fig-align="center"}

The only meaningful comparison:

```         
Model A (clinical) : age + sex + BMI + family history + labs
Model B (clinical + PRS)

           ΔAUC  =  AUC(B) − AUC(A)
  incremental R²  =  R²(B)  − R²(A)      (Nagelkerke, for binary outcomes)
```

-   ❌ "PRS AUC = 0.64" — uninterpretable on its own
-   ✅ "Adding PRS raised AUC from 0.72 to 0.74 (*p* = 0.003)" — now you can judge it

💡 ΔAUC of 0.01–0.03 is typical for a good PRS. In a biobank of 400,000 that is comfortably significant, and for one patient it is often negligible. Both statements are true at once.

## 2) Calibration — the metric nobody shows {#w4-s3-2}

**Discrimination** asks: does the model rank people correctly? \
**Calibration** asks: are the predicted probabilities *right*?

Of the people told "10% risk over 10 years", do about 10% develop the disease?

::: {.callout-warning collapse="true"}
## Good discrimination with terrible calibration is easy

Look at the right-hand panel above. Both lines rank patients identically. One of them tells patients a number that is twice the truth.

-   ⚠️ A PRS calibrated in the UK Biobank is **not** calibrated in a Korean hospital population — different baseline incidence, age structure, and competing risks
-   ✅ Recalibration (updating the intercept or baseline hazard) is a routine fix — but it must be *done*, in a population-representative cohort
-   ⚠️ Biobank volunteers are healthier than the general population, so absolute risks derived from them run systematically low

→ If a paper reports AUC and no calibration plot, the absolute risks it quotes are not trustworthy.
:::

## 3) The framing decides what the patient hears {#w4-s3-3}

| Framing | Example | ⚠️ Caveat |
|-------------------|--------------------|---------------------------------|
| **Percentile** | "top 5% of PRS" | Says nothing about how much risk that is |
| **Relative risk / OR** | "2.3× the population average" | 2.3 × a 0.4% baseline is still 0.9% |
| **Absolute risk** | "18% over 10 years" | Needs incidence data for *your* population |

✅ **Absolute risk tied to a guideline action** is the only framing that is both honest and usable. Everything else needs translating before it reaches the patient.

## 4) ⚠️ Which deciles are being compared? {#w4-s3-4}

![Nothing here is fabricated. All three numbers describe the same ten bars.](/assets/img/w4/decile_trick.svg){fig-align="center"}

⚠️ A headline "**5-fold increased risk**" is almost always top-versus-bottom decile. Against the middle, the same score gives 2.8×. Against everyone else — which is how it would actually be used — 2.4×.

💡 Also check whether risk rises **smoothly** across deciles or only in the extreme tail. A score that separates only the top 1% is a screening tool for very few people.

## 5) Whose genomes built it? {#w4-s3-5}

Look for the ancestry of the **base** GWAS before anything else. §5 is entirely about why.

------------------------------------------------------------------------

# 4. Is It Ready for the Clinic? 🏥 {#w4-s4}

## 1) ACCE, not AUC {#w4-s4-1}

The standard framework for judging a genomic test:

-   **A**nalytical validity — does the pipeline compute the score accurately and reproducibly?
-   **C**linical validity — does the score associate with disease **in the target population**?
-   **C**linical utility — does *using* it improve outcomes?
-   **E**LSI — ethical, legal and social implications: equity, consent, insurance, communication

💡 Most published PRSs clear the first two. The bottleneck is **utility** and **ELSI** — which is why PRS is not yet routine care despite twenty years of GWAS.

## 2) What the trials actually show {#w4-s4-2}

| Trial | Setting | What it found / is testing |
|-----------------|-------------------|------------------------------------|
| **MI-GENES** | CHD | PRS-integrated risk disclosure → **lower LDL at 6 months**, via earlier and longer statin use |
| **GenoVA** | 6 diseases (CHD, T2D, AF, breast/colorectal/prostate cancer) | Effect of PRS results on time to diagnosis |
| **WISDOM / PERSPECTIVE** | Breast cancer | PRS-guided mammography intensity |
| **eMERGE**, **Our Future Health** | 10 conditions / 5 M participants | Implementation at scale, not efficacy |

::: {.callout-note collapse="true"}
## Read MI-GENES carefully

The mechanism of benefit was **behavioural** — patients told they were high-risk took statins earlier and stayed on them.

That is a real clinical effect. But note what it implies: much of the demonstrated value of PRS so far comes from **motivating adherence to treatments we already had**, not from identifying a biologically distinct group.

→ Worth remembering when a paper claims PRS "reveals a new high-risk phenotype".
:::

## 3) Where PRS is already inside a clinical equation {#w4-s4-3}

-   ✅ **Breast cancer — BOADICEA / CanRisk**: the one widely used equation combining PRS **with** family history, rare variants (BRCA1/2), and clinical factors. Lifetime risk of ER-positive disease ranges roughly **2% to 31%** between bottom and top centile
-   ✅ **CHD — Pooled Cohort Equations**: PRS is close to orthogonal to the clinical factors, so it can be added log-additively
-   ⚠️ **Everything else**: no validated absolute-risk equation. Absolute risk has to be reconstructed from population incidence, which is exactly where calibration problems enter

## 4) The human barriers {#w4-s4-4}

::: {.callout-warning collapse="true"}
## Three failure modes that have nothing to do with the mathematics

-   ⚠️ **Patients hear genetics as destiny.** Risk is probabilistic and modifiable — MI-GENES succeeded partly by framing it that way explicitly
-   ⚠️ **Run twenty PRSs and almost everyone is top-decile for something.** Alarm fatigue, and a strong argument against untargeted panels
-   ⚠️ **Most physicians have no genomics training.** Deployment needs decision support in the EHR and a trained champion in the department, not a PDF report
:::

------------------------------------------------------------------------

# 5. ⚠️ Portability: The Korean Problem 🌏 {#w4-s5}

## 1) PRS does not transfer across ancestries {#w4-s5-1}

![Nothing about the score changed. Only the patients did.](/assets/img/w4/portability.svg){fig-align="center"}

The reasons are structural, not fixable by collecting more of the same people:

-   **Allele frequencies differ** — a variant common in Europeans may be rare or absent
-   **LD patterns differ** — the tag SNP no longer tags the causal variant (🔗 [W3 §3.1](w3_gwas.qmd#w3-s3-1))
-   **Effect sizes differ** — gene–environment and genetic-background interactions

::: {.callout-important collapse="true"}
## This is a clinical safety issue, not only an equity issue

Deploying a miscalibrated PRS does not fail neutrally. It systematically under-flags the population it was not trained on.

→ Applying a European CAD PRS to Korean patients means the high-risk group you identify is enriched for people who resemble the training population — and the patients you miss are **not a random sample**.

⚠️ Widening a disparity with a tool marketed as precision medicine is worse than not using the tool.
:::

## 2) What to do instead {#w4-s5-2}

-   ✅ Prefer an **ancestry-matched base GWAS** — Korean or East Asian
-   ✅ Use **PRS-CSx** or multi-ancestry meta-analysis when a matched base is too small
-   ✅ **Recalibrate** in your own population and report ancestry-specific performance
-   ✅ Korean resources: **KoGES**, Korea Biobank Array (**KCHIP**), **National Biobank of Korea**; regionally **Biobank Japan**, **China Kadoorie**

💡 Diversity is not only fairness — different LD structures break haplotype blocks and produce **smaller credible sets** in fine-mapping (🔗 [W3 §6](w3_gwas.qmd#w3-s6)).

------------------------------------------------------------------------

# 6. What a PRS Cannot Do 🚫 {#w4-s6}

-   ❌ **Not causal** — a high PRS does not mean disease will occur
-   ❌ **Misses rare high-penetrance variants** — BRCA1/2, Lynch syndrome need dedicated testing. A reassuring PRS does not exclude them
-   ❌ **Misses non-SNP variation** — CNVs, structural variants, repeat expansions (🔗 [W3 §4](w3_gwas.qmd#w3-s4))
-   ❌ **Ignores gene × environment** — the same score means different things in a smoker and a non-smoker
-   ❌ **Not a diagnosis** — one input, not a verdict
-   ⚠️ **Not static in practice** — methods improve and reference panels grow. Keep raw genotypes so the score can be recomputed

::: {.callout-tip collapse="true"}
## Back to the motivating question

A 42-year-old man, CAD PRS 92nd percentile, LDL 130, no family history.

What you would want to know before answering:

-   ① **Which base GWAS?** If European and the patient is Korean → the percentile is not interpretable as stated
-   ② **92nd percentile of what absolute risk?** For a 42-year-old man, even 3× a low baseline may stay below the treatment threshold
-   ③ **Does it cross a guideline line?** 10-year risk moving from 4% to 7% is a real conversation. From 1.2% to 2.1% is not
-   ④ **What would change?** If the answer is "discuss statins and lifestyle" — that conversation was available without the test

→ The PRS is not useless here. It is **not sufficient on its own**, and the percentile is the least informative part of the report.
:::

------------------------------------------------------------------------

# 7. Hands-On 💻 {#w4-s7}

## 1) No code: read a real score in the PGS Catalog {#w4-s7-1}

The **PGS Catalog** ([pgscatalog.org](https://www.pgscatalog.org)) is the public registry of published scores — weights, metadata, and evaluated performance.

Pick a disease relevant to your specialty, then answer:

-   ① How many variants does the score use? (Published scores range from 6 to over a million)
-   ② What was the **ancestry** of the development cohort?
-   ③ Has it been **evaluated in an East Asian sample**? What happened to the effect size?
-   ④ Is calibration reported anywhere, or only discrimination?
-   ⑤ Are there several scores for the same disease? Do they agree on which individuals are high-risk?

💡 ⑤ is the underrated one. Two scores with near-identical population AUC routinely disagree about **which specific person** sits in the top decile. Some groups therefore require top-tier status across ≥2 scores before flagging anyone.

## 2) Minimal code: turn a score into a clinical figure {#w4-s7-2}

We provide a pre-computed score table (`prs_scores.tsv`) so the session is about **interpretation**, not compute. One row per person: `IID`, `PRS`, `disease`, `age`, `sex`.

```{r}
library(tidyverse)
library(here)

prs <- read_tsv(here("rawdata", "prs_scores.tsv"))

# Standardise so the score reads as SD units — always do this before reporting
prs <- prs |> mutate(PRS_z = as.numeric(scale(PRS)))
```

**Incremental value over a clinical model:**

```{r}
library(pROC)

m_clin <- glm(disease ~ age + sex,         data = prs, family = binomial)
m_prs  <- glm(disease ~ age + sex + PRS_z, data = prs, family = binomial)

roc_clin <- roc(prs$disease, predict(m_clin, type = "response"))
roc_prs  <- roc(prs$disease, predict(m_prs,  type = "response"))

roc.test(roc_clin, roc_prs)   # ← the number that belongs in the abstract
```

**Risk by decile — against an honest reference:**

```{r}
by_decile <- prs |>
  mutate(decile = ntile(PRS_z, 10)) |>
  group_by(decile) |>
  summarise(n = n(), cases = sum(disease), .groups = "drop") |>
  mutate(risk = cases / n,
         OR_vs_middle = (risk / (1 - risk)) /
                        (risk[decile == 5] / (1 - risk[decile == 5])))

ggplot(by_decile, aes(decile, OR_vs_middle)) +
  geom_col(fill = "#185FA5") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "PRS decile", y = "OR vs middle decile",
       title = "Risk by PRS decile — referenced to the middle, not the bottom") +
  theme_minimal()
```

✅ **Now change one thing:** re-run with `decile == 1` as the reference and compare the top bar. That difference is the one you will meet in abstracts.

::: {.callout-tip collapse="true"}
## If you want to build one yourself

Not required for this course, but the entry points are:

-   `bigsnpr` (R) — implements C+T and LDpred2; [privefl.github.io/bigsnpr](https://privefl.github.io/bigsnpr)
-   `PRSice-2` — command-line, auto-tunes the threshold; [prsice.info](https://www.prsice.info)
-   `PRS-CSx` — Python, multi-ancestry; [github.com/getian107/PRScsx](https://github.com/getian107/PRScsx)

⚠️ Each needs an LD reference panel matched to your target ancestry. That requirement, not the modelling, is what usually stops a clinical project.
:::

## 3) Before next week {#w4-s7-3}

**W5 — single-cell RNA-seq: technology, QC, clustering.** Install in advance:

```{r}
install.packages("Seurat")
BiocManager::install(c("SingleR", "celldex"))
renv::snapshot()
```

------------------------------------------------------------------------

# 📝 Summary

-   A PRS aggregates thousands to millions of small effects into one number — a **risk indicator**, never a diagnosis
-   ⚠️ **Base and target must be independent.** Tuning and evaluating on the same data inflates every reported number
-   Method names (C+T, PRSice-2, LDpred2, PRS-CSx) matter less than ancestry and cohort — recognise them, don't implement them
-   ❌ PRS AUC alone is meaningless. ✅ **ΔAUC and incremental R² over a clinical model** are the claim
-   ⚠️ **Calibration is usually missing**, and it is what makes an absolute risk trustworthy
-   ⚠️ "5-fold risk" is almost always top-vs-bottom decile. Ask for the comparison against the middle, or against everyone else
-   Percentile → relative risk → **absolute risk tied to a guideline action**: only the last is usable at the bedside
-   Clinical readiness is **ACCE**, not AUC. Utility evidence is thin, and much of it works through adherence
-   ⚠️ **A European-derived PRS is not valid for a Korean patient unchanged** — use matched or multi-ancestry scores, and recalibrate
-   The hard parts are portability, communication, and EHR integration. Not the mathematics

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|---------------------|---------------------------------------------------|
| PRS / PGS | Weighted sum of risk alleles across many variants |
| Base GWAS | The published study supplying the effect sizes |
| Target cohort | The individuals being scored |
| C+T | Clumping and *p*-value thresholding |
| LDpred2 | Bayesian method reweighting all variants using an LD reference |
| PRS-CSx | Continuous-shrinkage method spanning multiple ancestries |
| Incremental R² | Variance explained **beyond** a clinical model |
| ΔAUC | Gain in discrimination from adding the PRS |
| Calibration | Whether predicted probabilities match observed frequencies |
| Recalibration | Adjusting the baseline risk for a new population |
| Portability | How well a PRS transfers to another ancestry |
| ACCE | Analytical validity, Clinical validity, Clinical utility, ELSI |
| PGS Catalog | Public repository of published scores and their weights |
| BOADICEA / CanRisk | Breast cancer equation combining PRS, family history, rare variants |

------------------------------------------------------------------------

# ➕ References

### Clinical use and readiness

Kullo IJ, et al. [*Polygenic scores in biomedical research.*](https://doi.org/10.1038/s41576-022-00470-z) Nat Rev Genet. 2022.\
Khera AV, et al. [*Genome-wide polygenic scores for common diseases identify individuals with risk equivalent to monogenic mutations.*](https://doi.org/10.1038/s41588-018-0183-z) Nat Genet. 2018.\
Wand H, et al. [*Improving reporting standards for polygenic scores in risk prediction studies.*](https://doi.org/10.1038/s41586-021-03243-6) Nature. 2021. **← the PRS-RS checklist; use it as a reviewing aid**

### Methods, for recognition

Choi SW, Mak TS, O'Reilly PF. [*Tutorial: a guide to performing polygenic risk score analyses.*](https://doi.org/10.1038/s41596-020-0353-1) Nat Protoc. 2020.\
Privé F, Arbel J, Vilhjálmsson BJ. [*LDpred2: better, faster, stronger.*](https://doi.org/10.1093/bioinformatics/btaa1029) Bioinformatics. 2020.\
Ruan Y, et al. [*Improving polygenic prediction in ancestrally diverse populations.*](https://doi.org/10.1038/s41588-022-01054-7) Nat Genet. 2022. (PRS-CSx)

### Portability and diversity

Martin AR, et al. [*Clinical use of current polygenic risk scores may exacerbate health disparities.*](https://doi.org/10.1038/s41588-019-0379-x) Nat Genet. 2019.\
Wojcik GL, et al. [*Genetic analyses of diverse populations improves discovery for complex traits.*](https://doi.org/10.1038/s41586-019-1310-4) Nature. 2019.

### Implementation and communication

Lambert SA, et al. [*The Polygenic Score Catalog as an open database for reproducibility and systematic evaluation.*](https://doi.org/10.1038/s41588-021-00783-5) Nat Genet. 2021.\
Li R, Chen Y, Ritchie MD, Moore JH. [*Electronic health records and polygenic risk scores for predicting disease risk.*](https://doi.org/10.1038/s41576-020-0224-1) Nat Rev Genet. 2020.\
Lee A, et al. [*BOADICEA: a comprehensive breast cancer risk prediction model incorporating genetic and nongenetic risk factors.*](https://doi.org/10.1038/s41436-018-0406-9) Genet Med. 2019.

### Resources

[PGS Catalog](https://www.pgscatalog.org) · [CanRisk (BOADICEA)](https://www.canrisk.org) · [Korea Biobank](https://biobank.nih.go.kr) · [PRSice-2](https://www.prsice.info)
