---
title: "Genome-wide Association Studies (GWAS) & Post-GWAS"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 2"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W2 — GWAS & Post-GWAS"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Why GWAS? {background-color="#1D9E75"}

---

## GWAS Is Directly Clinical

::: {.incremental}
- **Drug target validation** — GWAS hits are 2× more likely to succeed in clinical trials
- **Disease subtyping** — same phenotype, different genetic architecture → different treatment
- **Risk stratification** — polygenic risk scores in preventive medicine (W3)
- **Comorbidity understanding** — shared genetic architecture between diseases
- **Mendelian randomization** — causal inference without RCT
:::

::: {.fragment}
> If you read a Nature Medicine paper in 2024, there's a good chance GWAS data is somewhere in it.
:::

---

## What GWAS Actually Asks

> **"Which genetic variants are more common in people with disease X than in healthy controls?"**

- Compare allele frequencies at **millions of SNPs** simultaneously
- No prior hypothesis about which gene — agnostic, genome-wide
- Result: list of associated loci, not causal genes (yet)

::: {.callout-note}
## SNP
Single Nucleotide Polymorphism — a single base difference at a given position in the genome.  
~10 million common SNPs exist in humans; GWAS arrays typically genotype 500K–5M.
:::

---

# Study Design & QC {background-color="#534AB7"}

---

## GWAS Study Design

```
Participants
├── Cases   (n = 5,000+)   patients with disease
└── Controls (n = 10,000+) matched healthy individuals
         ↓
Genotyping array (SNP chip)
         ↓
Quality Control
         ↓
Statistical testing (per-SNP logistic regression)
         ↓
Results → Manhattan plot
```

::: {.fragment}
**Sample size matters enormously.**  
Early GWAS (n ~2,000) found almost nothing.  
Modern biobank GWAS (n ~500,000) finds hundreds of loci.
:::

---

## Quality Control — What We Filter Out

| QC Step | Threshold | Why |
|---------|-----------|-----|
| Sample call rate | > 95% | poor DNA quality |
| SNP call rate | > 99% | unreliable genotyping |
| Hardy-Weinberg equilibrium | p > 1×10⁻⁶ | genotyping error |
| Minor allele frequency | > 1% | underpowered for rare variants |
| Heterozygosity outliers | ±3 SD | contamination or inbreeding |
| Relatedness (IBD) | π̂ < 0.125 | inflated statistics |
| **Population stratification** | PCA | **biggest confounder** |

---

## Population Stratification — The Key Confounder

:::: {.columns}

::: {.column width="55%"}
If cases and controls differ in **ancestry**, any ancestry-associated variant looks like a disease hit.

**Solution:** Principal Component Analysis (PCA) on genotype data → include top PCs as covariates in the model.
:::

::: {.column width="45%"}
```r
# Model with PC adjustment
glm(disease ~ SNP + PC1 + PC2 + 
    PC3 + PC4 + PC5 + age + sex,
    family = binomial)
```
:::

::::

::: {.callout-important}
## Always report your ancestry
Mixing populations without correction = false positives.  
Restricting to one ancestry = limits generalizability (a real ongoing problem in the field).
:::

---

# Reading GWAS Results {background-color="#185FA5"}

---

## The Manhattan Plot

```{r}
library(qqman)

# Basic Manhattan plot from summary stats
manhattan(gwas_results,
          chr = "CHR",
          bp  = "BP",
          p   = "P",
          snp = "SNP",
          col = c("#185FA5", "#1D9E75"),
          suggestiveline = -log10(1e-5),   # dashed blue
          genomewideline = -log10(5e-8),   # solid red
          highlight = lead_snps,
          main = "GWAS — Type 2 Diabetes (n = 150,000)")
```

---

## The Significance Threshold — Why 5×10⁻⁸?

Not arbitrary:

- ~1 million independent tests in the genome (after LD pruning)
- Bonferroni: 0.05 / 1,000,000 = **5×10⁻⁸**
- Genome-wide significance: `p < 5×10⁻⁸`
- Suggestive: `p < 1×10⁻⁵`

::: {.callout-tip}
## For reviewers
A "nearly significant" GWAS hit at p = 10⁻⁶ is **not significant**.  
Replication in an independent cohort is the gold standard.
:::

---

## The QQ Plot — Diagnosing Inflation

```{r}
# QQ plot
qq(gwas_results$P,
   main = "QQ Plot — observed vs expected -log10(p)")
```

- **Perfect**: points follow the diagonal, then deviate at the tail
- **Inflated** (λ > 1.05): all points above diagonal → stratification not corrected
- **Deflated** (λ < 0.95): overly conservative QC → lost power

```{r}
# Calculate genomic inflation factor (lambda)
chisq <- qchisq(gwas_results$P, df = 1, lower.tail = FALSE)
lambda <- median(chisq) / qchisq(0.5, df = 1)
cat("Genomic inflation factor λ =", round(lambda, 3))
```

---

# Post-GWAS: From Hit to Biology {background-color="#3B6D11"}

---

## The Problem with GWAS Hits

A GWAS hit is rarely the causal variant.

```
rs1234567  p = 3×10⁻²⁴   ← lead SNP (what GWAS finds)
    │
    ├── rs1234568  r² = 0.98   
    ├── rs1234591  r² = 0.95   ← one of these is actually causal
    └── rs1234612  r² = 0.91   
    
(all in high linkage disequilibrium — inherited together)
```

**Post-GWAS workflow:**
1. Fine-mapping — narrow down the credible set
2. Functional annotation — is it in a regulatory element?
3. Gene prioritization — which gene does it affect?
4. eQTL colocalization — does it regulate gene expression?

---

## Linkage Disequilibrium (LD)

```{r}
library(LDlinkR)

# Get LD between variants in a region
ld_matrix <- LDmatrix(
  snps  = c("rs1234567", "rs1234568", "rs1234591"),
  pop   = "CEU",           # 1000 Genomes population
  r2d   = "r2",
  token = "your_token"     # free at ldlink.nci.nih.gov
)

# LD proxy search
LDproxy(
  snp   = "rs1234567",
  pop   = "CEU",
  r2d   = "r2",
  token = "your_token"
)
```

---

## Fine-Mapping — Credible Sets

```{r}
library(susieR)

# SuSiE: Sum of Single Effects model
# Produces a credible set of variants likely containing the causal SNP
susie_fit <- susie(
  X = genotype_matrix,    # n_samples × n_snps
  y = phenotype_vector,
  L = 10                  # max number of causal signals
)

# 95% credible set
credible_set <- susie_get_cs(susie_fit, coverage = 0.95)
```

::: {.fragment}
Fine-mapping shrinks from **200 correlated SNPs** to a **credible set of 5–20** most likely causal.
:::

---

## Functional Annotation

Where is the lead SNP, and what might it do?

| Location | Functional implication |
|----------|----------------------|
| Coding (missense) | Direct protein change |
| Splice site | Altered mRNA |
| Promoter / 5'UTR | Transcription regulation |
| Enhancer | Distal gene regulation |
| **Intergenic** | **Unknown — needs eQTL** |

> ~90% of GWAS hits are non-coding. This is why epigenomics (W4–W5) is essential for interpretation.

---

## eQTL Colocalization — Connecting SNP to Gene

**eQTL**: expression QTL — a variant that affects gene expression level

```{r}
library(coloc)

# Test whether GWAS signal and eQTL signal share the same causal variant
coloc_result <- coloc.abf(
  dataset1 = list(pvalues = gwas_pvals,
                  N       = 150000,
                  type    = "cc",     # case-control
                  s       = 0.3),     # proportion cases
  dataset2 = list(pvalues = eqtl_pvals,
                  N       = 500,
                  type    = "quant")  # quantitative (expression)
)

coloc_result$summary
# PP.H4: probability that GWAS and eQTL share the same causal variant
```

::: {.callout-tip}
PP.H4 > 0.8 → strong evidence for colocalization → **that gene is likely the target.**
:::

---

## Mendelian Randomization — Causal Inference

> "Does elevated LDL *cause* coronary artery disease, or are they just correlated?"

```
Genetic variants → LDL levels → CAD
(instrument)       (exposure)    (outcome)
```

```{r}
library(TwoSampleMR)

# Instrument: GWAS SNPs for LDL
exposure_dat <- extract_instruments("ieu-a-300")  # LDL GWAS

# Outcome: CAD GWAS
outcome_dat  <- extract_outcome_data(
  snps       = exposure_dat$SNP,
  outcomes   = "ieu-a-7"       # CAD GWAS
)

# Harmonize and run MR
dat    <- harmonise_data(exposure_dat, outcome_dat)
result <- mr(dat)
```

::: {.fragment}
MR uses genetic variants as "natural randomization" — immune to confounding by lifestyle, SES, etc.
:::

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: T2D GWAS Summary Statistics

We'll use publicly available T2D GWAS summary stats (Mahajan et al. 2022, Nature Genetics)

```{r}
library(tidyverse)
library(qqman)
library(here)

# Load summary statistics
gwas <- read_tsv(here("data", "raw", "t2d_gwas_summary.tsv"))
glimpse(gwas)

# Expected columns:
# CHR, BP, SNP, A1, A2, FREQ, BETA, SE, P, N
```

---

## Step 1: Basic QC on Summary Stats

```{r}
gwas_clean <- gwas |>
  filter(
    !is.na(P),
    P > 0, P <= 1,
    MAF > 0.01,            # minor allele frequency filter
    INFO > 0.8             # imputation quality filter
  ) |>
  mutate(
    CHR = as.integer(CHR),
    neglogP = -log10(P)
  )

cat("SNPs before QC:", nrow(gwas), "\n")
cat("SNPs after QC: ", nrow(gwas_clean), "\n")
```

---

## Step 2: QQ Plot

```{r}
# Genomic inflation
chisq  <- qchisq(gwas_clean$P, df = 1, lower.tail = FALSE)
lambda <- median(chisq) / qchisq(0.5, df = 1)
cat("Lambda (GC) =", round(lambda, 4), "\n")

# QQ plot
qq(gwas_clean$P,
   main = paste0("QQ Plot — T2D GWAS  (λ = ", round(lambda, 3), ")"),
   col  = "#185FA5")
```

---

## Step 3: Manhattan Plot

```{r}
# Genome-wide significant hits
sig_snps <- gwas_clean |>
  filter(P < 5e-8) |>
  arrange(P)

cat("Genome-wide significant loci:", nrow(sig_snps), "\n")

# Plot
manhattan(gwas_clean,
          chr  = "CHR",
          bp   = "BP",
          p    = "P",
          snp  = "SNP",
          col  = c("#185FA5", "#1D9E75"),
          suggestiveline = -log10(1e-5),
          genomewideline = -log10(5e-8),
          highlight      = sig_snps$SNP[1:10],
          main           = "T2D GWAS — Manhattan Plot")
```

---

## Step 4: Annotate Top Hits

```{r}
library(biomaRt)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Get genes near top hits
top_hits <- sig_snps |> slice_min(P, n = 20)

annotations <- getBM(
  attributes = c("chromosome_name", "start_position",
                 "end_position", "external_gene_name"),
  filters    = "chromosomal_region",
  values     = paste0(top_hits$CHR, ":", 
                      top_hits$BP - 500000, ":",
                      top_hits$BP + 500000),
  mart       = mart
)

head(annotations)
```

---

## Step 5: Regional Association Plot (Locus Zoom)

```{r}
library(gassocplot)

# Focus on TCF7L2 locus — most replicated T2D locus
tcf7l2_region <- gwas_clean |>
  filter(CHR == 10,
         BP  >= 114700000,
         BP  <= 115000000)

stack_assoc_plot(
  markers = tcf7l2_region,
  z       = tcf7l2_region$neglogP,
  corr    = ld_matrix_tcf7l2,     # from LDlinkR
  traits  = "T2D",
  ylab    = "-log10(p)"
)
```

---

## What to Interpret from a Locus

For any GWAS hit, ask:

::: {.incremental}
1. **Is it replicated?** — in an independent cohort?
2. **What's the LD structure?** — how many SNPs in the credible set?
3. **What's the functional context?** — coding, regulatory, eQTL?
4. **Which gene?** — nearest gene ≠ causal gene
5. **What's the effect size?** — OR / beta; GWAS hits are usually small
6. **Any drug target?** — is the gene druggable?
:::

---

# Limitations & Caveats {background-color="#993C1D"}

---

## What GWAS Cannot Tell You

::: {.incremental}
- **Causality** — association ≠ causation (MR helps, but has assumptions)
- **Rare variants** — standard GWAS misses MAF < 1% (need WES/WGS)
- **Mechanism** — a hit at an enhancer tells you *something* regulates *something*
- **Diverse populations** — most GWAS are European-ancestry; results may not transfer
- **Gene-environment interaction** — static genetic associations miss dynamic biology
:::

::: {.fragment}
> **GWAS is the beginning of the story, not the end.**  
> This is exactly why the rest of the semester exists.
:::

---

## GWAS + The Rest of the Curriculum

```
GWAS hit identified
    │
    ├─ W4–W5  Epigenomics    → Is the hit in an active enhancer?
    ├─ W6–W7  Metagenomics   → Does the microbiome mediate the effect?
    ├─ W8–W9  scRNA-seq      → Which cell type expresses the gene?
    ├─ W10    Spatial        → Where in the tissue?
    ├─ W11    Multi-omics    → Convergent evidence across layers
    └─ W12    Drug target    → Is there a druggable molecule?
```

GWAS is the **entry point** — everything else gives it biological meaning.

---

# Next Week {background-color="#534AB7"}

---

## W3 Preview — Polygenic Risk Scores (PRS)

**What we'll cover:**

- From single GWAS hits → genome-wide prediction
- PRS construction methods (C+T, LDpred2, PRSice2)
- Clinical utility — screening, risk stratification
- Hands-on: building a PRS from summary stats

**Install before next class:**

```{r}
install.packages("bigsnpr")
install.packages("bigstatsr")

# PRSice2 is a command-line tool — download binary:
# https://www.prsice.info/
```

---

## Key Resources

- **GWAS Catalog** — [www.ebi.ac.uk/gwas](https://www.ebi.ac.uk/gwas) — all published GWAS hits
- **UK Biobank** — [www.ukbiobank.ac.uk](https://www.ukbiobank.ac.uk) — 500K participants
- **IEU Open GWAS** — [gwas.mrcieu.ac.uk](https://gwas.mrcieu.ac.uk) — MR-ready summary stats
- **GTEx** — [gtexportal.org](https://gtexportal.org) — eQTL reference
- **LDlink** — [ldlink.nci.nih.gov](https://ldlink.nci.nih.gov) — LD calculations (free token)

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Polygenic Risk Scores (PRS)**

Course materials → [your GitHub Pages URL here]
:::
