---
title: "Polygenic Risk Scores (PRS)"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 3"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W3 — Polygenic Risk Scores"
    css: custom.css
execute:
  echo: true
  eval: false
---

# From GWAS Hits to Clinical Risk {background-color="#1D9E75"}

---

## The Limitation of Single GWAS Hits

From last week:

- GWAS finds variants with **p < 5×10⁻⁸**
- But most complex diseases are influenced by **thousands of variants**
- Each with a **tiny effect size** (OR 1.02–1.15)
- No single variant is clinically useful alone

::: {.fragment}
> **PRS aggregates the effects of thousands of SNPs into a single score per individual.**
:::

---

## What a PRS Looks Like

```
Individual A:
  rs1001  risk allele × 0.03
+ rs2045  risk allele × 0.02
+ rs8821  non-risk    × 0.00
+ ... (thousands more)
─────────────────────────────
= PRS: 1.42  (top 20% of population → 3× increased risk)
```

::: {.fragment}
**Clinically:**  
One number, derived from a saliva sample at birth,  
that captures lifetime genetic predisposition to a disease.
:::

---

## Clinical Utility — Where PRS Is Already Used

::: {.incremental}
- **Coronary artery disease** — PRS identifies high-risk individuals who benefit from early statin therapy
- **Breast cancer** — PRS improves risk stratification beyond BRCA1/2
- **Type 2 diabetes** — PRS + lifestyle factors → personalized prevention
- **Atrial fibrillation** — PRS predicts incident AF beyond clinical risk scores
- **Prostate cancer** — PRS guides screening intensity
:::

::: {.fragment}
::: {.callout-tip}
## The key point for clinicians
PRS adds information **on top of** traditional risk factors — age, BMI, family history, biomarkers.  
The question is always: does it change clinical management?
:::
:::

---

# PRS Methods {background-color="#534AB7"}

---

## Method 1: C+T (Clumping and Thresholding)

The simplest approach:

```
1. Clumping: keep only one SNP per LD block
   (remove correlated SNPs → independent signals only)

2. Thresholding: keep SNPs below a p-value cutoff
   (e.g., p < 0.05, p < 0.01, p < 5×10⁻⁸)

3. PRS = Σ (effect size × allele count)
```

```{r}
library(bigsnpr)

# C+T with PRSice2 (via system call) or bigsnpr
obj.bigsnp <- snp_attach(here("data", "raw", "target.rds"))
G <- obj.bigsnp$genotypes

# Clumping
ind.keep <- snp_clumping(
  G,
  infos.chr = obj.bigsnp$map$chromosome,
  infos.pos = obj.bigsnp$map$physical.pos,
  S         = abs(sumstats$Z),     # use Z-score for clumping
  thr.r2    = 0.1,
  size      = 500
)
```

---

## Method 2: LDpred2 — The Current Standard

Accounts for LD structure more accurately using a Bayesian framework:

```{r}
library(bigsnpr)

# Step 1: Match summary stats to target genotypes
sumstats <- snp_match(sumstats, obj.bigsnp$map)

# Step 2: Compute LD reference matrix
POS2 <- snp_asGeneticPos(
  infos.chr = obj.bigsnp$map$chromosome,
  infos.pos = obj.bigsnp$map$physical.pos,
  dir       = here("data", "raw", "genetic_maps")
)

corr <- snp_cor(G, ind.col = sumstats$`_NUM_ID_`,
                infos.pos = POS2, size = 3 / 1000)

# Step 3: LDpred2-auto (no tuning needed)
ldsc <- with(sumstats,
             snp_ldsc(ld_score, ld_size = nrow(G),
                      chi2  = (beta / beta_se)^2,
                      sample_size = n_eff))

h2_est <- ldsc[["h2"]]

multi_auto <- snp_ldpred2_auto(
  corr, sumstats, h2_init = h2_est,
  vec_p_init = seq_log(1e-4, 0.2, length.out = 30),
  burn_in = 500, num_iter = 500, report_step = 20
)
```

---

## Method 3: PRSice2 — Practical Workhorse

```{r}
# PRSice2 is a command-line tool
# Call from R via system()
system(paste(
  "PRSice",
  "--base",    here("data", "raw", "sumstats.txt"),
  "--target",  here("data", "raw", "target_plink"),
  "--binary-target", "T",
  "--pheno",   here("data", "raw", "phenotype.txt"),
  "--cov",     here("data", "raw", "covariates.txt"),
  "--out",     here("results", "prsice_output"),
  "--thread",  4
))
```

::: {.callout-note}
PRSice2 automatically optimizes the p-value threshold and produces  
a bar chart, high-resolution plot, and quantile plot.
:::

---

## Comparing Methods

| Method | Pros | Cons |
|--------|------|------|
| **C+T** | Fast, interpretable | Loses information, arbitrary threshold |
| **LDpred2** | Best predictive accuracy, principled | Slower, needs LD reference |
| **PRSice2** | Easy to use, good visualization | Threshold-based |
| **PRS-CS** | Handles diverse ancestries better | Needs Python |

::: {.fragment}
**For clinical research:** LDpred2 or PRSice2.  
**For multi-ancestry:** PRS-CS or PRS-CSx.
:::

---

# Evaluating PRS Performance {background-color="#185FA5"}

---

## Key Performance Metrics

```{r}
library(tidyverse)

# Load PRS scores + phenotype
prs_data <- read_tsv(here("results", "prsice_output.best")) |>
  left_join(phenotype, by = "IID")

# Nagelkerke R² — variance explained by PRS
library(DescTools)

# Null model (covariates only)
null_model <- glm(disease ~ age + sex + PC1 + PC2 + PC3 + PC4 + PC5,
                  data   = prs_data,
                  family = binomial)

# Full model (covariates + PRS)
full_model <- glm(disease ~ PRS + age + sex + PC1 + PC2 + PC3 + PC4 + PC5,
                  data   = prs_data,
                  family = binomial)

r2_prs <- PseudoR2(full_model, "Nagelkerke") -
          PseudoR2(null_model, "Nagelkerke")
cat("R² explained by PRS:", round(r2_prs, 4), "\n")
```

---

## AUC and Discrimination

```{r}
library(pROC)

# ROC curve — does PRS improve discrimination?
roc_null <- roc(prs_data$disease,
                predict(null_model, type = "response"))

roc_full <- roc(prs_data$disease,
                predict(full_model, type = "response"))

# Compare AUCs
roc.test(roc_null, roc_full)

# Plot
ggroc(list(Clinical = roc_null,
           `Clinical + PRS` = roc_full)) +
  labs(title = "ROC: Does PRS improve prediction?",
       color  = "Model") +
  theme_minimal()
```

---

## Quantile Plot — Clinical Interpretation

```{r}
# Divide PRS into deciles
prs_data <- prs_data |>
  mutate(prs_decile = ntile(PRS, 10))

# Odds ratio per decile vs. middle decile (reference)
or_by_decile <- prs_data |>
  group_by(prs_decile) |>
  summarise(
    n_cases    = sum(disease == 1),
    n_controls = sum(disease == 0)
  ) |>
  mutate(
    OR = (n_cases / n_controls) /
         (n_cases[prs_decile == 5] / n_controls[prs_decile == 5])
  )

ggplot(or_by_decile, aes(prs_decile, OR)) +
  geom_col(fill = "#185FA5") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(x = "PRS decile",
       y = "Odds Ratio vs. middle decile",
       title = "Disease risk by PRS decile") +
  theme_minimal()
```

---

# Clinical Considerations {background-color="#0F6E56"}

---

## PRS in Diverse Populations — A Real Problem

::: {.incremental}
- Most GWAS (and therefore PRS) are trained on **European-ancestry individuals**
- PRS built on European GWAS **performs 4–5× worse** in African-ancestry populations
- Applying biased PRS clinically could **widen health disparities**
- Active research: PRS-CSx, multi-ancestry GWAS meta-analysis
:::

::: {.fragment}
::: {.callout-important}
## For Korean patients
Korean-specific GWAS (KoGES, KCHIP, TOPMed Korean) exist for major diseases.  
Always check whether the base GWAS matches your target population's ancestry.
:::
:::

---

## What PRS Cannot Do

::: {.incremental}
- **Causality** — high PRS ≠ will develop disease
- **Rare variants** — PRS misses high-penetrance rare variants (BRCA1/2, Lynch syndrome)
- **Gene–environment interaction** — genetic risk can be modified by lifestyle
- **Replace clinical judgment** — PRS is one input, not a diagnosis
:::

---

## PRS + Clinical Risk Scores

```{r}
# Example: CAD risk — PRS vs. traditional risk factors
library(tidyverse)

# Framingham Risk Score + PRS
combined_model <- glm(
  cad ~ PRS + age + sex + total_chol + hdl_chol +
        systolic_bp + smoking + diabetes,
  data   = prs_data,
  family = binomial
)

# Net reclassification index
library(PredictABEL)
reclassification(
  data         = prs_data,
  cOutcome     = "cad",
  predrisk1    = predict(null_model, type = "response"),
  predrisk2    = predict(combined_model, type = "response"),
  cutoff       = c(0, 0.1, 0.2, 1)
)
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: CAD PRS from Public Summary Stats

Using Khera et al. 2018 (Nature Genetics) CAD GWAS summary statistics  
+ 1000 Genomes Project target genotypes for demonstration

```{r}
library(bigsnpr)
library(tidyverse)
library(here)

# Load pre-processed bigSNP object
obj.bigsnp <- snp_attach(here("data", "raw", "1kg_eur.rds"))
G   <- obj.bigsnp$genotypes
map <- obj.bigsnp$map

# Load CAD summary statistics
sumstats <- read_tsv(here("data", "raw", "cad_sumstats.tsv"))
glimpse(sumstats)
```

---

## Hands-On: Match & Compute PRS

```{r}
# Match summary stats to target SNPs
sumstats <- snp_match(sumstats, map)

# Compute simple C+T PRS
beta <- sumstats$beta
ind  <- match(sumstats$`_NUM_ID_`, cols_along(G))

prs_scores <- big_prodVec(G, beta[!is.na(ind)],
                           ind.col = ind[!is.na(ind)])

# Add to metadata
prs_df <- obj.bigsnp$fam |>
  mutate(PRS = prs_scores)

summary(prs_df$PRS)
```

---

## Hands-On: Visualize Distribution

```{r}
# PRS distribution by phenotype
ggplot(prs_df, aes(PRS, fill = factor(affection))) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("#1D9E75", "#D85A30"),
                    labels  = c("Control", "Case")) +
  labs(title = "PRS distribution — CAD cases vs. controls",
       x     = "Polygenic Risk Score",
       fill  = "Status") +
  theme_minimal()

# Odds ratio by PRS percentile
prs_df <- prs_df |>
  mutate(prs_pct = ntile(PRS, 100),
         top10   = prs_pct >= 90,
         bottom10 = prs_pct <= 10)

# Risk in top 10% vs. bottom 10%
table(prs_df$top10, prs_df$affection)
```

---

## Hands-On: LDpred2-auto

```{r}
# Compute LD matrix (pre-computed for speed)
corr <- readRDS(here("data", "processed", "ld_matrix.rds"))

# Run LDpred2-auto
ldsc    <- with(sumstats,
                snp_ldsc(ld_score, ld_size = nrow(G),
                         chi2  = (beta / beta_se)^2,
                         sample_size = n_eff))
h2_est  <- ldsc[["h2"]]

multi_auto <- snp_ldpred2_auto(
  corr, sumstats, h2_init = h2_est,
  vec_p_init = seq_log(1e-4, 0.2, length.out = 30),
  burn_in    = 200, num_iter  = 200
)

# Extract best beta weights
beta_auto <- sapply(multi_auto, function(a) a$beta_est)
prs_ldpred2 <- big_prodMat(G, beta_auto)
```

---

## Save & Document

```{r}
# Save PRS scores
prs_final <- prs_df |>
  select(FID, IID, PRS_CT = PRS) |>
  mutate(PRS_LDpred2 = rowMeans(prs_ldpred2))

write_tsv(prs_final,
          here("results", "tables", "prs_scores.tsv"))

# Session info for reproducibility
renv::snapshot()
sessionInfo()
```

---

# Next Week {background-color="#534AB7"}

---

## W4 Preview — Epigenomics I: DNA Methylation

**What we'll cover:**

- CpG methylation — what it is and what it regulates
- Methylation arrays (EPIC, 450K) vs. WGBS
- Differential methylation analysis with `minfi` and `ChAMP`
- Epigenetic clocks — biological age from methylation data

**Install before next class:**

```{r}
BiocManager::install(c("minfi", "ChAMP",
                       "IlluminaHumanMethylationEPICanno.ilm10b4.hg19",
                       "missMethyl"))
```

---

## Key Resources

- **PGS Catalog** — [pgscatalog.org](https://www.pgscatalog.org) — published PRS for hundreds of traits
- **PRSice2** — [prsice.info](https://www.prsice.info)
- **bigsnpr tutorial** — [privefl.github.io/bigsnpr](https://privefl.github.io/bigsnpr)
- **Khera et al. 2018** — landmark PRS paper for CAD, T2D, AF, IBD, breast cancer

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Epigenomics I — DNA Methylation**

Course materials → [your GitHub Pages URL here]
:::
