---
title: "Epigenomics I: DNA Methylation"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 4"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W4 — DNA Methylation"
    css: custom.css
execute:
  echo: true
  eval: false
---

# What Is Epigenomics? {background-color="#1D9E75"}

---

## Beyond the DNA Sequence

Same genome. Different cell types. Different functions.

```
Neuron  ─── DNA ──→  expresses NEFL, SNAP25, SYP
Hepatocyte ─ DNA ──→  expresses ALB, CYP3A4, F2
T cell  ─── DNA ──→  expresses CD3E, CD8A, GZMB
```

::: {.fragment}
**The genome is the same. The epigenome is different.**

Epigenomics studies the regulatory layer **on top of** the DNA sequence:
- DNA methylation
- Histone modifications
- Chromatin accessibility (W5)
- 3D genome organization (W5)
:::

---

## Why Epigenomics for Clinicians?

::: {.incremental}
- **Cancer** — global hypomethylation + gene-specific hypermethylation → tumor suppressor silencing
- **Aging** — epigenetic clocks predict biological age, mortality, disease risk
- **Drug response** — methylation status of promoters predicts chemotherapy sensitivity
- **Developmental disorders** — imprinting diseases (Prader-Willi, Angelman)
- **Liquid biopsy** — cell-free DNA methylation → tissue of origin for cancer detection
:::

---

# DNA Methylation {background-color="#534AB7"}

---

## The Biology

```
CpG dinucleotide:  5'─C─G─3'
                   3'─G─C─5'

Methylation:   5'─[m]C─G─3'   (5-methylcytosine)

Maintained by: DNMT1 (maintenance methyltransferase)
Written by:    DNMT3A / DNMT3B (de novo)
Erased by:     TET enzymes (oxidation → demethylation)
```

::: {.fragment}
**Effect on gene expression:**
- Promoter methylation → gene **silencing**
- Gene body methylation → associated with **active transcription** (context-dependent)
- Enhancer methylation → **reduced enhancer activity**
:::

---

## CpG Islands, Shores, and Shelves

```
                shore    CGI      shore
──────────────[====][===========][====]──────────────
              shelf                     shelf

CpG Island (CGI): high CpG density, usually unmethylated in normal tissue
Shore (0–2 kb):   variable methylation, most disease-relevant
Shelf (2–4 kb):   lower CpG density
Open sea:         remaining genome
```

::: {.callout-tip}
## Clinically important
Most cancer-associated methylation changes happen at **CGI shores**, not the islands themselves.
:::

---

## Measurement Technologies

| Technology | Resolution | Coverage | Throughput | Cost |
|-----------|-----------|---------|-----------|------|
| **Illumina EPIC array** | CpG-level | 935K sites | High | Low |
| **450K array** | CpG-level | 485K sites | High | Low |
| **WGBS** | CpG-level | All ~28M CpGs | Low | High |
| **RRBS** | CpG-level | CpG-rich regions | Medium | Medium |
| **MeDIP-seq** | Regional | Genome-wide | Medium | Medium |

::: {.fragment}
**For most clinical research:** EPIC array — cost-effective, validated, huge reference datasets  
(GEO has thousands of EPIC datasets available now)
:::

---

# Analysis Workflow {background-color="#185FA5"}

---

## Overview

```
Raw IDAT files (Illumina array)
    │
    ├── 1. Load & QC          → minfi / ChAMP
    ├── 2. Normalization       → BMIQ / Noob
    ├── 3. Filtering           → remove SNP-associated, cross-reactive probes
    ├── 4. DMP analysis        → differentially methylated positions
    ├── 5. DMR analysis        → differentially methylated regions
    ├── 6. Functional annotation → gene, CGI context
    └── 7. Epigenetic clock    → biological age estimation
```

---

## Step 1: Load IDAT Files

```{r}
library(minfi)
library(here)

# Read sample sheet
targets <- read.metharray.sheet(here("data", "raw", "methylation"),
                                 pattern = "SampleSheet.csv")

# Load raw IDAT files
rgSet <- read.metharray.exp(targets = targets,
                             verbose  = TRUE)
rgSet

# Check: should see ~935,000 probes (EPIC) or ~485,000 (450K)
```

---

## Step 2: QC

```{r}
# QC report
qcReport(rgSet,
         sampNames = targets$Sample_Name,
         sampGroups = targets$Sample_Group,
         pdf       = here("results", "qc_report.pdf"))

# Detection p-values — probes with p > 0.01 are unreliable
detP <- detectionP(rgSet)
failed <- detP > 0.01
cat("Failed probes per sample:\n")
colSums(failed)

# Remove samples with > 1% failed probes
keep <- colMeans(failed) < 0.01
rgSet <- rgSet[, keep]
cat("Samples remaining:", ncol(rgSet), "\n")
```

---

## Step 3: Normalization

```{r}
# Noob + BMIQ normalization — recommended for EPIC
mSet <- preprocessNoob(rgSet)
mSet <- BMIQ(mSet)   # within-sample normalization

# Extract beta values (0–1, proportion methylated)
beta <- getBeta(mSet)

# Extract M-values (logit of beta, better for statistics)
mval <- getM(mSet)

# Beta values for visualization, M-values for statistical testing
cat("Beta matrix:", nrow(beta), "probes ×", ncol(beta), "samples\n")
```

---

## Step 4: Probe Filtering

```{r}
library(maxprobes)   # cross-reactive probes list

# Remove probes that:
# 1. Failed QC in any sample
keep_probes <- rowSums(detP > 0.01) == 0

# 2. Are on sex chromosomes (if not studying sex differences)
annot <- getAnnotation(mSet)
keep_probes <- keep_probes & !(annot$chr %in% c("chrX", "chrY"))

# 3. Contain known SNPs at CpG site
keep_probes <- keep_probes & is.na(annot$Probe_maf)

# 4. Are known cross-reactive
xreact <- xreactive_probes(array_type = "EPIC")
keep_probes <- keep_probes & !(rownames(mSet) %in% xreact)

mSet_filtered <- mSet[keep_probes, ]
cat("Probes after filtering:", nrow(mSet_filtered), "\n")
```

---

## Step 5: Differential Methylation — DMPs

```{r}
library(limma)

# Design matrix — adjust for age, sex, cell type composition
design <- model.matrix(
  ~ Group + age + sex + CD8T + CD4T + NK + Bcell + Mono + Gran,
  data = targets
)

# Fit linear model on M-values
fit <- lmFit(mval[keep_probes, ], design)
fit <- eBayes(fit)

# Extract DMPs
dmps <- topTable(fit,
                 coef    = "GroupCase",
                 number  = Inf,
                 sort.by = "p") |>
  as_tibble(rownames = "probe")

# Significant DMPs
sig_dmps <- dmps |>
  filter(adj.P.Val < 0.05, abs(logFC) > 0.1)

cat("Significant DMPs:", nrow(sig_dmps), "\n")
```

---

## Step 6: Differential Methylation — DMRs

```{r}
library(DMRcate)

# DMR analysis — finds regions of co-methylated probes
myannotation <- cpg.annotate(
  object       = mval,
  datatype     = "array",
  what         = "M",
  analysis.type = "differential",
  design       = design,
  coef         = "GroupCase",
  arraytype    = "EPIC"
)

# Find DMRs
dmrcoutput <- dmrcate(myannotation,
                      lambda   = 1000,
                      C        = 2)

dmrs <- extractRanges(dmrcoutput, genome = "hg19")
dmrs
```

---

## Step 7: Visualization

```{r}
library(ggplot2)
library(patchwork)

# Volcano plot of DMPs
ggplot(dmps, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = abs(logFC) > 0.1 & adj.P.Val < 0.05),
             alpha = 0.5, size = 0.8) +
  scale_color_manual(values = c("gray70", "#D85A30")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed") +
  labs(x     = "Δβ (M-value logFC)",
       y     = "-log10(adj. p-value)",
       title = "Differentially Methylated Positions",
       color = "Significant") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Cell Type Deconvolution — Critical Confounder

```{r}
library(minfi)

# Whole blood methylation is a mixture of cell types
# Must estimate and correct for cell type composition

cell_counts <- estimateCellCounts2(
  rgSet,
  referencePlatform = "IlluminaHumanMethylationEPIC",
  returnAll         = FALSE
)

# Cell type proportions
head(cell_counts)
# CD8T  CD4T  NK  Bcell  Mono  Gran

# Visualize
cell_counts |>
  as_tibble(rownames = "sample") |>
  pivot_longer(-sample, names_to = "cell_type", values_to = "proportion") |>
  ggplot(aes(sample, proportion, fill = cell_type)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Estimated blood cell type composition")
```

---

# Epigenetic Clocks {background-color="#0F6E56"}

---

## Biological Age vs. Chronological Age

::: {.incremental}
- **Horvath clock (2013)** — multi-tissue, 353 CpGs → predicts chronological age
- **Hannum clock (2013)** — blood-specific, 71 CpGs
- **PhenoAge (2018)** — predicts mortality, not just age
- **GrimAge (2019)** — predicts time-to-death and disease onset
- **DunedinPACE (2022)** — pace of aging, not biological age point estimate
:::

::: {.fragment}
**Epigenetic age acceleration** = (biological age) − (chronological age)

Positive acceleration → aging faster → higher mortality, disease risk
:::

---

## Computing Epigenetic Age

```{r}
library(methylclock)

# Compute multiple epigenetic clocks at once
clocks <- DNAmAge(
  x        = beta,           # beta value matrix (probes × samples)
  clocks   = c("Horvath",
               "Hannum",
               "PhenoAge",
               "GrimAge"),
  cell.count = TRUE          # adjust for cell composition
)

# Age acceleration
clocks <- clocks |>
  mutate(
    AgeAccel_Horvath  = Horvath - age,
    AgeAccel_Hannum   = Hannum  - age,
    AgeAccel_PhenoAge = PhenoAge - age
  )

# Do cases age faster?
t.test(AgeAccel_GrimAge ~ Group, data = clocks)
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: GEO — Colorectal Cancer Methylation

GSE48684 — Colorectal cancer vs. normal tissue (450K array)

```{r}
library(GEOquery)
library(minfi)
library(here)

# Download from GEO (pre-processed beta values)
gse <- getGEO("GSE48684", GSEMatrix = TRUE)
beta_geo <- exprs(gse[[1]])
pheno    <- pData(gse[[1]])

cat("Probes:", nrow(beta_geo), "\n")
cat("Samples:", ncol(beta_geo), "\n")
table(pheno$`tissue type:ch1`)
```

---

## Hands-On: DMP Analysis

```{r}
library(limma)

# M-values
mval_geo <- logit2(beta_geo)
mval_geo[!is.finite(mval_geo)] <- NA

# Design
group <- factor(pheno$`tissue type:ch1`)
design <- model.matrix(~ group)

# Fit
fit    <- lmFit(mval_geo, design)
fit    <- eBayes(fit)
dmps   <- topTable(fit, coef = 2, number = Inf)

# Top hypermethylated in cancer
dmps |>
  filter(logFC > 0, adj.P.Val < 0.01) |>
  arrange(desc(logFC)) |>
  head(20)
```

---

## Hands-On: Pathway Enrichment on DMPs

```{r}
library(missMethyl)

# Gene set testing for methylation data
# (accounts for variable number of probes per gene)
gst <- gometh(
  sig.cpg   = rownames(sig_dmps),
  all.cpg   = rownames(mval_geo),
  plot.bias = TRUE,
  collection = "KEGG"
)

topGSA(gst, number = 20)
```

---

# Next Week {background-color="#534AB7"}

---

## W5 Preview — Epigenomics II: Chromatin Accessibility & 3D Genome

**What we'll cover:**

- ATAC-seq — open chromatin profiling
- Peak calling, differential accessibility
- Hi-C / 3D genome — TADs, compartments, loops
- Connecting GWAS hits to regulatory elements

**Install before next class:**

```{r}
BiocManager::install(c("ArchR", "chromVAR",
                       "BSgenome.Hsapiens.UCSC.hg38",
                       "TFBSTools"))
install.packages("strawr")   # for Hi-C .hic files
```

---

## Key Resources

- **GEO** — [ncbi.nlm.nih.gov/geo](https://www.ncbi.nlm.nih.gov/geo) — thousands of methylation datasets
- **ENCODE** — [encodeproject.org](https://www.encodeproject.org) — reference epigenomes
- **minfi vignette** — [bioconductor.org/packages/minfi](https://bioconductor.org/packages/minfi)
- **methylclock** — epigenetic clocks in R
- **Illumina EPIC annotation** — pre-built annotation for all probes

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Epigenomics II — Chromatin Accessibility & 3D Genome**

Course materials → [your GitHub Pages URL here]
:::
