---
title: "W10. Clinical Proteomics & Biomarkers"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   ⚠️ Explain why missing values in proteomics are not missing at random, and how the choice of imputation changes conclusions
-   Run a differential abundance analysis with `limma` and say why it is not just a t-test
-   ⚠️ Place a biomarker claim on the discovery → verification → validation → clinical ladder
-   ⚠️ Recognise the wrong-control-group error, and design around it
-   Say what an AUC of 0.94 in a discovery study is worth
-   Distinguish diagnostic, prognostic and predictive biomarkers — and say which one a study actually tested
-   Write the design for a biomarker study you could run in your own department
:::

> 🤔 **Motivating question**
>
> A paper reports a plasma protein separating pancreatic cancer from healthy controls with AUC = 0.94. Would you order it for the patient in front of you with vague abdominal pain?

------------------------------------------------------------------------

# 1. ⚠️ Missing Values Come First 🕳️ {#w10-s1}

Before any statistics, proteomics forces a decision that transcriptomics does not.

![A zero in RNA-seq and a blank in proteomics mean very different things.](/assets/img/w10/missing_values.svg){fig-align="center"}

## 1) Why they are not random {#w10-s1-1}

A protein is missing from a run mostly because it was **below the limit of detection** — that is, because it was low. Statisticians call this **MNAR** (missing not at random), and it is the opposite of the assumption most default methods make.

-   ⚠️ **DDA adds a second, technical layer** of missingness on top (🔗 [W9 §2.2](w9_proteomics_tech.qmd#w9-s2-2)) — a protein can be absent because it was not selected, not because it was low
-   ✅ **DIA data is much cleaner in this respect**, which is most of why it is now the default

## 2) The four choices, and what each does to your result {#w10-s1-2}

::: {.callout-important collapse="true"}
## Each one is an assumption, not a cleaning step

-   ❌ **Drop any protein with a missing value.** You keep only the abundant, ubiquitously detected proteins — which is exactly the opposite of the set a novel biomarker would come from
-   ❌ **Impute with the row mean.** Low values get pulled up toward the middle, shrinking precisely the difference you are testing
-   ⚠️ **Impute from a low distribution** (e.g. a shifted, narrowed normal, as in Perseus). Usually a reasonable match to the MNAR mechanism — and still an assumption that belongs in the methods section
-   ✅ **Model missingness explicitly** — `MSstats`, or report a presence/absence test alongside the intensity test. Most defensible, because it stops pretending the NAs are not data

💡 **The reflexive check:** rerun with a different imputation and see whether your hit survives. If it does not, you have learned something important before a reviewer does.
:::

------------------------------------------------------------------------

# 2. Testing for Difference 📊 {#w10-s2}

## 1) Why `limma` and not a t-test {#w10-s2-1}

Proteomics datasets are wide and short — thousands of proteins, tens of samples. A per-protein t-test is unstable because each variance estimate rests on very few observations.

`limma` uses **empirical Bayes** to shrink each protein's variance toward the overall trend. In practice this means a protein that happens to look very consistent by chance is not automatically declared significant.

```{r}
library(limma); library(tidyverse)

# mat: log2 intensities, proteins × samples. meta: one row per sample.
design <- model.matrix(~ condition + sex + age + batch, data = meta)   # ⚠️ batch belongs here

fit <- lmFit(mat, design) |> eBayes(trend = TRUE, robust = TRUE)
res <- topTable(fit, coef = "conditionCase", number = Inf) |>
  as_tibble(rownames = "protein")
```

-   ✅ `trend = TRUE` accounts for the intensity–variance relationship typical of MS data
-   ⚠️ Include **batch, sex and age** in the design. Leaving them out does not make them go away
-   ✅ For MS designs with technical replicates or fractionation, **`MSstats`** models the hierarchy properly

## 2) What to report {#w10-s2-2}

-   Effect size (log2 fold change) **and** adjusted p-value — the same discipline as DESeq2 last semester
-   How many proteins were tested, and how many were dropped for missingness. ⚠️ Multiple testing is over the proteins you tested, not the ones that survived (🔗 [W2 §2.4](../part1/w2_scripts_ai.qmd#w2-s2-4))
-   Whether the result survives a different imputation

------------------------------------------------------------------------

# 3. ⚠️ From a Difference to a Biomarker 🪜 {#w10-s3}

A significant difference between groups is the beginning of a very long process.

![Each row loses about 90% of what entered it.](/assets/img/w10/biomarker_funnel.svg){fig-align="center"}

## 1) Three kinds of biomarker, often confused {#w10-s3-1}

| Type | Asks | Requires |
|--------------------|--------------------------|--------------------------|
| **Diagnostic** | Does this patient have the disease *now*? | Cases and the right differential diagnosis, at presentation |
| **Prognostic** | What will happen to this patient? | Follow-up, outcomes, and a survival model |
| **Predictive** | Will this patient respond to *this* drug? | ⚠️ A treated and an untreated arm — otherwise you cannot separate prediction from prognosis |

⚠️ **Predictive is the hardest and the most often claimed.** "Responders had higher levels than non-responders" is compatible with the marker simply indicating better-prognosis disease, regardless of treatment.

## 2) ⚠️ Why an AUC of 0.94 can be worthless {#w10-s3-2}

![The number is real. The study answered a question nobody asks.](/assets/img/w10/wrong_comparison.svg){fig-align="center"}

::: {.callout-warning collapse="true"}
## Back to the motivating question

**No** — not from that study.

The paper compared advanced cancer with healthy volunteers. Your patient is neither. To be useful in your clinic, the marker must separate **early cancer from the conditions that mimic it, sampled at presentation**.

And there is a second problem hiding underneath. If cases were bled in an oncology ward and controls at a screening centre, then collection tube, time to processing and fasting state all differ systematically between groups (🔗 [W9 §5](w9_proteomics_tech.qmd#w9-s5)). Some of that AUC is measuring **where the blood was drawn**.

→ ✅ Ask: were cases and controls sampled at the same point in the care pathway, in the same place, into the same tubes?
:::

## 3) Prevalence changes everything at the bedside {#w10-s3-3}

AUC is prevalence-independent; **positive predictive value is not**.

```
Sensitivity 90%, specificity 90%:

  prevalence 50% (a trial cohort)   →  PPV ≈ 90%
  prevalence  1% (a screening clinic) →  PPV ≈  8%
```

⚠️ The same test. Ninety-two false positives per hundred positive results, and every one of them gets a CT scan.

💡 This is the calculation you can do and the bioinformatician cannot — because it needs the prevalence in **your** clinic.

------------------------------------------------------------------------

# 4. Designing One You Could Actually Run 📋 {#w10-s4}

**① Name the clinical decision.** Not "find biomarkers for X" — *"distinguish A from B, at the moment I currently have to guess."*

**② Name the comparator group**, and make it the one you would really have to tell apart.

**③ Fix the collection protocol before the first sample.** Same tubes, same time-to-spin, same aliquot volume, freeze–thaw logged. Written down, one page.

**④ Randomise across batches.** 🔗 [W1 §3.3](../part1/w1_intro.qmd#w1-s3-3) — this is where it bites hardest.

**⑤ Reserve a validation set before you look at anything.** Split by patient, ideally by recruitment site or time period.

**⑥ Decide in advance what would count as success**, in terms of a decision changed — not an AUC threshold.

::: {.callout-tip collapse="true"}
## The version of this a resident can actually do

You do not need a new prospective cohort to start.

-   ✅ Many hospitals bank residual serum or plasma. Ask what exists and under what consent
-   ✅ A retrospective set selected **by outcome** — responders vs non-responders, relapsed vs not — is a stronger design than consecutive recruitment, and it already has follow-up attached
-   ⚠️ The catch: banked samples were collected under whatever conditions prevailed at the time, so pre-analytical variation is a real risk. Check whether collection was protocolised
-   ✅ An affinity panel on 300 banked samples is often cheaper and far more informative than MS on 30 fresh ones
:::

------------------------------------------------------------------------

# 5. Hands-On 💻 {#w10-s5}

## 1) Handle the missingness deliberately {#w10-s5-1}

⚠️ The dataset is simulated (patient specimens cannot be shared), and **batch is partly confounded with condition** — B1 holds 28 controls and 12 cases, B2 the reverse. That is deliberate: it is what a real banked cohort looks like, because cases and controls were collected at different times.

```{r}
library(tidyverse); library(limma)

prot <- readRDS(here::here("rawdata", "plasma_proteomics.rds"))  # proteins × samples, log2
meta <- readRDS(here::here("rawdata", "plasma_meta.rds"))

table(meta$batch, meta$condition)    # ⚠️ look at this before anything else

# ⚠️ Look before deciding
miss <- rowMeans(is.na(prot))
hist(miss, breaks = 40, main = "Fraction missing per protein")

# Is missingness related to abundance? (if yes → MNAR)
plot(rowMeans(prot, na.rm = TRUE), miss,
     xlab = "mean intensity", ylab = "fraction missing", pch = 16, col = "#185FA5")
```

💡 A clear negative trend in that second plot **is** the MNAR evidence. Show it in your supplement.

## 2) Test, with the covariates that matter {#w10-s5-2}

```{r}
keep <- miss < 0.30                       # state this threshold in the methods
mat  <- prot[keep, ]

design <- model.matrix(~ condition + age + sex + batch, data = meta)
fit    <- lmFit(mat, design) |> eBayes(trend = TRUE, robust = TRUE)
res    <- topTable(fit, coef = "conditionCase", number = Inf) |>
  as_tibble(rownames = "protein")

sum(res$adj.P.Val < 0.05)
```

## 3) ✅ Change one thing — three times {#w10-s5-3}

```{r}
# ① Drop batch from the design. How many hits change?
#    (you should go from ~40 proteins to ~209 — that gap is 169 false positives)
# ② Impute from a low distribution instead of filtering. Do the top 10 survive?
# ③ Split by patient into train/test, fit on train, evaluate AUC on test.
#    Compare that AUC with the one from the full dataset.
```

⚠️ ③ is the important one. The gap between the in-sample AUC and the held-out AUC is the size of your overfitting, and it is usually larger than people expect.

## 4) Design task, continued {#w10-s5-4}

Take the five sentences you wrote in 🔗 [W9 §6.3](w9_proteomics_tech.qmd#w9-s6-3) and add:

-   the **comparator group** you would actually need
-   the **decision** the result would change
-   the **prevalence** of the condition in the clinic where it would be used

Bring it to W14 — this is a legitimate capstone project.

------------------------------------------------------------------------

# 📝 Summary

-   ⚠️ **Proteomics missing values are mostly "too low to detect"** — MNAR, not MCAR
-   ❌ Dropping proteins with NAs keeps only abundant ones; ❌ mean imputation shrinks the effect you are testing
-   ✅ Impute from a low distribution, or model missingness explicitly (`MSstats`), and **state which**
-   ✅ Rerun with a different imputation. If the hit does not survive, better you find out than a reviewer
-   `limma` with empirical Bayes is more stable than per-protein t-tests; ⚠️ put batch, age and sex in the design
-   ⚠️ Discovery → verification → validation → clinical. **Each step loses ~90%**, and the second step is the one that gets skipped
-   ⚠️ **Diagnostic, prognostic and predictive are different claims.** Predictive needs an untreated arm
-   ⚠️ **AUC 0.94 against healthy volunteers may be a detector for being very ill** — and partly for where the blood was drawn
-   ✅ The control group is whoever you would actually have to tell apart, sampled at the same point in the pathway
-   ⚠️ **PPV collapses at low prevalence.** 90/90 at 1% prevalence gives a PPV near 8%
-   ✅ Banked, outcome-selected samples plus an affinity panel is the realistic entry point for a resident

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| MNAR / MCAR | Missing not at random / completely at random |
| Left-censoring | Values below the detection limit recorded as absent |
| Imputation | Filling missing values — always an assumption |
| MSstats | Proteomics package that models missingness and study structure |
| Empirical Bayes | Shrinking per-protein variance toward the overall trend (`limma`) |
| Discovery / verification / validation | The three pre-clinical stages of biomarker work |
| Diagnostic biomarker | Does the patient have the disease now? |
| Prognostic biomarker | What will happen regardless of treatment? |
| Predictive biomarker | Will this patient benefit from this drug? |
| Comparator group | The population the test must actually distinguish cases from |
| PPV | Positive predictive value — depends on prevalence |
| Overfitting | Performance that does not survive a held-out cohort |

------------------------------------------------------------------------

# ➕ References

### Missing values and statistics

Lazar C, et al. [*Accounting for the multiple natures of missing values in label-free quantitative proteomics data sets.*](https://doi.org/10.1021/acs.jproteome.5b00981) J Proteome Res. 2016. **← the reference on MNAR vs MCAR in proteomics**\
Choi M, et al. [*MSstats: an R package for statistical analysis of quantitative mass spectrometry-based proteomic experiments.*](https://doi.org/10.1093/bioinformatics/btu305) Bioinformatics. 2014.\
Ritchie ME, et al. [*limma powers differential expression analyses for RNA-sequencing and microarray studies.*](https://doi.org/10.1093/nar/gkv007) Nucleic Acids Res. 2015.

### Biomarker development and its failures

Pepe MS, et al. [*Phases of biomarker development for early detection of cancer.*](https://doi.org/10.1093/jnci/93.14.1054) J Natl Cancer Inst. 2001. **← the five-phase framework; still the clearest statement of the problem**\
Ioannidis JPA, Bossuyt PMM. [*Waste, leaks, and failures in the biomarker pipeline.*](https://doi.org/10.1373/clinchem.2016.254649) Clin Chem. 2017.\
Geyer PE, et al. [*Revisiting biomarker discovery by plasma proteomics.*](https://doi.org/10.15252/msb.20156297) Mol Syst Biol. 2017.\
Collins GS, et al. [*Transparent reporting of a multivariable prediction model for individual prognosis or diagnosis (TRIPOD).*](https://doi.org/10.7326/M14-0697) Ann Intern Med. 2015. **← the reporting checklist to write against**

### Study design

Bossuyt PM, et al. [*STARD 2015: an updated list of essential items for reporting diagnostic accuracy studies.*](https://doi.org/10.1136/bmj.h5527) BMJ. 2015.\
Rifai N, Gillette MA, Carr SA. [*Protein biomarker discovery and validation: the long and uncertain path to clinical utility.*](https://doi.org/10.1038/nbt1235) Nat Biotechnol. 2006.
