---
title: "Data Interpretation & Visualization"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 13"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W13 — Visualization"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Making Figures That Tell a Story {background-color="#1D9E75"}

---

## Why Visualization Matters

::: {.incremental}
- A great analysis with bad figures **doesn't get published**
- A mediocre analysis with compelling figures **gets attention**
- Reviewers form first impressions from Figure 1 before reading methods
- For your capstone: your hypothesis lives or dies by how clearly you present it
:::

::: {.fragment}
> **The figure is the argument.** The methods section just explains how you got there.
:::

---

## The Five Principles

1. **One message per panel** — don't overload a single plot
2. **Direct labeling over legends** — reduce eye travel
3. **Color encodes meaning** — don't use color just to be colorful
4. **Whitespace is not wasted** — clutter kills clarity
5. **Think about the reader's path** — A → B → C

---

# ggplot2 — Advanced Techniques {background-color="#534AB7"}

---

## The Grammar of Graphics Mindset

```{r}
library(ggplot2)
library(tidyverse)
library(here)

# Everything in ggplot2 is a layer
ggplot(data, aes(x, y)) +          # data + aesthetics
  geom_point() +                    # geometry
  stat_smooth(method = "lm") +      # statistical transformation
  scale_color_manual(...) +         # scale
  facet_wrap(~ group) +             # facet
  labs(title = "...", x = "...") +  # labels
  theme_minimal()                    # theme
```

::: {.fragment}
::: {.callout-tip}
## Build plots incrementally
Start with `ggplot(data, aes()) + geom_point()`.  
Add one layer at a time. Print after each addition.  
Never write a 20-line ggplot in one go without checking.
:::
:::

---

## Volcano Plot — Publication Ready

```{r}
library(ggplot2)
library(ggrepel)

# Assume: de_results from DESeq2
de_results <- de_results |>
  mutate(
    status = case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE                               ~ "NS"
    ),
    label = ifelse(padj < 1e-10 | abs(log2FoldChange) > 3,
                   gene, NA)
  )

ggplot(de_results, aes(x = log2FoldChange,
                        y = -log10(padj),
                        color = status)) +
  geom_point(alpha = 0.6, size = 0.8) +
  geom_text_repel(aes(label = label),
                   size     = 3,
                   max.overlaps = 20,
                   box.padding  = 0.3) +
  scale_color_manual(values = c(Up   = "#D85A30",
                                 Down = "#185FA5",
                                 NS   = "gray70")) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "gray40") +
  labs(x     = "log₂ fold change",
       y     = "-log₁₀(adjusted p-value)",
       color = "",
       title = "Differential expression — Case vs. Control") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank())
```

---

## Heatmap — ComplexHeatmap

```{r}
library(ComplexHeatmap)
library(circlize)

# Color scale — always center on 0 for scaled data
col_fun <- colorRamp2(c(-2, 0, 2),
                       c("#185FA5", "white", "#D85A30"))

# Column annotation
ha <- HeatmapAnnotation(
  Group    = sample_metadata$group,
  Age      = sample_metadata$age,
  col      = list(
    Group = c(Control = "#1D9E75", Case = "#D85A30"),
    Age   = colorRamp2(c(20, 80), c("white", "#534AB7"))
  ),
  annotation_name_side = "left"
)

# Main heatmap
ht <- Heatmap(
  scaled_matrix,
  name                    = "z-score",
  col                     = col_fun,
  top_annotation          = ha,
  show_column_names       = FALSE,
  clustering_distance_rows = "pearson",
  clustering_method_rows  = "ward.D2",
  row_names_gp            = gpar(fontsize = 8),
  border                  = TRUE,
  row_title               = "Top DE genes"
)

draw(ht, heatmap_legend_side = "right")
```

---

## Combining Plots — patchwork

```{r}
library(patchwork)

# Build individual panels
p1 <- ggplot(data, aes(group, expression, fill = group)) +
  geom_boxplot() + labs(tag = "A")

p2 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = cell_type)) +
  geom_point(size = 0.5) + labs(tag = "B")

p3 <- ggplot(de_results, aes(log2FoldChange, -log10(padj))) +
  geom_point() + labs(tag = "C")

p4 <- ggplot(pathway_df, aes(NES, pathway)) +
  geom_col() + labs(tag = "D")

# Combine into figure layout
(p1 | p2) /
(p3 | p4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

# Save as high-resolution PDF
ggsave(here("results", "figures", "figure1.pdf"),
       width = 10, height = 8,
       device = cairo_pdf)
```

---

## Color — The Most Misused Element

```{r}
library(RColorBrewer)
library(ggsci)     # Nature, Science, NEJM color palettes

# For categorical data — colorblind-friendly
scale_color_manual(values = c(
  "#0072B2",   # blue
  "#D55E00",   # orange
  "#009E73",   # green
  "#CC79A7",   # pink
  "#E69F00",   # yellow
  "#56B4E9",   # light blue
  "#F0E442"    # yellow (use sparingly)
))

# For sequential data
scale_fill_distiller(palette = "Blues", direction = 1)

# For diverging data (fold change, correlation)
scale_fill_gradient2(low = "#185FA5", mid = "white",
                      high = "#D85A30", midpoint = 0)

# Nature journal palette
scale_color_npg()    # ggsci::scale_color_npg()

# NEJM palette
scale_color_nejm()
```

---

## Dot Plot for Pathway Results

```{r}
# Better than bar charts for enrichment results
pathway_df |>
  slice_max(abs(NES), n = 20) |>
  ggplot(aes(x = NES,
             y = reorder(pathway, NES),
             size  = setSize,
             color = padj)) +
  geom_point() +
  scale_color_gradient(low  = "#D85A30",
                       high = "gray80",
                       name = "adj. p-value",
                       trans = "log10") +
  scale_size_continuous(range = c(2, 8),
                        name  = "Gene set size") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x     = "Normalized enrichment score (NES)",
       y     = "",
       title = "GSEA — top pathways") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_line(color = "gray90"))
```

---

# Figure Design for Publications {background-color="#185FA5"}

---

## Panel Layout Principles

```{r}
# Use consistent sizing
# Journal column widths:
# Single column: 3.5 inches
# 1.5 column:    5 inches
# Double column: 7 inches

# Always set figure size explicitly in ggsave
ggsave(
  filename = here("results", "figures", "fig2.pdf"),
  plot     = my_figure,
  width    = 7,      # double column
  height   = 5,
  units    = "in",
  device   = cairo_pdf   # for proper font embedding
)

# For raster formats (e.g., presentations):
ggsave(
  filename = here("results", "figures", "fig2.png"),
  plot     = my_figure,
  width    = 7,
  height   = 5,
  dpi      = 300    # minimum for publication
)
```

---

## The Axis Trap

```{r}
# BAD: truncated y-axis exaggerates differences
ggplot(data, aes(group, value)) +
  geom_col() +
  coord_cartesian(ylim = c(95, 102))   # NEVER do this

# GOOD: start from 0 for bar charts
ggplot(data, aes(group, value)) +
  geom_col() +
  expand_limits(y = 0)

# Exception: show variation when absolute value is less important
# (e.g., gene expression changes) → boxplot / violinplot, not bar
ggplot(data, aes(group, expression)) +
  geom_violin(aes(fill = group)) +
  geom_boxplot(width = 0.1, fill = "white") +
  geom_jitter(width = 0.05, alpha = 0.3, size = 0.5)
```

---

## Statistical Annotation

```{r}
library(ggpubr)
library(rstatix)

# Compute statistics
stat_test <- data |>
  t_test(expression ~ group) |>
  add_significance() |>
  add_xy_position(x = "group")

# Add to plot
ggplot(data, aes(group, expression, fill = group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1) +
  stat_pvalue_manual(stat_test,
                     label     = "p.adj.signif",
                     tip.length = 0.01) +
  scale_fill_manual(values = c("#1D9E75", "#D85A30")) +
  labs(y = "Expression (normalized)") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Reporting Standards {background-color="#0F6E56"}

---

## What Must Be Reported

::: {.incremental}
- **Sample sizes** — always, for every group
- **Statistical tests** — name the test, not just "p < 0.05"
- **Multiple testing correction** — BH, Bonferroni — which and why
- **Effect sizes** — OR, HR, log2FC — not just p-values
- **Software versions** — R version + package versions (`sessionInfo()`)
- **Data availability** — GEO accession numbers, GitHub code repository
- **Reproducibility** — seeds set, renv snapshot included
:::

---

## sessionInfo() — Non-Negotiable

```{r}
# Always include at the end of your analysis script
writeLines(
  capture.output(sessionInfo()),
  here("results", "session_info.txt")
)

# For renv — snapshot locks all package versions
renv::snapshot()

# For methods section — generate package citation list
citation("Seurat")
citation("DESeq2")
citation("ggplot2")
```

---

# Capstone Check-In {background-color="#BA7517"}

---

## Where Are You?

By now you should have:

::: {.incremental}
- ✅ A **disease or biological question** in mind
- ✅ A **public dataset** identified (GEO, TCGA, curatedMetagenomicData, UK Biobank)
- ✅ At least **one analysis method** from the course that fits
- 🔲 A **one-sentence hypothesis** — what do you expect to find?
:::

---

## Good Capstone Questions (Examples)

::: {.incremental}
- "Do patients with high PRS for T2D show distinct gut microbiome composition in the HMP2 cohort?"
- "Which cell types express the top AD GWAS genes, and do they show altered states in disease?"
- "Can CMAP identify drugs that reverse the transcriptomic signature of senescent cardiomyocytes?"
- "Are GWAS hits for IBD enriched in open chromatin of intestinal epithelial cells vs. immune cells?"
:::

::: {.fragment}
::: {.callout-important}
## Keep it focused
One question, one dataset, two methods.  
A clean answer to a focused question is better than a messy tour of everything.
:::
:::

---

## Hands-On: Your Visualization

```{r}
# Workshop time — work on your capstone figure

# Template: publication-ready plot skeleton
capstone_plot <- ggplot(your_data,
                         aes(x_var, y_var, color = group)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("#1D9E75", "#D85A30")) +
  labs(
    title    = "Your finding in one sentence",
    subtitle = "Dataset | n = XX",
    x        = "X axis label (unit)",
    y        = "Y axis label (unit)",
    color    = "Group",
    caption  = "Data: GEO accession GSEXXXXX"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(size = 14, face = "plain"),
    plot.subtitle = element_text(color = "gray50"),
    legend.position = "top"
  )

capstone_plot
```

---

# Next Week {background-color="#534AB7"}

---

## W14 — Rscript, Vibe Coding & Project Design

**What we'll cover:**

- Writing clean, shareable R scripts
- Vibe coding with Claude / GitHub Copilot — how to use AI assistance effectively
- Structuring your capstone analysis
- Individual project design session

**Bring:**

- Your dataset (downloaded and in `data/raw/`)
- Your one-sentence hypothesis
- A first attempt at loading the data

---

## Key Resources

- **ggplot2 book** — [ggplot2-book.org](https://ggplot2-book.org) — free online
- **Fundamentals of Data Visualization** — [clauswilke.com/dataviz](https://clauswilke.com/dataviz)
- **patchwork** — [patchwork.data-imaginist.com](https://patchwork.data-imaginist.com)
- **ComplexHeatmap** — [jokergoo.github.io/ComplexHeatmap-reference](https://jokergoo.github.io/ComplexHeatmap-reference)
- **ggsci** — journal color palettes (Nature, Science, NEJM, Lancet)

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Rscript, Vibe Coding & Individual Project Design**

Come with your dataset and hypothesis ready!

Course materials → [your GitHub Pages URL here]
:::
