---
title: "Rscript, Vibe Coding & Individual Project Design"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 14"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W14 — Vibe Coding & Project Design"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Writing Code That Lasts {background-color="#1D9E75"}

---

## The Problem with Analysis Scripts

Code written during analysis tends to be:

::: {.incremental}
- One massive 500-line script with no structure
- Variables named `df2`, `df_final`, `df_final2`, `df_FINAL_USE_THIS`
- Hardcoded paths that only work on your laptop
- No comments — you wrote it, you know what it does (for now)
- Three months later: no idea what `tmp3` is
:::

::: {.fragment}
> **For a capstone project that you might actually use in research:**  
> write code as if a colleague needs to run it without you.
:::

---

## Anatomy of a Good R Script

```r
# ============================================================
# W14 Demo: Clean R Script Structure
# Project: IBD microbiome analysis
# Author:  Your Name
# Date:    2025-01-01
# Input:   data/raw/hmp2_phyloseq.rds
# Output:  results/figures/diversity_plot.pdf
#          results/tables/da_taxa.tsv
# ============================================================

# 0. Libraries -----------------------------------------------
library(phyloseq)
library(tidyverse)
library(here)

# 1. Load data -----------------------------------------------
ps <- readRDS(here("data", "raw", "hmp2_phyloseq.rds"))

# 2. QC & filtering ------------------------------------------
ps_clean <- prune_samples(sample_sums(ps) > 1000, ps)
ps_clean <- prune_taxa(taxa_sums(ps_clean) > 10, ps_clean)

# 3. Analysis ------------------------------------------------
ps_rare  <- rarefy_even_depth(ps_clean, rngseed = 42)
alpha    <- estimate_richness(ps_rare, measures = "Shannon")

# 4. Visualization -------------------------------------------
p <- ggplot(...) + ...

# 5. Save outputs --------------------------------------------
ggsave(here("results", "figures", "diversity_plot.pdf"), p,
       width = 7, height = 5)
write_tsv(da_taxa, here("results", "tables", "da_taxa.tsv"))
```

---

## Named Code Sections in RStudio

```r
# Use ---- at end of comment to create collapsible sections
# (Ctrl/Cmd + Shift + R to insert section)

# Libraries ----
library(tidyverse)

# Load data ----
data <- read_tsv(...)

# Analysis ----
results <- ...

# Visualize ----
p <- ggplot(...) + ...
```

::: {.fragment}
Sections appear in the **outline panel** (bottom-left of script editor).  
Navigate large scripts instantly without scrolling.
:::

---

## Writing Functions — Don't Repeat Yourself

```{r}
# BAD — copy-pasted analysis per cell type
cd8_markers <- FindMarkers(seurat, ident.1 = "CD8 T", ...)
cd4_markers <- FindMarkers(seurat, ident.1 = "CD4 T", ...)
nk_markers  <- FindMarkers(seurat, ident.1 = "NK", ...)

# GOOD — write a function once
get_markers <- function(seurat_obj, cell_type,
                         min_pct = 0.25, logfc = 0.5) {
  FindMarkers(seurat_obj,
              ident.1         = cell_type,
              min.pct         = min_pct,
              logfc.threshold = logfc) |>
    as_tibble(rownames = "gene") |>
    mutate(cell_type = cell_type)
}

# Apply to all cell types
all_markers <- map_dfr(
  c("CD8 T", "CD4 T", "NK", "B cell", "Monocyte"),
  ~ get_markers(seurat, .x)
)
```

---

## Quarto Documents for Reproducible Reports

```{r}
# Your capstone should be a .qmd file that:
# 1. Loads data
# 2. Runs all analyses
# 3. Produces all figures and tables
# 4. Renders to HTML / PDF with one command

# quarto render capstone.qmd --to html
# quarto render capstone.qmd --to pdf

# Inline results in text:
# "We identified `r nrow(sig_genes)` significant genes
#  (adj. p < 0.05, |log2FC| > 1)."
# → auto-updates when you re-run the analysis

# No more "copy number from table → paste into Word" errors
```

---

# Vibe Coding with AI {background-color="#534AB7"}

---

## What Is Vibe Coding?

::: {.fragment}
> **Vibe coding** (Karpathy, 2025):  
> Using AI language models (Claude, GitHub Copilot, ChatGPT)  
> to write code by describing what you want in natural language,  
> then reviewing, testing, and iterating.
:::

::: {.fragment}
For bioinformatics:

- "Write a function that takes a DESeq2 results object and plots a volcano plot"
- "Convert this phyloseq object to a long-format tibble with sample metadata merged"
- "Explain what this ggplot code is doing and suggest improvements"
:::

---

## How to Use AI Assistance Effectively

**Good prompts:**

```
Context: I'm analyzing a Seurat object with cell types in the
         "cell_type" column and disease status in "condition".
Task:    Write a pseudobulk DESeq2 function that takes the Seurat
         object and a cell type name, aggregates counts per sample,
         and returns a DESeq2 results table with gene symbols.
Constraints: Use dplyr, DESeq2, and the here package.
             Return a tibble sorted by padj.
```

::: {.fragment}
**Bad prompt:**
```
write code for my analysis
```
:::

---

## The AI Assistance Workflow

```
1. Describe your goal clearly (context + task + constraints)
         ↓
2. Get AI-generated code
         ↓
3. READ the code — understand every line
         ↓
4. Run it on a small subset first
         ↓
5. Check the output makes biological sense
         ↓
6. Ask AI to explain anything unclear
         ↓
7. Iterate
```

::: {.callout-important}
## Non-negotiable
**You** are responsible for the code in your analysis.  
AI can write it. You must understand it.  
"The AI wrote it" is not an explanation in a methods section.
:::

---

## Useful Prompts for Bioinformatics

```
"Explain what this error message means and how to fix it:
[paste error]"

"I have a data frame with columns [x, y, z].
Reshape it so that..."

"What is the difference between rlog and vst normalization
in DESeq2? When should I use each?"

"Review this code and suggest improvements for readability
and performance: [paste code]"

"Generate a template for a ggplot volcano plot with labeled
top hits, colored by significance, publication-ready style."
```

---

## Debugging with AI

```{r}
# Common pattern: paste error + code context
# "I get this error when running this code. What's wrong?"

# Example error:
# Error in FindMarkers(seurat, ident.1 = "CD8 T") :
#   'ident.1' must be a valid identity class

# AI will recognize:
# - Seurat must have identities set with Idents()
# - "CD8 T" might not match exactly (space, case, spelling)

# Fix:
levels(Idents(seurat))    # check valid identity names
Idents(seurat) <- "cell_type"   # set correct column
FindMarkers(seurat, ident.1 = levels(Idents(seurat))[1])
```

---

# GitHub for Science {background-color="#185FA5"}

---

## Why GitHub for Your Capstone

::: {.incremental}
- **Version control** — never lose work again
- **Reproducibility** — reviewers can run your code
- **GitHub Pages** — host your Quarto slides (like this course!)
- **Portfolio** — show future employers/collaborators your work
- **Collaboration** — share with your PI / co-authors cleanly
:::

---

## Basic Git Workflow

```bash
# One-time setup
git config --global user.name "Your Name"
git config --global user.email "you@hospital.ac.kr"

# Start a new project
git init
git add .
git commit -m "Initial commit: project structure + renv"

# After each analysis session
git add R/analysis.R results/figures/
git commit -m "Add DESeq2 analysis and volcano plot"
git push origin main
```

---

## Publishing to GitHub Pages with Quarto

```bash
# In your project directory
quarto publish gh-pages

# That's it — Quarto handles the rest:
# 1. Renders all .qmd files
# 2. Pushes HTML to gh-pages branch
# 3. GitHub Pages serves at:
#    https://yourusername.github.io/your-repo/
```

::: {.fragment}
Your capstone presentation will be at a public URL  
you can share with anyone, on any device, immediately after W15.
:::

---

# Individual Project Design {background-color="#BA7517"}

---

## Capstone Structure (W15 Presentation)

```
1. Background (2 min)
   └── Disease, clinical problem, why it matters

2. Research question & hypothesis (1 min)
   └── One clear question, one directional hypothesis

3. Data & Methods (3 min)
   └── Dataset, analysis pipeline, key tools

4. Results (5–7 min)
   └── 3–5 key figures, each with one message

5. Interpretation & Discussion (2 min)
   └── Does this support your hypothesis?
   └── Limitations, next steps

6. Clinical implications (1 min)
   └── If this is true, what does it mean for patients?

Total: 15 min + 5 min Q&A
```

---

## Figure Plan Template

```
Figure 1: Overview / data quality
  → "We analyzed X samples from Y dataset. After QC, Z remained."

Figure 2: Main result — the answer to your question
  → The one plot that shows your hypothesis is supported or rejected

Figure 3: Mechanistic insight
  → "This is WHY we see what we see in Figure 2"

Figure 4 (optional): Clinical relevance
  → "This finding could translate to..."
```

---

## Workshop — Project Check-In

Work with a partner (10 min each):

1. **State your question** — one sentence
2. **Show your data** — `glimpse()` or `str()` output
3. **Sketch Figure 1** — even on paper
4. **Get feedback** — is the question focused enough?

::: {.callout-tip}
## Most common issue
The question is too broad:
❌ "I want to analyze the gut microbiome in my disease"  
✅ "I will test whether patients with high CRP have lower Faecalibacterium abundance in the HMP2 IBD cohort"
:::

---

## Common Issues — Troubleshooting Guide

```{r}
# Issue 1: Package won't install
# Solution: check Bioconductor version match
BiocManager::valid()

# Issue 2: Object not found
# Solution: check environment, renv status
ls()
renv::status()

# Issue 3: Memory error with large dataset
# Solution: use sparse matrices, subsample, or bigmemory
# For Seurat: downsample to 10k cells for exploration
seurat_small <- seurat[, sample(ncol(seurat), 10000)]

# Issue 4: Plot looks different from example
# Solution: check package version
packageVersion("Seurat")
packageVersion("ggplot2")
```

---

## The Checklist Before W15

::: {.incremental}
- ✅ Data loaded and QC'd
- ✅ At least one main analysis complete
- ✅ Figure 1 (overview) ready
- ✅ Figure 2 (main result) ready — even if rough
- ✅ One-slide methods summary
- ✅ GitHub repo created, code committed
- ✅ QMD renders without errors
- 🔲 Practice the 15-minute talk (time yourself)
:::

---

# Next Week {background-color="#534AB7"}

---

## W15 — Final Presentations

**Format:**
- 15 minutes presentation + 5 minutes Q&A
- Use your GitHub Pages QMD slides
- Focus on the story, not exhaustive results

**Judging criteria:**
- Clarity of research question
- Appropriateness of methods
- Quality of figures
- Biological interpretation
- Clinical relevance

::: {.fragment}
> **It's okay if your hypothesis is wrong.**  
> A well-conducted analysis that disproves a hypothesis  
> is better science than p-hacking until something "works."
:::

---

## Key Resources for This Week

- **Happy Git with R** — [happygitwithr.com](https://happygitwithr.com) — Git setup for R users
- **Quarto publishing** — [quarto.org/docs/publishing/github-pages](https://quarto.org/docs/publishing/github-pages)
- **R for Data Science** — [r4ds.hadley.nz](https://r4ds.hadley.nz) — modern R coding style
- **GitHub Student Pack** — free private repos if you have an academic email

---

## Final Thought {background-color="#1D9E75"}

::: {.center}
You started this course knowing bulk RNA-seq.

You now know GWAS, PRS, methylation, chromatin, microbiome,
single-cell, spatial transcriptomics, multi-omics, and drug discovery.

**That is not a small thing.**

Next week: show us what you can do with it.

Course materials → [your GitHub Pages URL here]
:::
