---
title: "Introduction to Advanced Bioinformatics & R Environment Setup"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 1"
author: "Eunji Ha"
date: last-modified
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
    code-fold: false
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W1"
    logo: ""
    css: custom.css
execute:
  echo: true
  eval: false
---

# Why Are We Here? {background-color="#1D9E75"}

------------------------------------------------------------------------

## The Gap No One Talks About

::::: columns
::: {.column width="50%"}
**What's happening in research**

-   Whole genome sequencing at clinical scale
-   Single-cell atlases of every tissue
-   Multi-omics datasets in public repositories
-   AI/ML-driven drug discovery pipelines
:::

::: {.column width="50%"}
**What most clinicians can do with it**

-   Read the abstract
-   Trust the figures
-   Ask a bioinformatician
:::
:::::

::: fragment
> **Clinician-scientists who can analyze their own data ask better questions.**
:::

------------------------------------------------------------------------

## What You Already Know

From last semester:

-   ✅ R basics — data wrangling, ggplot2, statistics\
-   ✅ Bulk RNA-seq — QC, alignment, DESeq2, pathway analysis\
-   ✅ Thinking computationally about biological questions

::: fragment
This semester: we go **one (or several) levels deeper.**
:::

------------------------------------------------------------------------

# The Big Picture {background-color="#534AB7"}

------------------------------------------------------------------------

## One Story, Many Tools

From a GWAS hit to a drug target — this is what the semester builds toward:

```         
GWAS / PRS          → Which genetic variants increase disease risk?
       ↓
Epigenomics         → Which regulatory elements are involved?
       ↓
Metagenomics        → Does the microbiome play a role?
       ↓
scRNA-seq / Spatial → Which cell types? In which tissue context?
       ↓
Multi-omics         → How do all these layers converge on a target?
       ↓
Drug Discovery      → Can we reposition an existing drug?
```

::: fragment
Every tool we learn is a **piece of the same puzzle.**
:::

------------------------------------------------------------------------

## Semester Roadmap

| Section | Weeks | Topics |
|----|----|----|
| **Introduction** | W1 | R environment setup |
| **Clinical Genomics & Risk Prediction** | W2–W3 | GWAS, Post-GWAS, PRS |
| **High-Resolution Omics** | W4–W10 | Epigenomics, Metagenomics, scRNA-seq, Spatial |
| **Multi-omics & Target Discovery** | W11–W12 | Integration, Network, Drug repositioning |
| **Capstone** | W13–W15 | Your own data story |

------------------------------------------------------------------------

## What You'll Be Able to Do by W15

::: incremental
-   Read a methods section and understand **what was actually done**
-   Download public omics datasets and **run analyses yourself**
-   Combine multiple data types to **generate a testable hypothesis**
-   **Present a data-driven research proposal** at the final session
:::

------------------------------------------------------------------------

## Capstone Project (W13–W15)

**W13** — Data interpretation & visualization\
**W14** — Rscript, vibe coding tips, individual project design\
**W15** — Final presentation: *Pitching the Hypothesis*

::: callout-tip
## Pick your topic early

The capstone works best when you bring a question from your own clinical domain.\
You don't need novel data — public repositories (GEO, GTEx, TCGA, UK Biobank) are fair game.
:::

------------------------------------------------------------------------

# Tools We'll Use {background-color="#185FA5"}

------------------------------------------------------------------------

## The R Ecosystem This Semester

| Topic          | Key Packages                       |
|----------------|------------------------------------|
| GWAS / PRS     | `bigsnpr`, `PRSice`, `LDlinkR`     |
| Epigenomics    | `minfi`, `ChAMP`, `ArchR`          |
| Metagenomics   | `phyloseq`, `vegan`, `DADA2`       |
| scRNA-seq      | `Seurat`, `SingleR`, `Monocle3`    |
| Spatial        | `Seurat`, `SpatialDE`, `NNSVG`     |
| Multi-omics    | `MOFA2`, `mixOmics`                |
| Drug discovery | `igraph`, `drugTargetInteractions` |

::: fragment
Don't worry — we install these as we go. Today: just the foundation.
:::

------------------------------------------------------------------------

# R Environment Setup {background-color="#0F6E56"}

------------------------------------------------------------------------

## Principles of a Reproducible Workflow

Three things that will save you later:

1.  **Project-based working** — never `setwd()` again\
2.  **Consistent directory structure** — know where everything lives\
3.  **Locked package versions** — your code runs the same in 2 years

------------------------------------------------------------------------

## R Project Structure

```         
my_project/
├── my_project.Rproj       # always open via this
├── renv.lock              # package versions locked here
├── data/
│   ├── raw/               # never modify these
│   └── processed/
├── R/
│   ├── 01_qc.R
│   └── 02_analysis.R
├── results/
│   ├── figures/
│   └── tables/
└── docs/                  # reports, QMD files
```

------------------------------------------------------------------------

## The `here` Package — No More Path Pain

```{r}
library(here)

# Instead of:
read.csv("C:/Users/me/Desktop/project/data/raw/counts.csv")

# Use:
read.csv(here("data", "raw", "counts.csv"))
```

Works identically on Windows, Mac, and Linux.\
Works regardless of where your Rproj file is.

------------------------------------------------------------------------

## Reproducible Environments with `renv`

```{r}
# 1. Initialize — run once when starting a project
renv::init()

# 2. Install packages as usual
install.packages("Seurat")
BiocManager::install("DESeq2")

# 3. Snapshot — lock your current package versions
renv::snapshot()

# 4. Restore — on a new machine or for a collaborator
renv::restore()
```

::: callout-important
## Why this matters

Without `renv`, a Seurat update 6 months from now may silently break your code.
:::

------------------------------------------------------------------------

# Hands-On {background-color="#BA7517"}

------------------------------------------------------------------------

## Let's Set Up Together

**Step 1** — Create a new R Project

```         
File → New Project → New Directory → New Project
Name: "bioinformatics_2025"
```

**Step 2** — Create directory structure

```{r}
dirs <- c("data/raw", "data/processed", "R", "results/figures", 
          "results/tables", "docs")
sapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```

**Step 3** — Initialize renv

```{r}
install.packages("renv")
renv::init()
```

------------------------------------------------------------------------

## Install Core Packages for This Semester

```{r}
# CRAN packages
install.packages(c("here", "tidyverse", "BiocManager", 
                   "ggplot2", "patchwork", "data.table"))

# Bioconductor
BiocManager::install(c("DESeq2", "minfi", "phyloseq", 
                       "SingleR", "celldex"))

# Snapshot your environment
renv::snapshot()
```

::: callout-note
Some packages (Seurat, ArchR) we'll install in the relevant week.\
Bioconductor installs can take a few minutes — this is normal.
:::

------------------------------------------------------------------------

## Verify Your Setup

```{r}
# Run this — everything should load without errors
library(here)
library(tidyverse)
library(BiocManager)

sessionInfo()
```

If you hit errors → paste the error message in our class channel.\
Common fix: `BiocManager::install(version = "3.19")`

------------------------------------------------------------------------

# Next Week {background-color="#533AB7"}

------------------------------------------------------------------------

## W2 Preview — GWAS & Post-GWAS

**What we'll cover:**

-   How GWAS works (and what it doesn't tell you)
-   Reading a Manhattan plot like a reviewer
-   LD, fine-mapping, functional annotation
-   Hands-on: analyzing a public GWAS summary stat dataset

**Before next class:**

```{r}
# Install in advance
install.packages("bigsnpr")
install.packages("LDlinkR")
```

Download data → link will be shared on the course page.

------------------------------------------------------------------------

## Questions? {background-color="#1D9E75"}

::: center
Next week: **Genome-wide Association Studies (GWAS) & Post-GWAS**

Course materials → \[your GitHub Pages URL here\]
:::
