---
title: "Single-cell RNA-seq Fundamentals: Cell-type Specificity"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 8"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W8 — scRNA-seq Fundamentals"
    css: custom.css
execute:
  echo: true
  eval: false
---

# Why Single-Cell? {background-color="#1D9E75"}

---

## The Problem with Bulk RNA-seq

From last semester — bulk RNA-seq:

```
Tumor biopsy → homogenize → RNA → counts per gene

Gene X: 150 counts
```

::: {.fragment}
But what does that mean?

```
Tumor cell A:    ████████  (high expression)
Tumor cell B:    ██        (medium)
Macrophage:      ████████  (high — confounding!)
Fibroblast:      ░░        (zero)
T cell:          ░░        (zero)

Bulk average: ████  ← misleading
```
:::

---

## What scRNA-seq Resolves

::: {.incremental}
- Every cell gets its own expression profile
- Identify **cell types** without prior knowledge
- Find **rare cell populations** (<1% of tissue)
- Discover **cell states** within a cell type
- Map **gene expression heterogeneity** within tumors
- Connect to **GWAS** — in which cell type is the risk gene expressed?
:::

::: {.fragment}
> scRNA-seq is now standard in Nature/Cell/Science papers.  
> As a clinician-scientist, you need to be able to interpret and run these analyses.
:::

---

## How 10x Chromium Works

```
Single-cell suspension
    ↓
Microfluidic droplet formation
(one cell + one bead + oil → GEM)
    ↓
Cell barcoding + UMI tagging
    ↓
cDNA synthesis + amplification
    ↓
Library sequencing
    ↓
Cell Ranger → count matrix
(cells × genes)
```

---

## The scRNA-seq Count Matrix

```
         Gene1  Gene2  Gene3  Gene4  ...
Cell_1      0      3      0     12
Cell_2      5      0      8      0
Cell_3      0      0      1      0
Cell_4      2      7      0      6
...

n_cells: 5,000–50,000
n_genes: ~20,000 (but 80–90% are zeros → sparse matrix)
```

::: {.fragment}
**The sparsity problem:**  
Each cell captures only ~5–15% of all transcripts → lots of zeros  
This is a technical artifact ("dropout") not true zero expression
:::

---

# The Seurat Pipeline {background-color="#534AB7"}

---

## Overview

```
Cell Ranger output (barcodes, features, matrix)
    │
    ├── 1. Load & create Seurat object
    ├── 2. QC → filter low-quality cells
    ├── 3. Normalization → SCTransform
    ├── 4. Feature selection → highly variable genes
    ├── 5. Scaling & PCA
    ├── 6. Clustering → FindNeighbors + FindClusters
    ├── 7. UMAP visualization
    ├── 8. Marker gene identification
    └── 9. Cell type annotation → SingleR / manual
```

---

## Step 1: Load Data

```{r}
library(Seurat)
library(tidyverse)
library(here)

# Load 10x Cell Ranger output
counts <- Read10X(data.dir = here("data", "raw",
                                   "cellranger_output",
                                   "filtered_feature_bc_matrix"))

# Create Seurat object
seurat <- CreateSeuratObject(
  counts    = counts,
  project   = "PBMC",
  min.cells = 3,      # gene must be in ≥3 cells
  min.features = 200  # cell must have ≥200 genes
)

seurat
# 23,000 features × 8,000 samples (cells)
```

---

## Step 2: QC — Three Key Metrics

```{r}
# Mitochondrial gene percentage
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat,
                                                pattern = "^MT-")

# Ribosomal gene percentage
seurat[["percent.rb"]] <- PercentageFeatureSet(seurat,
                                                pattern = "^RP[SL]")

# Visualize
VlnPlot(seurat,
        features = c("nFeature_RNA",
                     "nCount_RNA",
                     "percent.mt"),
        ncol     = 3,
        pt.size  = 0)

# Scatter plot — doublets have high nCount + high nFeature
FeatureScatter(seurat, "nCount_RNA", "nFeature_RNA",
               group.by = "orig.ident")
```

---

## QC Thresholds

```{r}
# Filter cells
seurat <- subset(seurat,
  nFeature_RNA > 200   &    # too few genes = dead cell
  nFeature_RNA < 5000  &    # too many genes = doublet
  nCount_RNA   < 25000 &    # too many UMIs  = doublet
  percent.mt   < 20)        # high MT% = damaged cell

cat("Cells before QC:", ncol(counts), "\n")
cat("Cells after QC: ", ncol(seurat), "\n")
```

::: {.callout-tip}
## Thresholds are dataset-specific
PBMC: MT% < 20% fine.  
Cardiomyocytes: MT% up to 40% is normal (high mitochondrial content).  
Always look at the distribution first, then set thresholds.
:::

---

## Step 3: Normalization — SCTransform

```{r}
# SCTransform: regression-based normalization
# Accounts for sequencing depth differences between cells
seurat <- SCTransform(
  seurat,
  vars.to.regress = "percent.mt",   # regress out MT%
  verbose         = FALSE
)

# Result: "SCT" assay added to Seurat object
# Use SCT for all downstream analysis
```

---

## Step 4–6: PCA & Clustering

```{r}
seurat <- RunPCA(seurat, verbose = FALSE)

# Elbow plot — how many PCs to use?
ElbowPlot(seurat, ndims = 50)

# Use top 30 PCs (adjust based on elbow plot)
seurat <- FindNeighbors(seurat, dims = 1:30, verbose = FALSE)
seurat <- FindClusters(seurat,
                        resolution = 0.5,   # higher = more clusters
                        verbose    = FALSE)

seurat <- RunUMAP(seurat, dims = 1:30, verbose = FALSE)

DimPlot(seurat, reduction = "umap",
        label = TRUE, pt.size = 0.5)
```

---

## Step 7: Marker Genes per Cluster

```{r}
# Find markers for each cluster vs. all other cells
all_markers <- FindAllMarkers(
  seurat,
  only.pos        = TRUE,    # only upregulated
  min.pct         = 0.25,    # expressed in ≥25% of cells
  logfc.threshold = 0.5,
  test.use        = "wilcox"
)

# Top 5 markers per cluster
top_markers <- all_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5)

# Heatmap of top markers
DoHeatmap(seurat,
          features = top_markers$gene,
          size     = 3) +
  theme(axis.text.y = element_text(size = 7))
```

---

## Visualizing Marker Genes

```{r}
# Feature plots — expression on UMAP
FeaturePlot(seurat,
            features = c("CD3E",    # T cells
                         "CD8A",    # CD8 T cells
                         "MS4A1",   # B cells
                         "CD14",    # Monocytes
                         "GNLY",    # NK cells
                         "PPBP"),   # Platelets
            ncol = 3,
            order = TRUE)

# Violin plots
VlnPlot(seurat,
        features = c("CD3E", "CD8A", "MS4A1"),
        ncol     = 3)
```

---

# Cell Type Annotation {background-color="#185FA5"}

---

## Manual vs. Automated Annotation

**Manual:**
- Look at top marker genes per cluster
- Check canonical markers (CD3E = T cell, CD14 = Monocyte...)
- Time-consuming but gives full control

**Automated — SingleR:**
- Compares each cell to reference transcriptomes
- Fast, reproducible, reference-dependent
- Best used as a starting point → manually validate

---

## SingleR — Automated Annotation

```{r}
library(SingleR)
library(celldex)

# Reference dataset — Human Primary Cell Atlas
hpca_ref <- HumanPrimaryCellAtlasData()

# Get normalized expression matrix
sce <- as.SingleCellExperiment(seurat)

# Run SingleR
predictions <- SingleR(
  test      = sce,
  ref       = hpca_ref,
  labels    = hpca_ref$label.fine,
  assay.type.test = "logcounts"
)

# Add predictions to Seurat object
seurat$singler_labels <- predictions$labels

# Visualize
DimPlot(seurat, group.by = "singler_labels",
        label = TRUE, repel = TRUE) +
  NoLegend()
```

---

## Doublet Detection

```{r}
library(DoubletFinder)

# Doublets = two cells captured in one droplet
# Look like cells with very high nFeature_RNA
# Must be removed before analysis

# Step 1: parameter sweep
sweep_res <- paramSweep_v3(seurat, PCs = 1:30,
                            sct = TRUE)
sweep_stats <- summarizeSweep(sweep_res, GT = FALSE)
bcmvn     <- find.pK(sweep_stats)
pK_val    <- as.numeric(as.character(
  bcmvn$pK[which.max(bcmvn$BCmetric)]))

# Step 2: run DoubletFinder
seurat <- doubletFinder_v3(
  seurat,
  PCs          = 1:30,
  pN           = 0.25,
  pK           = pK_val,
  nExp         = round(0.075 * ncol(seurat)),  # 7.5% expected
  reuse.pANN   = FALSE,
  sct          = TRUE
)

# Remove doublets
seurat <- subset(seurat, DF.classifications == "Singlet")
```

---

## Finalizing Annotations

```{r}
# Add manual annotations based on marker genes + SingleR
cell_type_map <- c(
  "0" = "CD4 T cell",
  "1" = "CD8 T cell",
  "2" = "B cell",
  "3" = "Classical Monocyte",
  "4" = "NK cell",
  "5" = "Non-classical Monocyte",
  "6" = "Dendritic cell",
  "7" = "Platelet"
)

seurat$cell_type <- recode(as.character(seurat$seurat_clusters),
                            !!!cell_type_map)

# Final UMAP
DimPlot(seurat, group.by = "cell_type",
        label = TRUE, repel = TRUE,
        label.size = 4) +
  ggtitle("PBMC — Cell type annotation") +
  theme(legend.position = "none")
```

---

# Differential Expression in scRNA-seq {background-color="#0F6E56"}

---

## Pseudobulk — The Right Way

```{r}
library(DESeq2)

# Wrong: treat each cell as independent observation → inflated p-values
# Right: aggregate cells per sample → pseudobulk → DESeq2

# Aggregate counts per cell type per sample
pseudobulk <- AggregateExpression(
  seurat,
  group.by     = c("cell_type", "sample_id"),
  assays       = "RNA",
  return.seurat = FALSE
)

# For each cell type, run DESeq2
run_pseudobulk_de <- function(cell_type_name, seurat_obj) {
  cells <- subset(seurat_obj, cell_type == cell_type_name)

  bulk  <- AggregateExpression(cells,
                                group.by   = "sample_id",
                                assays     = "RNA")$RNA

  meta  <- cells@meta.data |>
    distinct(sample_id, .keep_all = TRUE) |>
    column_to_rownames("sample_id")

  dds   <- DESeqDataSetFromMatrix(bulk, meta, ~ condition)
  dds   <- DESeq(dds)
  results(dds)
}

de_cd8 <- run_pseudobulk_de("CD8 T cell", seurat)
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: PBMC 3k (10x Genomics)

```{r}
library(Seurat)
library(SeuratData)
library(tidyverse)
library(here)

# Install & load PBMC 3k dataset
InstallData("pbmc3k")
data("pbmc3k")

pbmc3k
# 2,700 cells, 13,714 genes
```

---

## Hands-On: Full Pipeline

```{r}
pbmc <- pbmc3k

# QC
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- subset(pbmc,
               nFeature_RNA > 200 &
               nFeature_RNA < 2500 &
               percent.mt   < 5)

# Normalize + Variable features + Scale
pbmc <- NormalizeData(pbmc)
pbmc <- FindVariableFeatures(pbmc, nfeatures = 2000)
pbmc <- ScaleData(pbmc)

# Dim reduction + cluster
pbmc <- RunPCA(pbmc, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
pbmc <- RunUMAP(pbmc, dims = 1:10)

DimPlot(pbmc, label = TRUE)
```

---

## Hands-On: Annotate & Explore

```{r}
# Top markers per cluster
markers <- FindAllMarkers(pbmc, only.pos = TRUE,
                           min.pct = 0.25,
                           logfc.threshold = 0.25)

markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 3) |>
  print(n = Inf)

# Known PBMC markers
FeaturePlot(pbmc,
            features = c("MS4A1",   # B cells
                         "CD8A",    # CD8 T
                         "LYZ",     # Monocytes
                         "GNLY",    # NK
                         "FCGR3A",  # NK / Mono
                         "FCER1A"), # DC
            ncol = 3)
```

---

## Save for Next Week

```{r}
# Annotate clusters
pbmc$cell_type <- recode(pbmc$seurat_clusters,
  `0` = "CD4 T",    `1` = "CD14 Mono", `2` = "CD4 T",
  `3` = "B cell",   `4` = "CD8 T",     `5` = "FCGR3A Mono",
  `6` = "NK",       `7` = "DC",        `8` = "Platelet")

# Save for W9
saveRDS(pbmc, here("data", "processed", "pbmc_annotated.rds"))
```

---

# Next Week {background-color="#534AB7"}

---

## W9 Preview — Advanced scRNA-seq: Trajectory & Cell–Cell Communication

**What we'll cover:**

- Pseudotime trajectory analysis with Monocle3
- RNA velocity — predicting cell fate direction
- Cell–cell communication with CellChat
- Multi-sample integration with Harmony

**Install before next class:**

```{r}
BiocManager::install("monocle3")
install.packages(c("CellChat", "harmony", "velocyto.R"))
```

---

## Key Resources

- **Seurat** — [satijalab.org/seurat](https://satijalab.org/seurat) — comprehensive tutorials
- **SingleR** — [bioconductor.org/packages/SingleR](https://bioconductor.org/packages/SingleR)
- **Human Cell Atlas** — [humancellatlas.org](https://www.humancellatlas.org) — reference for annotation
- **10x Genomics datasets** — [10xgenomics.com/datasets](https://www.10xgenomics.com/datasets)
- **Chan Zuckerberg CELLxGENE** — [cellxgene.cziscience.com](https://cellxgene.cziscience.com) — millions of cells

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Advanced Single-cell Analysis: Trajectory & Cell–Cell Communication**

Course materials → [your GitHub Pages URL here]
:::
