---
title: "Multi-omics Integration & GWAS Gene Prioritization"
subtitle: "Advanced Bioinformatics for Clinician-Scientists | Week 11"
author: "Your Name"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    code-copy: true
    highlight-style: github
    transition: fade
    background-transition: fade
    footer: "Advanced Bioinformatics | W11 — Multi-omics Integration"
    css: custom.css
execute:
  echo: true
  eval: false
---

# The Full Picture {background-color="#1D9E75"}

---

## What We've Built So Far

```
W2–W3   GWAS / PRS       → genetic variants associated with disease
W4–W5   Epigenomics      → regulatory landscape (methylation, chromatin)
W6–W7   Metagenomics     → microbial community structure & function
W8–W9   scRNA-seq        → cell-type-specific expression & dynamics
W10     Spatial          → tissue-level context
```

::: {.fragment}
Each layer tells a **partial story**.

> **Multi-omics integration** combines these layers to find  
> convergent, cross-validated biological signals.
:::

---

## Why Integration?

:::: {.columns}
::: {.column width="50%"}
**Single-omics**

- High-dimensional
- Noisy
- Incomplete picture
- Many false positives
:::
::: {.column width="50%"}
**Multi-omics**

- Cross-layer validation
- More robust signals
- Mechanistic links
- Better biomarkers
:::
::::

::: {.fragment}
> A gene that is a GWAS hit **and** differentially methylated **and** overexpressed in a specific cell type  
> is far more likely to be causally relevant than any one signal alone.
:::

---

## Integration Strategies

| Strategy | Approach | Best for |
|---------|---------|---------|
| **Correlation** | Pairwise association between layers | Exploratory, hypothesis generation |
| **MOFA2** | Factor analysis across layers | Discovering shared variation |
| **mixOmics** | Supervised multiblock methods | Prediction, biomarker discovery |
| **DIABLO** | Supervised multi-omics | Classification with omics data |
| **SNF** | Similarity network fusion | Patient subtyping |
| **eQTL / colocalization** | GWAS + expression | Gene prioritization |

---

# MOFA2 — Multi-Omics Factor Analysis {background-color="#534AB7"}

---

## The MOFA2 Idea

```
Omics layer 1: RNA-seq    (n_samples × n_genes)
Omics layer 2: Methylation (n_samples × n_CpGs)
Omics layer 3: Metabolomics (n_samples × n_metabolites)
         ↓
MOFA2: find latent factors that explain variance ACROSS layers

Factor 1 → driven by cell cycle genes + methylation changes
Factor 2 → driven by immune activity in RNA + microbiome
Factor 3 → driven by metabolic genes + metabolites
```

::: {.fragment}
Each factor = a **biological axis of variation** shared across omics layers
:::

---

## Building a MOFA2 Model

```{r}
library(MOFA2)
library(tidyverse)
library(here)

# Load each omics layer (same samples, different features)
rna   <- readRDS(here("data", "processed", "rna_matrix.rds"))     # genes × samples
meth  <- readRDS(here("data", "processed", "meth_matrix.rds"))    # CpGs × samples
micro <- readRDS(here("data", "processed", "microbiome_matrix.rds")) # taxa × samples

# Create MOFA object
mofa <- create_mofa(list(
  RNA          = rna,
  Methylation  = meth,
  Microbiome   = micro
))

# Visualize data overview
plot_data_overview(mofa)
```

---

## Set Options & Train

```{r}
# Model options
model_opts <- get_default_model_options(mofa)
model_opts$num_factors <- 15

# Training options
train_opts <- get_default_training_options(mofa)
train_opts$seed           <- 42
train_opts$convergence_mode <- "slow"   # more accurate

mofa <- prepare_mofa(mofa,
                     model_options   = model_opts,
                     training_options = train_opts)

# Train (takes ~5–10 min)
mofa <- run_mofa(mofa,
                 outfile = here("results", "mofa_model.hdf5"))

# Load pre-trained model
mofa <- load_model(here("results", "mofa_model.hdf5"))
```

---

## Inspect Factors

```{r}
# Variance explained per factor per view
plot_variance_explained(mofa, max_r2 = 15)

# Factor values per sample (colored by metadata)
plot_factor(mofa,
            factor  = 1:5,
            color_by = "diagnosis",
            shape_by = "batch")

# Which features drive each factor?
plot_top_weights(mofa,
                 view   = "RNA",
                 factor = 1,
                 nfeatures = 20)
```

---

## Associating Factors with Clinical Variables

```{r}
# Add sample metadata
samples_metadata(mofa) <- sample_metadata   # data.frame with sample info

# Correlation of factors with clinical variables
correlate_factors_with_covariates(
  mofa,
  covariates  = c("age", "bmi", "disease_duration",
                  "CRP", "diagnosis"),
  plot        = "log_pval"
)

# Factor 3 correlates with CRP → inflammatory factor?
# Inspect top RNA features in Factor 3
get_weights(mofa, view = "RNA", factor = 3) |>
  as_tibble(rownames = "gene") |>
  arrange(desc(abs(value))) |>
  head(20)
```

---

# mixOmics / DIABLO — Supervised Integration {background-color="#185FA5"}

---

## DIABLO — Multi-block Discriminant Analysis

```
Goal: find a combination of features across omics layers
      that best discriminates between groups (e.g., responder vs. non-responder)

Input:
  X = list of omics matrices (RNA, methylation, proteomics)
  Y = outcome variable (factor: responder / non-responder)

Output:
  - Selected features per layer
  - Correlation structure between layers
  - Classification model
```

---

## Running DIABLO

```{r}
library(mixOmics)

# Input data
X <- list(
  RNA    = t(rna_filtered),      # samples × features
  Methyl = t(meth_filtered),
  Micro  = t(micro_filtered)
)
Y <- factor(sample_metadata$response)   # "Responder" / "Non-responder"

# Design matrix — how correlated should layers be?
# 1 = maximize inter-layer correlation
# 0 = maximize discrimination only
design <- matrix(0.1,
                  ncol = length(X),
                  nrow = length(X),
                  dimnames = list(names(X), names(X)))
diag(design) <- 0

# Initial DIABLO run
diablo <- block.splsda(X, Y,
                        ncomp  = 5,
                        design = design)
```

---

## Tuning & Final Model

```{r}
# Tune number of features to select per layer
set.seed(42)
perf_diablo <- perf(diablo,
                     validation    = "Mfold",
                     folds         = 5,
                     nrepeat       = 10,
                     dist          = "centroids.dist")

# Optimal number of components
plot(perf_diablo)
ncomp_opt <- perf_diablo$choice.ncomp$WeightedVote["Overall.BER",
                                                     "centroids.dist"]

# Tune number of features
test_keepX <- list(RNA    = c(5, 10, 15, 20),
                   Methyl = c(5, 10, 15),
                   Micro  = c(3, 5, 10))

tune_diablo <- tune.block.splsda(X, Y,
                                  ncomp      = ncomp_opt,
                                  test.keepX = test_keepX,
                                  design     = design,
                                  validation = "Mfold",
                                  folds      = 5,
                                  nrepeat    = 10)

# Final model
final_diablo <- block.splsda(X, Y,
                              ncomp    = ncomp_opt,
                              keepX    = tune_diablo$choice.keepX,
                              design   = design)
```

---

## DIABLO Visualizations

```{r}
# Circos plot — cross-layer correlations of selected features
circosPlot(final_diablo,
           cutoff        = 0.7,
           line          = TRUE,
           color.blocks  = c("#1D9E75", "#534AB7", "#D85A30"),
           color.cor     = c("#D85A30", "#1D9E75"))

# Sample plot
plotDiablo(final_diablo, ncomp = 1)

# Heatmap of selected features
cimDiablo(final_diablo,
          color.blocks = c("#1D9E75", "#534AB7", "#D85A30"),
          comp         = 1,
          margin       = c(8, 20),
          legend.position = "topright")
```

---

# GWAS Gene Prioritization {background-color="#0F6E56"}

---

## The Problem: GWAS Hit → Causal Gene

A GWAS locus typically contains:
- 1 lead SNP
- Dozens of correlated SNPs in LD
- Multiple nearby genes

Which gene is the actual effector?

```
Evidence layers for gene prioritization:
├── Positional     → nearest gene
├── eQTL           → gene whose expression is controlled by the SNP
├── Chromatin      → SNP overlaps enhancer contacting gene X (Hi-C)
├── Epigenomic     → SNP in cell-type-specific open chromatin
├── scRNA-seq      → gene expressed in disease-relevant cell type
└── Multi-omics    → gene appears across multiple evidence layers
```

---

## MAGMA — Gene-Level GWAS Analysis

```{r}
# MAGMA: aggregates SNP p-values to gene-level scores
# Run from command line (pre-computed)
system(paste(
  "magma",
  "--bfile",   here("data", "raw", "reference_panel"),
  "--gene-annot", here("data", "raw", "magma.genes.annot"),
  "--pval",    here("data", "raw", "gwas_sumstats.txt"),
                  "N=150000",
  "--gene-model", "snp-wise=mean",
  "--out",     here("results", "magma_gene")
))

# Load gene-level results
magma_genes <- read_tsv(here("results", "magma_gene.genes.out"),
                         col_types = cols()) |>
  arrange(P)

head(magma_genes)
```

---

## H-MAGMA — Cell-Type-Specific Prioritization

```{r}
# H-MAGMA: uses Hi-C contacts from specific cell types
# to map SNPs to genes via chromatin interactions

# Pre-computed H-MAGMA results for different tissues
# (neurons, cardiomyocytes, immune cells, etc.)

hmagma_neuron <- read_tsv(
  here("data", "processed", "hmagma_adult_brain.genes.out")
) |> arrange(P)

hmagma_immune <- read_tsv(
  here("data", "processed", "hmagma_immune.genes.out")
) |> arrange(P)

# Compare: different genes prioritized in neurons vs. immune cells
# for same GWAS → cell-type specificity of disease mechanism
inner_join(
  hmagma_neuron  |> select(GENE, P_neuron  = P),
  hmagma_immune  |> select(GENE, P_immune  = P)
) |>
  filter(P_neuron < 0.05 | P_immune < 0.05) |>
  mutate(cell_type_specific = case_when(
    P_neuron < 0.05 & P_immune > 0.05 ~ "Neuron-specific",
    P_immune < 0.05 & P_neuron > 0.05 ~ "Immune-specific",
    TRUE ~ "Both"
  ))
```

---

## Convergent Evidence Scoring

```{r}
# Combine multiple lines of evidence
# Each layer contributes points toward gene prioritization

gene_scores <- tibble(gene = all_candidate_genes) |>

  # 1. GWAS: MAGMA gene p-value
  left_join(magma_genes |> select(gene = GENE, magma_p = P)) |>
  mutate(score_gwas = -log10(magma_p)) |>

  # 2. eQTL colocalization
  left_join(coloc_results |> select(gene, pp_h4)) |>
  mutate(score_coloc = ifelse(pp_h4 > 0.8, 3, 0)) |>

  # 3. scRNA-seq — expressed in disease cell type?
  left_join(scrna_markers |> select(gene, avg_log2FC)) |>
  mutate(score_scrna = ifelse(avg_log2FC > 0.5, 2, 0)) |>

  # 4. ATAC-seq — SNP in open chromatin?
  left_join(atac_overlap |> select(gene, in_peak)) |>
  mutate(score_atac = ifelse(in_peak, 2, 0)) |>

  # Total convergent score
  mutate(total_score = score_gwas + score_coloc +
                       score_scrna + score_atac) |>
  arrange(desc(total_score))

head(gene_scores, 20)
```

---

## Visualizing Convergent Evidence

```{r}
library(ggplot2)
library(patchwork)

# Lollipop chart — top prioritized genes
top_genes <- gene_scores |> slice_max(total_score, n = 20)

ggplot(top_genes,
       aes(x = total_score,
           y = reorder(gene, total_score))) +
  geom_segment(aes(xend = 0, yend = gene),
               color = "gray70", linewidth = 0.8) +
  geom_point(aes(color = score_coloc > 0,
                 size  = score_gwas),
             alpha = 0.9) +
  scale_color_manual(values = c("#888780", "#1D9E75"),
                     labels  = c("No colocalization",
                                 "eQTL colocalized")) +
  labs(x     = "Convergent evidence score",
       y     = "Gene",
       color = "eQTL",
       size  = "GWAS score",
       title = "Multi-evidence gene prioritization") +
  theme_minimal()
```

---

# Hands-On {background-color="#BA7517"}

---

## Dataset: TCGA Multi-omics (BRCA)

TCGA Breast Cancer — RNA-seq + DNA methylation + miRNA

```{r}
library(TCGAbiolinks)
library(MOFA2)
library(tidyverse)
library(here)

# Download TCGA BRCA multi-omics (pre-queried)
# RNA-seq, methylation (450K), miRNA
rna_brca   <- readRDS(here("data", "raw", "tcga_brca_rna.rds"))
meth_brca  <- readRDS(here("data", "raw", "tcga_brca_meth.rds"))
mirna_brca <- readRDS(here("data", "raw", "tcga_brca_mirna.rds"))

# Align samples
common_samples <- Reduce(intersect,
                          list(colnames(rna_brca),
                               colnames(meth_brca),
                               colnames(mirna_brca)))

cat("Common samples:", length(common_samples), "\n")
```

---

## Hands-On: MOFA2

```{r}
# Subset to highly variable features per layer
select_hv <- function(mat, n = 1000) {
  rv <- rowVars(mat)
  mat[order(rv, decreasing = TRUE)[1:n], ]
}

mofa_obj <- create_mofa(list(
  RNA   = select_hv(rna_brca[, common_samples],  1000),
  Methyl = select_hv(meth_brca[, common_samples], 500),
  miRNA = select_hv(mirna_brca[, common_samples], 200)
))

model_opts            <- get_default_model_options(mofa_obj)
model_opts$num_factors <- 10

mofa_obj <- prepare_mofa(mofa_obj, model_options = model_opts)
mofa_obj <- run_mofa(mofa_obj, use_basilisk = TRUE)

plot_variance_explained(mofa_obj)
```

---

## Hands-On: Factor–Subtype Association

```{r}
# PAM50 subtypes as metadata
samples_metadata(mofa_obj) <- tcga_metadata |>
  filter(sample %in% common_samples) |>
  column_to_rownames("sample")

# Which factors separate PAM50 subtypes?
plot_factor(mofa_obj,
            factor   = 1:5,
            color_by = "PAM50",
            dodge    = TRUE,
            legend   = TRUE)

# Top RNA drivers of Factor 1
plot_top_weights(mofa_obj,
                 view      = "RNA",
                 factor    = 1,
                 nfeatures = 15,
                 scale     = TRUE)
```

---

# Next Week {background-color="#534AB7"}

---

## W12 Preview — Network Analysis & Drug Repositioning

**What we'll cover:**

- Protein–protein interaction networks
- Community detection — finding gene modules
- Drug target identification from networks
- Drug repositioning using transcriptomic signatures (CMAP/L1000)

**Install before next class:**

```{r}
install.packages(c("igraph", "ggraph", "tidygraph"))
BiocManager::install(c("STRINGdb", "fgsea"))
install.packages("cmapR")   # CMAP/L1000 interface
```

---

## Key Resources

- **MOFA2** — [biofam.github.io/MOFA2](https://biofam.github.io/MOFA2)
- **mixOmics / DIABLO** — [mixomics.org](http://www.mixomics.org)
- **TCGA** — [portal.gdc.cancer.gov](https://portal.gdc.cancer.gov)
- **H-MAGMA** — [github.com/thewonlab/H-MAGMA](https://github.com/thewonlab/H-MAGMA)
- **Open Targets** — [platform.opentargets.org](https://platform.opentargets.org) — convergent evidence scoring

---

## Questions? {background-color="#1D9E75"}

::: {.center}
Next week: **Network Analysis & Drug Repositioning**

Course materials → [your GitHub Pages URL here]
:::
