---
title: "W11. Multi-omics Integration"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Distinguish **sequential** from **joint** integration, and say which your question needs
-   Use convergent evidence across layers to narrow a locus to a single candidate gene
-   ⚠️ Say why an absent signal in one layer is not a negative result
-   Read a MOFA variance-explained plot, and ⚠️ recognise a technical factor
-   ⚠️ Anticipate the three failures that happen before any model is fitted: n collapse, scale imbalance, uninterpretable factors
-   Say what integration genuinely buys, and what it cannot manufacture
:::

> 🤔 **Motivating question**
>
> You have genotypes, RNA-seq, proteomics and clinical data on the same cohort. Should you put them all into one model?

------------------------------------------------------------------------

# 1. Two Ways to Integrate 🧭 {#w11-s1}

## 1) Sequential — each layer filters the last {#w11-s1-1}

You already did this without calling it integration. W3 narrowed a locus; W5 named a cell type; W9 asked whether the protein exists. Each layer removes candidates.

-   ✅ Keeps your **full n** at every step, because each analysis uses only the samples that have that layer
-   ✅ Every step is interpretable on its own and defensible in a methods section
-   ✅ Layers can come from **different cohorts** — public eQTL data, a published atlas, your own samples
-   ⚠️ Cannot discover a pattern that only exists *across* layers

## 2) Joint — model the layers together {#w11-s1-2}

**MOFA+**, **mixOmics/DIABLO**, **iCluster** and similar methods find latent factors that explain variance across several layers at once.

-   ✅ Can reveal coordinated programmes no single layer shows
-   ✅ Unsupervised — can find structure you did not ask about (patient subtypes)
-   ⚠️ Needs the **same samples** measured on every layer
-   ⚠️ Output is a factor, not a mechanism

💡 **Answer to the motivating question:** usually **no, not first**. Sequential prioritisation answers most clinical questions, keeps your n, and produces a figure a reviewer can follow. Reach for joint modelling when you specifically suspect a cross-layer programme, or when you want unsupervised patient subtypes.

------------------------------------------------------------------------

# 2. Sequential Integration in Practice 🎯 {#w11-s2}

![The most useful form of integration for a clinician, and the least glamorous.](/assets/img/w11/prioritization.svg){fig-align="center"}

## 1) Why convergence is strong evidence {#w11-s2-1}

Each layer has its own false positives, and they are largely **independent**: an eQTL colocalisation artifact has nothing to do with a proteomics artifact. A gene surviving five independent filters is unlikely to be five separate coincidences.

## 2) ⚠️ Why an absence is not evidence {#w11-s2-2}

::: {.callout-important collapse="true"}
## The asymmetry that gets misread constantly

A ✕ in one column almost always means **"not measured in the right context"**, not "measured and absent".

-   ⚠️ eQTL panels come from bulk tissue — effects confined to a rare cell type are invisible
-   ⚠️ Plasma proteomics misses low-abundance proteins entirely (🔗 [W9 §3](../part5/w9_proteomics_tech.qmd#w9-s3))
-   ⚠️ Some genes have no protein product to measure
-   ⚠️ MR needs a strong instrument; a gene without one cannot be tested, which is not the same as being tested and failing

→ ✅ Report convergence as positive evidence. Report gaps as **coverage limitations**, in a sentence, in the results.

⚠️ And the failure mode this creates: prioritisation pipelines systematically favour **well-measured, well-studied genes**. The same bias returns in W12 as network hubs.
:::

------------------------------------------------------------------------

# 3. Joint Models — Reading the Output 📐 {#w11-s3}

![Read the columns to find the technical factors, then the rows to find the biology.](/assets/img/w11/mofa_factors.svg){fig-align="center"}

## 1) What a factor is {#w11-s3-1}

A latent axis along which samples vary, together with **loadings** saying which features in which layers drive it. The variance-explained plot says how much of each layer each factor accounts for.

## 2) ⚠️ Annotate factors before interpreting them {#w11-s3-2}

::: {.callout-warning collapse="true"}
## Factor 1 is almost always technical

The strongest shared source of variation in a real cohort is rarely disease. It is batch, RNA integrity, sex, or age.

✅ The annotation routine, in order:

1.  **Technical** — batch, run date, RIN/DV200, operator, storage time
2.  **Demographic** — age, sex, BMI, ancestry
3.  **Clinical confounders** — medication, comorbidity, disease duration
4.  **Only then** — disease status or outcome

⚠️ If a factor correlates with batch at r = 0.8 and with disease at r = 0.3, it is a batch factor that happens to be slightly confounded with disease. Reporting it as a "disease signature" is a real and common error.

💡 A factor that loads across **several** layers is more likely to be biological than one confined to a single layer — but only after steps 1–3 come back empty.
:::

------------------------------------------------------------------------

# 4. ⚠️ Three Failures Before the Model 🕳️ {#w11-s4}

![None of these is a modelling problem, and all of them are decided at design time.](/assets/img/w11/integration_traps.svg){fig-align="center"}

**① Your n collapses.** Joint models use complete cases. A cohort of 300 with four layers may have 41 patients measured on all four — and that is your real sample size.

-   ✅ Check the intersection **before** planning the analysis
-   ✅ Some methods handle missing views; using one is a decision to state, not a default

**② Scale imbalance.** Methylation contributes 450,000 features and your clinical variables contribute 12. Without per-layer scaling, the factors describe methylation.

**③ The factor is not a mechanism.** See §3.2.

::: {.callout-tip collapse="true"}
## What integration cannot do

-   ❌ It cannot create power. Four noisy layers on 41 patients is not stronger than one good layer on 300
-   ❌ It cannot fix confounding. If batch tracks disease, every layer inherits it and the joint model concentrates it into a factor
-   ❌ It cannot make an underpowered study publishable. Reviewers have become good at spotting integration used as a substitute for n

→ ✅ What it genuinely buys: **cross-layer coordination you could not otherwise see**, and **unsupervised subtypes** you did not know to look for.
:::

------------------------------------------------------------------------

# 5. Hands-On 💻 {#w11-s5}

## 1) Build a prioritisation table {#w11-s5-1}

Take one locus for a disease you treat, and fill in the table from §2 by hand, in the browser:

-   **eQTL** — [Open Targets Genetics](https://genetics.opentargets.org) or [GTEx](https://gtexportal.org)
-   **Cell type** — [CELLxGENE](https://cellxgene.cziscience.com)
-   **Protein** — [Human Protein Atlas](https://www.proteinatlas.org)
-   **Direction** — [IEU OpenGWAS](https://gwas.mrcieu.ac.uk) for MR results
-   **Tractability** — [Open Targets](https://platform.opentargets.org) target profile

✅ One row per candidate gene, one column per layer. This is an afternoon, and it is a real figure.

## 2) Check the intersection first {#w11-s5-2}

```{r}
library(tidyverse); library(UpSetR)

have <- meta |> transmute(
  clinical   = 1L,
  genotype   = as.integer(!is.na(geno_id)),
  rna        = as.integer(!is.na(rna_id)),
  proteomics = as.integer(!is.na(prot_id)))

upset(as.data.frame(have), order.by = "freq")
cat("complete cases:", sum(rowSums(have) == ncol(have)), "\n")   # ⚠️ this is your n
```

## 3) MOFA, and annotating its factors {#w11-s5-3}

```{r}
library(MOFA2)

mofa <- create_mofa(list(RNA = rna_mat, Protein = prot_mat, Methyl = meth_mat)) |>
  prepare_mofa() |> run_mofa()

plot_variance_explained(mofa)          # the figure above

# ⚠️ the step that must come next
Z <- get_factors(mofa)[[1]]
covars <- meta |> select(batch, RIN, age, sex, bmi, condition)

cors <- sapply(covars, \(v) apply(Z, 2, \(f) {
  if (is.numeric(v)) abs(cor(f, v, use = "pairwise")) else summary(aov(f ~ factor(v)))[[1]][1, "F value"]
}))
round(cors, 2)     # look here before you look at biology
```

## 4) ✅ Change one thing {#w11-s5-4}

```{r}
# Re-run MOFA with the batch effect removed from each layer first.
# Does Factor 1 disappear? Does anything else change rank?
```

💡 If removing batch reshuffles your factors, that tells you how much of the original result was batch — a number worth knowing before it appears in review.

## 5) Before next week {#w11-s5-5}

**W12 — networks and drug targets.** Bring the prioritisation table from §5.1. We will ask whether the surviving gene is druggable.

------------------------------------------------------------------------

# 📝 Summary

-   **Sequential** integration filters candidates layer by layer; **joint** integration models layers together
-   ✅ Sequential keeps your full n, works across cohorts, and answers most clinical questions
-   ✅ Convergence across layers is strong because each layer's false positives are **independent**
-   ⚠️ **An absent signal is a coverage gap, not a negative result** — and it biases prioritisation toward well-studied genes
-   ⚠️ **Factor 1 is almost always technical.** Annotate against batch, RIN, age and sex *before* disease
-   ⚠️ Joint models use complete cases — **your n is the intersection**, not the cohort
-   ⚠️ Without per-layer scaling, the layer with most features dominates every factor
-   ❌ Integration cannot create power, fix confounding, or rescue an underpowered study

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| Sequential integration | Each layer filters candidates from the previous one |
| Joint integration | Modelling several layers simultaneously |
| Latent factor | An axis of variation shared across layers |
| Loadings | Which features drive a factor, per layer |
| Variance explained | How much of each layer a factor accounts for |
| MOFA+ | Unsupervised multi-omics factor analysis |
| DIABLO / mixOmics | Supervised multi-omics discriminant analysis |
| Complete case | A sample measured on every layer |
| Per-layer scaling | Preventing the largest layer from dominating |
| Convergent evidence | Independent layers pointing at the same gene |
| Coverage gap | A layer that could not have detected the signal |

------------------------------------------------------------------------

# ➕ References

### Methods

Argelaguet R, et al. [*MOFA+: a statistical framework for comprehensive integration of multi-modal single-cell data.*](https://doi.org/10.1186/s13059-020-02015-1) Genome Biol. 2020.\
Argelaguet R, et al. [*Multi-Omics Factor Analysis — a framework for unsupervised integration of multi-omics data sets.*](https://doi.org/10.15252/msb.20178124) Mol Syst Biol. 2018.\
Singh A, et al. [*DIABLO: an integrative approach for identifying key molecular drivers from multi-omics assays.*](https://doi.org/10.1093/bioinformatics/bty1054) Bioinformatics. 2019.

### Perspective and pitfalls

Argelaguet R, Cuomo ASE, Stegle O, Marioni JC. [*Computational principles and challenges in single-cell data integration.*](https://doi.org/10.1038/s41587-021-00895-7) Nat Biotechnol. 2021.\
Vasaikar SV, et al. [*LinkedOmics: analyzing multi-omics data within and across 32 cancer types.*](https://doi.org/10.1093/nar/gkx1090) Nucleic Acids Res. 2018.\
Subramanian I, et al. [*Multi-omics data integration, interpretation, and its application.*](https://doi.org/10.1177/1177932219899051) Bioinform Biol Insights. 2020.

### Gene prioritisation

Mountjoy E, et al. [*An open approach to systematically prioritize causal variants and genes at all published human GWAS trait-associated loci.*](https://doi.org/10.1038/s41588-021-00945-5) Nat Genet. 2021. (Open Targets Genetics locus-to-gene)\
Ochoa D, et al. [*The next-generation Open Targets Platform.*](https://doi.org/10.1093/nar/gkac1046) Nucleic Acids Res. 2023.
