---
title: "W12. Network Analysis & Drug Targets"
author: "Eunji Ha"
date: last-modified
format: html
execute:
  echo: true
  eval: false
---

::: {.callout-note collapse="true"}
## 🎯 Learning objectives

By the end of this chapter you should be able to:

-   Say what a biological network is built from, and ⚠️ why that construction biases every result
-   ⚠️ Recognise the study-bias artifact behind most "hub gene" findings
-   Compare the three routes to drug repurposing, and say what each is worth as evidence
-   ⚠️ Explain why direction of effect decides agonist versus antagonist
-   Work through six questions that separate a top-ranked gene from an actual target
-   Use PheWAS and loss-of-function carriers as a free safety readout
:::

> 🤔 **Motivating question**
>
> A network analysis of your disease reports TP53, EGFR and AKT1 as the key regulators. How surprised should you be?

------------------------------------------------------------------------

# 1. ⚠️ What a Network Is Made Of 🕸️ {#w12-s1}

## 1) The construction, and the bias it creates {#w12-s1-1}

A protein–protein interaction network is an accumulation of **published experiments**. STRING, BioGRID and IntAct are curated literature, plus predictions trained on that literature.

Experiments concentrate on genes people already study. So node degree tracks **publication count**.

![The correlation is not subtle.](/assets/img/w12/network_bias.svg){fig-align="center"}

::: {.callout-important collapse="true"}
## The answer to the motivating question is: not at all surprised

TP53, EGFR and AKT1 come out as hubs for nearly every disease that has ever been networked. They have tens of thousands of papers each, so they have edges to everything.

⚠️ Concretely, this means:

-   "Hub" analyses rediscover the same genes regardless of your data
-   **Network proximity** scores favour well-studied drugs and well-studied targets
-   Genes with no publications have **degree ≈ 0** and can never be prioritised, however important

✅ What to do:

-   Compare against a **degree-matched null** — permute while preserving each node's degree, so "well-connected" stops being an automatic win
-   Prefer networks built from a **single systematic experiment** (a proteomics interactome, a CRISPR screen) over literature aggregates, where you can
-   ✅ Report degree alongside any hub claim, and let the reader see it
:::

## 2) What networks are still good for {#w12-s1-2}

-   ✅ **Modules** — a set of genes that interact more than expected, giving a process-level rather than gene-level result
-   ✅ **Connecting your hits** — do your 40 GWAS loci converge on a pathway?
-   ✅ **Suggesting a mechanism** for a gene of unknown function, from its neighbours
-   ⚠️ Always as a hypothesis. The edges are other people's experiments, not yours

------------------------------------------------------------------------

# 2. From Gene to Drug 💊 {#w12-s2}

![Three routes, three very different evidence classes.](/assets/img/w12/repurposing_routes.svg){fig-align="center"}

## 1) Genetic support is the strongest route {#w12-s2-1}

Targets with human genetic support are roughly **twice as likely** to succeed through clinical development. The mechanism is simple: a variant that alters the target is a natural, lifelong, randomised experiment in humans — the argument from 🔗 [W3 §3.2](../part2/w3_gwas.qmd#w3-s3-2).

-   **PCSK9** — loss-of-function carriers have low LDL and low cardiovascular risk → inhibitors work
-   **IL6R** — a variant mimicking IL-6 blockade associates with lower CHD risk → tocilizumab's cardiovascular signal

## 2) ⚠️ Direction of effect {#w12-s2-2}

::: {.callout-warning collapse="true"}
## The mistake with the worst consequences in this course

If the risk allele **increases** target activity, you want an inhibitor. If it **decreases** it, the target may be protective — and inhibiting it would make patients worse.

Two complications:

-   ⚠️ **Incoherence.** A risk allele can down-regulate a gene that appears up-regulated in patients. Often that is an impaired protective response, and the therapeutic direction is the opposite of what the expression data suggests
-   ⚠️ **Tissue-specificity.** Direction can differ between tissues. The relevant tissue is the one where the disease process happens (🔗 [W5](../part3/w5_scrna_qc.qmd), 🔗 [W7](../part4/w7_spatial_basics.qmd))

→ ✅ Establish direction from **genetics plus MR**, not from a fold change in patient tissue.
:::

## 3) Signature reversal and network proximity {#w12-s2-3}

-   **Signature reversal** (CMap, LINCS) — find a drug whose expression signature is anti-correlated with the disease signature. ⚠️ Reference profiles come from cell lines at high doses. Useful for generating candidates; not evidence
-   **Network proximity** — is a drug's target close to the disease module? ⚠️ Inherits every bias in §1, so it needs a degree-matched null to mean anything

------------------------------------------------------------------------

# 3. Six Questions Before Calling It a Target ✅ {#w12-s3}

![Answering all six for one gene is a complete piece of work.](/assets/img/w12/target_checklist.svg){fig-align="center"}

::: {.callout-tip collapse="true"}
## Question 5 in more detail: humans have already run your safety study

People carrying natural loss-of-function variants in your target are a preview of what inhibiting it does over a lifetime.

-   ✅ **gnomAD** — how many LoF carriers exist? Is the gene depleted of them (a **LOEUF** near 0 means intolerant, and probably a poor target)?
-   ✅ **PheWAS** on the variant — what else changes? Those are your predictable side effects
-   ✅ Homozygotes, where they exist, are the strongest signal available short of a trial

⚠️ A gene that is highly intolerant to loss of function is telling you something. Inhibiting it is likely to have consequences beyond the disease.
:::

------------------------------------------------------------------------

# 4. Hands-On 💻 {#w12-s4}

Bring the prioritisation table from 🔗 [W11 §5.1](w11_multiomics.qmd#w11-s5-1).

## 1) No code: profile your surviving gene {#w12-s4-1}

[**Open Targets**](https://platform.opentargets.org) → your gene → its target profile.

-   ① **Tractability** — small molecule? antibody? Which clinical phase, for any indication?
-   ② **Safety** — any flagged organ toxicity?
-   ③ **Known drugs** — is something already approved against it, for anything?
-   ④ [**gnomAD**](https://gnomad.broadinstitute.org) → **LOEUF**. Is the gene tolerant of loss of function?
-   ⑤ [**Open Targets Genetics**](https://genetics.opentargets.org) → PheWAS for the lead variant. What else does it affect?

✅ Write the answers as five sentences. That is §3's checklist, done.

## 2) Build a small network, then check yourself {#w12-s4-2}

```{r}
library(STRINGdb); library(igraph); library(tidyverse)

sdb <- STRINGdb$new(version = "12", species = 9606, score_threshold = 700)
mapped <- sdb$map(data.frame(gene = my_genes), "gene", removeUnmappedRows = TRUE)
g <- sdb$get_subnetwork(mapped$STRING_id) |> igraph::simplify()

deg <- sort(degree(g), decreasing = TRUE)
head(deg, 10)
```

⚠️ **Now the step that is usually missing:**

```{r}
# Is your "hub" just a well-studied gene?
# Fetch PubMed counts for the same genes and correlate with degree.
pubs <- readRDS(here::here("rawdata", "pubmed_counts.rds"))   # gene → n papers
cor(log10(pubs[names(deg)] + 1), deg, use = "complete", method = "spearman")
```

💡 In most literature-derived networks this correlation is strong. Report it, and your hub claim becomes credible instead of automatic.

## 3) A degree-matched null {#w12-s4-3}

```{r}
observed <- mean(distances(g, v = disease_genes, to = drug_targets))

null <- replicate(1000, {
  rw <- sample_degseq(degree(g), method = "vl")   # same degrees, rewired edges
  mean(distances(rw, v = disease_genes, to = drug_targets))
})

cat("observed:", round(observed, 2), " null mean:", round(mean(null), 2), "\n")
```

✅ If the observed proximity is not clearly better than the degree-matched null, you have measured popularity.

## 4) Before next week {#w12-s4-4}

**W13 — interpretation and visualisation.** Bring **one figure** from your own work or from a paper you are reading. We will take it apart.

------------------------------------------------------------------------

# 📝 Summary

-   Biological networks are aggregated published experiments, so ⚠️ **node degree tracks publication count**
-   ⚠️ Hub analyses rediscover TP53, EGFR and AKT1 for almost every disease. ✅ Use a degree-matched null
-   ⚠️ Genes with no literature have degree ≈ 0 and can never be prioritised, however important
-   ✅ Networks are good for **modules**, for connecting scattered hits, and for suggesting mechanism — as hypotheses
-   ✅ **Genetic support roughly doubles the chance a target reaches approval** — the strongest evidence class available
-   ⚠️ **Direction of effect decides agonist vs antagonist**, and getting it backwards harms patients
-   ⚠️ Signature reversal uses cell lines at high doses. Candidates, not evidence
-   ✅ Six questions: genetics · direction · expression · tractability · safety · **would it change management**
-   ✅ LoF carriers and PheWAS are a free, human, lifelong safety readout — use them before proposing a target

------------------------------------------------------------------------

# 🔑 Key terms

| Term | One-line meaning |
|-------------------------|-----------------------------------------|
| PPI network | Graph of protein–protein interactions, mostly from literature |
| Degree | Number of edges at a node |
| Hub | High-degree node — often a well-studied gene |
| Study bias | Degree reflecting research attention rather than biology |
| Degree-matched null | Rewiring that preserves degrees, to test a network claim |
| Module | A densely interconnected set of genes |
| Network proximity | Graph distance between drug targets and disease genes |
| Signature reversal | Finding a drug that anti-correlates with the disease signature |
| CMap / LINCS | Databases of drug-induced expression profiles |
| Tractability | Whether a target can be drugged at all |
| LOEUF | gnomAD metric of intolerance to loss of function |
| Human knockout | A natural homozygous LoF carrier — a lifetime safety preview |

------------------------------------------------------------------------

# ➕ References

### Networks and their biases

Schaefer MH, Serrano L, Andrade-Navarro MA. [*Correcting for the study bias associated with protein–protein interaction measurements reveals differences between protein degree distributions from different cancer types.*](https://doi.org/10.3389/fgene.2015.00260) Front Genet. 2015. **← the study-bias problem, measured**\
Gillis J, Ballouz S, Pavlidis P. [*Bias tradeoffs in the creation and analysis of protein–protein interaction networks.*](https://doi.org/10.1016/j.jprot.2014.01.020) J Proteomics. 2014.\
Barabási AL, Gulbahce N, Loscalzo J. [*Network medicine: a network-based approach to human disease.*](https://doi.org/10.1038/nrg2918) Nat Rev Genet. 2011.

### Drug targets and genetic support

Nelson MR, et al. [*The support of human genetic evidence for approved drug indications.*](https://doi.org/10.1038/ng.3314) Nat Genet. 2015. **← the paper behind "roughly doubles"**\
King EA, Davis JW, Degner JF. [*Are drug targets with genetic support twice as likely to be approved?*](https://doi.org/10.1371/journal.pgen.1008489) PLoS Genet. 2019.\
Reay WR, Cairns MJ. [*Advancing the use of GWAS for drug repurposing.*](https://doi.org/10.1038/s41576-020-00387-5) Nat Rev Genet. 2021.\
Minikel EV, et al. [*Evaluating drug targets through human loss-of-function genetic variation.*](https://doi.org/10.1038/s41586-020-2267-z) Nature. 2020. **← human knockouts as a safety readout**

### Repurposing methods

Subramanian A, et al. [*A next generation connectivity map: L1000 platform and the first 1,000,000 profiles.*](https://doi.org/10.1016/j.cell.2017.10.049) Cell. 2017.\
Guney E, Menche J, Vidal M, Barabási AL. [*Network-based in silico drug efficacy screening.*](https://doi.org/10.1038/ncomms10331) Nat Commun. 2016.

### Resources

[Open Targets](https://platform.opentargets.org) · [STRING](https://string-db.org) · [gnomAD](https://gnomad.broadinstitute.org) · [CLUE / LINCS](https://clue.io) · [DGIdb](https://dgidb.org)
